\documentclass[margin,line]{res}
\usepackage{verbatim}
\usepackage{url}
\urlstyle{same}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\oddsidemargin -.5in
\evensidemargin -.5in
\textwidth=6.0in
\itemsep=0in
\parsep=0in

\newenvironment{list1}{
  \begin{list}{\ding{113}}{%
      \setlength{\itemsep}{0in}
      \setlength{\parsep}{0in} \setlength{\parskip}{0in}
      \setlength{\topsep}{0in} \setlength{\partopsep}{0in} 
      \setlength{\leftmargin}{0.17in}}}{\end{list}}
\newenvironment{list2}{
  \begin{list}{$\bullet$}{%
      \setlength{\itemsep}{0in}
      \setlength{\parsep}{0in} \setlength{\parskip}{0in}
      \setlength{\topsep}{0in} \setlength{\partopsep}{0in} 
      \setlength{\leftmargin}{0.2in}}}{\end{list}}

\begin{document}

\name{
\begin{tabular*}{7.4in} {@{\extracolsep{\fill}}lr}
Toby Dylan Hocking & Curriculum Vitae 
\end{tabular*}
}

\begin{resume}
\section{\sc Contact and General Info}
\vspace{.05in}
\begin{tabular*}{6.1in} {@{\extracolsep{\fill}}ll}
 NAU Building 90, Office 210 & Birth: 17 March 1984 in Newport Beach, California\\
 1295 S. Knoles Dr.  & Citizenship: USA \\            
  Flagstaff, AZ 86011 & Language skills:\\
  Web: \url{http://tdhock.github.io} & English (native)\\
  E-mail:  toby.hocking@nau.edu & French (fluent since 2009)\\
\end{tabular*}

\section{\sc Research Interests}

Machine learning algorithms, statistical software, and data
visualization techniques. Emphasis on efficient algorithms for large
datasets, based on constrained optimization (regression,
classification, ranking, clustering, segmentation, changepoint
detection, survival analysis). Application domains include medicine,
genomics, neuroscience, audio, internet, sensors, recommendation
systems.

\section{\sc Professional Experience}

{\bf Northern Arizona University}, Flagstaff, Arizona, USA (2018-present).\\
\vspace*{-.1in}
\begin{list2}
\item[] Assistant Professor, School of Informatics, Computing, and Cyber Systems.
\item[] ``Optimization algorithms for machine learning and interactive data analysis.''
\end{list2}

{\bf McGill University}, Montreal, Canada (2014-2018).\\
\vspace*{-.1in}
\begin{list2}
\item[] Postdoc with Guillaume Bourque, Department of Human Genetics.
\item[]``Changepoint detection and regression models for peak detection in genomic data.''
\end{list2}

{\bf Tokyo Institute of Technology}, Tokyo, Japan (2013).\\
\vspace*{-.1in}
\begin{list2}
\item[] Postdoc with Masashi Sugiyama, Department of Computer Science.
\item[] ``Support vector machines for ranking and comparing.''
\end{list2}

{\bf Sangamo BioSciences}, Richmond, CA, USA (2006-2008).\\
\vspace*{-.1in}
\begin{list2}
\item[] Research Assistant with Jeff Miller in the Technology group.
\item[] ``A web app for visualization and statistical analysis of experimental data.''
\end{list2}

\section{\sc Education}

{\bf \'{E}cole Normale Sup\'{e}rieure}, Cachan, France (2009-2012).\\
\vspace*{-.1in}
\begin{list2}
\item[] Ph.D. with Francis Bach, D\'{e}partement d'Informatique; Jean-Philippe Vert, Institut Curie.
\item[] ``Learning algorithms and statistical software, with applications to bioinformatics."
\end{list2}

{\bf Universit\'e Paris 6}, Paris, France (2008-2009).\\
\vspace*{-.1in}
\begin{list2}
\item[] Master of Statistics, internship at INRA with Mathieu Gautier and Jean-Louis Foulley.
\item[] ``A Bayesian Outlier Criterion to Detect SNPs under Selection in Large Data Sets."
\end{list2}

{\bf University of California, Berkeley}, CA, USA (2002-2006).\\
\vspace*{-.1in}
\begin{list2}
\item[] Double B.A. in Statistics, Molecular and Cell Biology; thesis in Statistics with Terry Speed.
\item[] ``Chromosomal copy number analysis using SNP microarrays and a binomial test statistic.'' 
\end{list2}

\section{\sc Honors and Awards (Selected)}

Co-PI on National Science Foundation grant, \$3,000,000, Sept 2021 to
Aug 2026. ``MIM: Discovering in reverse: using isotopic translation
of omics to reveal ecological interactions in microbiomes.''

Air Force Research Laboratory, Summer Faculty Fellowship, May--July
2020, ``Machine learning algorithms for understanding physically
unclonable functions based on resistive memory devices.''

R Consortium Grant, \$34,000, Jan-Dec 2020, ``RcppDeepState: an easy
way to fuzz test compiled code in R packages.''

``Mobilit\'e entrant'' travel award to work with Guillem Rigaill in
Universit\'e Evry, France, 2016.

International useR conference, Best Student Poster Award, ``Adding
direct labels to plots,'' 2011.

% INRIA/INRA (French computer science and agricultural research institutes), Ph.D. scholarship, 2009 (declined).

% UC Berkeley, Department of Statistics VIGRE research scholarship, 2001.

% UC Berkeley, Cal Band George Miller scholarship, 2000.

\section{\sc Peer-reviewed publications (Selected)}

In addition to peer-reviewed journals, I publish papers at
computer science conferences such as {\it ICML} and {\it NeurIPS}, with
double-blind peer reviews, and $\approx$20\% acceptance rates.

{\bf Hocking TD}, Rigaill G, Fearnhead P, Bourque G. Generalized
Functional Pruning Optimal Partitioning (GFPOP) for Constrained
Changepoint Detection in Genomic Data. Accepted for publication in
{\it Journal of Statistical Software}, arXiv:1810.00117.

Fotoohinasab A, {\bf Hocking TD}, Afghah F. A Greedy Graph Search
Algorithm Based on Changepoint Analysis for Automatic QRS-Complex
Detection. {\it Computers in Biology and Medicine} 130 (2021).

{\bf Hocking TD}. Wide-to-tall data reshaping using regular
expressions and the nc package. {\it R Journal} (2021),
doi:10.32614/RJ-2021-029.

{\bf Hocking TD}, Rigaill G, Fearnhead P, Bourque G. Constrained
dynamic programming and supervised penalty learning algorithms for
peak detection in genomic data. {\it Journal of Machine Learning
  Research} 21(87):1--40, 2020.

Jewell S, {\bf Hocking TD}, Fearnhead P, Witten D. Fast Nonconvex
Deconvolution of Calcium Imaging Data. {\it Biostatistics} (2019), doi:
10.1093/biostatistics/kxy083.

Drouin A, {\bf Hocking TD}, Laviolette F. Maximum margin interval
trees. {\it Neural Information Processing Systems (NeurIPS)}, 2017.

{\bf Hocking TD}, Joulin A, Bach F, Vert J-P. Clusterpath: an
Algorithm for Clustering using Convex Fusion Penalties. {\it International
Conference on Machine Learning (ICML)}, 2011.

\section{\sc Books, Chapters, Manuals}

{\bf Hocking TD}. Introduction to Machine Learning and Neural
Networks. Chapter in textbook {\it New Advances in Land Carbon Cycle
  Modeling} edited by Yiqi Luo. (in press)

{\bf Hocking TD}. Animated interactive data visualization using the
grammar of graphics (The animint2 Manual), 17 web pages/chapters with
interactive graphics and exercises. (2018)

\section{\sc Conference Tutorials}

{\bf Hocking TD}, Killick R. Introduction to optimal changepoint
detection algorithms, {\it useR} 2017.

{\bf Hocking TD}, Ekstr\o m CT. Understanding and creating interactive
graphics, {\it useR} 2016.

\section{\sc Professional Service}

Co-administrator and mentor for R project in Google Summer of Code,
since 2012.

Editor for Journal of Statistical Software since 2018, and rOpenSci
Statistical Software since 2020.

President of organizing committee for ``R in Montreal 2018'' conference.

Reviewer since 2010: Technometrics, International Conference on
Machine Learning (ICML), Advances in Neural Information Processing
Systems (NeurIPS), Journal of Machine Learning Research (JMLR),
Artificial Intelligence Review, Journal of Computational and Graphical
Statistics (JCGS), R Journal, Bioinformatics, PLOS Computational
Biology, BMC Bioinformatics, IEEE Transactions on Pattern Analysis and
Machine Intelligence, Information and Inference, etc.

\end{resume}
\end{document}




