---
layout: post
title: R packages on github
description: How to query CRAN meta-data
---

```{r Ropts, echo=FALSE}
if(FALSE){
  knitr::knit("2022-08-17-packages-on-github.Rmd")
}
```

The goal here is to find beta testers for
[RcppDeepState-action](https://github.com/FabrizioSandri/RcppDeepState-action).
Previously, Akhila Kolla created [a list of R packages with
RcppDeepState issues](https://akhikolla.github.io./packages-folders/).
Some of these packages may be hosted on github, but how to find that
programmatically? 

First, we download the web page which lists all of the candidate packages,

```{r}
if(!file.exists("problems.html")){
  download.file(
    "https://akhikolla.github.io./packages-folders/",
    "problems.html")
}
(prob.dt <- nc::capture_all_str(
  "problems.html",
  '<li><a href="',
  Package=".*?",
  '[.]html">'))
```

The code above shows that there are `r nrow(prob.dt)` packages on that
web page. Can we find the CRAN meta-data for each of them? First we
download the current CRAN meta-data,

```{r}
if(!file.exists("packages.rds")){
  download.file(
    "https://cloud.r-project.org/web/packages/packages.rds",
    "packages.rds")
}
meta.mat <- readRDS("packages.rds")
nrow(meta.mat)
```

Then we can subset the meta-data based on the packages on that web page,

```{r}
library(data.table)
meta.dt <- data.table(meta.mat)
meta.prob <- meta.dt[prob.dt, on="Package"]
meta.prob[, .(Package, URL.truncated=substr(URL, 1, 50))]
```

The output above shows that there are several packages which mention
github in the URL field. To find the repo URL we can do,

```{r}
pkg.repos <- meta.prob[, nc::capture_all_str(
  c("",URL), # to avoid attempting to download URL.
  repo.url="https://github.com/.*?/[^#/ ,]+"),
  by=Package]
pkg.repos$repo.url
```

Exercise for the reader: programmatically fork each of these repos,
then make a new branch, then add a
[RcppDeepState.yaml](https://github.com/tdhock/binsegRcpp/blob/32d09699bc55c32c09a70b8580b21c335791fb81/.github/workflows/RcppDeepState.yaml)
file, then push. This may be doable in R via
[gh](https://github.com/r-lib/gh), an R package for making calls to
the GitHub API.

UPDATE: my excellent GSOC student Fabrizio Sandri describes how to do
that [on his
blog](https://fabriziosandri.github.io/gsoc-2022-blog/github%20action/2022/08/23/rcppdeepstate-beta-test.html).
