---
layout: post
title: Positive and negative log transform
description: Non-linear transformations for heat maps and signed p-values
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2022-07-14-pos-and-neg-log-transform-"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=7)
if(FALSE){
  knitr::knit("2022-07-14-pos-and-neg-log-transform.Rmd")
}
```

## diverging color heat map

I recently used the following code/transformation to make an
[informative plot of linear model
coefficients](https://rcdata.nau.edu/genomic-ml/nn_embedding_with_interpretable_figures/compare_weights_heat_map.png),

```{r log.pos.neg}
normalize <- function(x)(x-min(x))/(max(x)-min(x))
log.pos.neg <- function(x)sign(x)*normalize(log10(abs(x)))
curve(log.pos.neg(x),-9,9.1)
```

This curve is a nonlinear transformation which makes it easy to
visualize log-scale changes in both positive and negative numbers, so
is ideal for a diverging color scale.

## signed p-value transform

When you do a 2-sided `t.test` in R, for example to see if a machine learning
algorithm is more/less accurate than another, you get two things:

* `estimate` is the [effect size](https://en.wikipedia.org/wiki/Effect_size),
* `p.value` is the [p-value](https://en.wikipedia.org/wiki/P-value), probability of observing these data/differences, given the null hypothesis.

So you can have a really big difference, in either the positive or
negative direction (one or the other algorithm is quite a bit more
accurate), and the p-values will be small (`1e-10` maybe).  The
p-value by itself does not tell you which direction the differences
are tending, so for visualization we can multiply the p-value by the
sign of the difference, and use the transformation below to transform
into log scale. This enables us to visualize differences between very
small/significant p-values such as `1e-8` and `1e-9` which would not
be visible on the linear scale (without this transformation).

```{r log.signed.p}
library(data.table)
p.val <- 10^seq(-10, 0, l=101)
signed.p <- sort(c(sapply(c(-1,1), function(s)s*p.val)))
log.signed.p <- function(signed.p){
  sign(signed.p)*abs(log10(abs(signed.p)))
}
sign.dt <- data.table(signed.p, log.trans=log.signed.p(signed.p))
library(ggplot2)
ggplot()+
  geom_line(aes(
    signed.p, log.trans, group=sign(signed.p)),
    data=sign.dt)
```