---
layout: post
title: Reformatting NEWS files
description: Regular expression example
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.name <- "2022-09-15-reformat-news"
post.id <- paste0(post.name,"/")
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=7)
if(FALSE){
  post.Rmd <- paste0(post.name,".Rmd")
  knitr::knit(post.Rmd)
}
```

To make my research about new machine learning algorithms more
reproducible, I provide free/open-source implementations as R
packages, published on the CRAN (Comprehensive R Archive Network). In
2022, CRAN introduced a new policy that NEWS files must be in a
standard format, so I started to get CRAN check messages as below,

```
Flavor: r-devel-linux-x86_64-debian-gcc
Check: package subdirectories, Result: NOTE
  Problems with news in 'NEWS':
    Cannot process chunk/lines:
      Line search C++ code review, exclude speed and line search vignettes from CRAN.
    Cannot process chunk/lines:
      aum_diffs_penalty now correctly handles denominator="rate" (previously
    Cannot process chunk/lines:
      there could be problems if there are more examples in error table than
```

The first few lines of my old NEWS file were

```{r}
NEWS <- "~/R/aum/NEWS"
NEWS <- "~/teaching/regex-tutorial/NEWS/old/aum.txt"
NEWS.lines <- readLines(NEWS)
cat(head(NEWS.lines, 19), sep="\n")
```

Since there is some structure, we can use a regular expression to
parse the news items into a data table...

```{r}
change.dt <- nc::capture_all_str(
  NEWS.lines,
  version="[0-9]+[.][0-9]+[.][0-9]+",
  "\\s*\n",
  changes="(?:[^0-9].*\n*)*")
change.dt[, .(version, changes=substr(changes,1,50))]
```

...then convert them into the correct format,

```{r}
change.dt[, change.list := strsplit(changes, "\n\n")]
change.dt[, new.str := sapply(change.list, function(change.vec){
  no.newline <- gsub("\n", " ", change.vec)
  with.dash <- paste0("- ", no.newline)
  paste(with.dash, collapse="\n")
})]
change.dt[, new.block := sprintf(
  "Changes in version %s\n\n%s", version, new.str)]
out.str <- paste(change.dt$new.block, collapse="\n\n")
cat(out.str)
```
