---
layout: post
title: Indirect reverse dependencies
description: Computing the entire graph, and histogram tutorial
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2022-10-19-indirect-rev-deps/"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=16,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
if(FALSE){
  knitr::knit("2022-10-19-indirect-rev-deps.Rmd")
}
```

For a recent grant proposal submission to the National Science
Foundation POSE program, I wanted to make an argument that the
`data.table` R package is one of the most used, out of all R
packages.

I therefore wrote some code to compute the number of indirect reverse
dependencies for all R packages. First, download meta-data from the
current CRAN,

```{r}
library(data.table)
if(!file.exists("packages.rds")){
  u <- paste0(
    "https://cloud.r-project.org/web/packages/packages.rds")
  download.file(u, "packages.rds")
}
packages <- readRDS("packages.rds")
pkg.dt <- data.table(packages)[is.na(Path)]
nrow(pkg.dt)
```

The output above shows the number of packages on CRAN.

## Direct reverse imports

Then, we can get direct reverse imports,

```{r}
imp.by <- pkg.dt[, .(
  imported.by=strsplit(`Reverse imports`, ", ")[[1]]
), by=.(Imports=Package)]
imp.by[!is.na(imported.by)]
```

The table above contains one row for each direct reverse import listed
on CRAN.

```{r}
imp.by[Imports=="data.table"]
```

The table above contains a row for each direct reverse import listed
for the `data.table` package.

## Indirect reverse imports

Then, we can use a loop to recursively compute indirect reverse
imports,

```{r}
order.i <- 0
ord.dt.list <- list()
order.pkgs <- pkg.dt[, .(Package, Imports=Package)]
while(nrow(order.pkgs)){
  print(order.i <- order.i+1)
  order.deps <- imp.by[
    order.pkgs, 
    on="Imports", nomatch=0L, allow.cartesian=TRUE]
  if(nrow(order.deps)){
    ord.dt.list[[order.i]] <- data.table(order.i, order.deps)
  }
  order.pkgs <- unique(order.deps[, .(Package, Imports=imported.by)])
}
(ord.dt <- do.call(rbind, ord.dt.list))
```

The table above contains one row for each reverse import (direct or
indirect). Direct reverse imports have `order.i=1` and indirect have
larger values.

Below we check that the number of packages in this table is the same
as the number of packages in the CRAN meta-data,

```{r}
rbind(pkgs.in.ord=length(unique(ord.dt$Package)), pkgs=nrow(pkg.dt))
```

The output above indicates that the table of reverse imports was
computed correctly (the total number of packages is correct).

## Subset of packages funded by NSF

Which packages were funded by NSF?

```{r}
pkg.dt[, Desc.no.newlines := gsub("\n\\s+", " ", Description)]
(nsf.pkgs <- pkg.dt[
  grep("NSF|National Science Foundation", Desc.no.newlines), 
  Package])
```

Which of the reverse imports were funded by NSF?

```{r}
dt.pkgs <- ord.dt[Package=="data.table", imported.by]
(int.pkgs <- intersect(nsf.pkgs, dt.pkgs))
```

The output above shows that there are four reverse imports that were
also funded by NSF.

## Computing path from reverse import in dependency graph

What is the path of packages from each reverse import in the
dependency graph?

```{r}
select.dt <- data.table(
  rev.dep=int.pkgs,
  imported.by=int.pkgs, 
  Package="data.table")
path.dt.list <- list()
iteration <- 0
while(nrow(select.dt)){
  print(iteration <- iteration+1)
  dep.dt <- ord.dt[select.dt, on=.(imported.by, Package)]
  path.dt.list[[iteration]] <- dep.dt
  select.dt <- dep.dt[
    Imports != Package, 
    .(rev.dep, imported.by=Imports, Package)]
}
(path.dt <- do.call(rbind, path.dt.list)[
  order(Package, rev.dep, -order.i),
  .(Package, rev.dep, order.i, imported.by, Imports)])
```

The table above shows that there are two direct reverse imports
(futureheatwaves, hurricaneexposure) for which the path length (max
`order.i`) is 1. The `fdapace` package imports `Hmisc` which imports
`data.table` (path length 2), whereas `bizicount` imports `DHARMa`
which imports `gap` which imports `plotly` which imports `data.table`
(path length 4).

## Unique reverse imports, direct or all

The code below counts the number of unique reverse imports (either
direct or all), for each CRAN package,

```{r}
(dep.type.counts <- rbind(
  data.table(dep.type="all", max.order=Inf),
  data.table(dep.type="direct", max.order=1)
)[, {
  ord.dt[order.i <= max.order, .(
    rev.imports=length(unique(na.omit(imported.by)))
  ), by=Package
  ][, `:=`(
    rank=rank(-rev.imports),
    prop.bigger=1-rank(rev.imports)/.N
  )][order(rank)]
}, by=dep.type])
```

The `data.table` package appears near the top in terms of number of
dependent packages,

```{r}
(dt.counts <- dep.type.counts[Package=="data.table"])
```

The table above shows that `data.table` has 

* 1326 direct reverse imports, which is rank 11 among CRAN packages
  (only 10 other packages, 0.05% of all CRAN packages, have a larger
  number of reverse imports).
* 2661 packages which import either directly or indirectly, which is
  rank 45 among CRAN packages (only 44 other packages, 0.2% of all
  CRAN packages, have a larger number).
  
## Comparison with other CRAN packages

The code below can be used to compute a histogram of reverse import
counts, among all CRAN packages. To compute a histogram using a
`data.table` rolling join, we first need to look at the range of the
data, and then use the min/max to define the sequence of midpoints of
histogram bins.

```{r}
log10(range(dep.type.counts$rev.imports))
log10.min <- 0
log10.max <- 4
(bin.dt <- data.table(
  log10.bin=c(-Inf,seq(log10.min, log10.max, by=0.5))
)[, bin := 10^log10.bin][, Bin := round(bin)][])
```

The bins defined above range from 1 to 10000 on the log scale, with an
additional bin for 0 (packages that have no reverse imports). Below we
use a rolling join to figure out which packages are closest to each
bin midpoint on the log scale,

```{r}
dep.type.counts[, log10.rev.imports := log10(rev.imports)]
(bin.log.join <- bin.dt[dep.type.counts, .(
  dep.type, Package, Bin, bin,
  log10.bin=x.log10.bin, 
  log10.rev.imports),
  roll="nearest", 
  on=.(log10.bin=log10.rev.imports)])
```

The table above has columns for bin midpoint on the log scale,
`log10.bin`, as well as the actual number of reverse imports,
`log10.rev.imports`. We can compute a histogram by summarizing for
each bin,

```{r}
(bin.log.hist <- bin.log.join[, .(
  n.packages=.N
), by=.(dep.type, bin)])
```

The table above has a column `n.packages` which shows the number of
packages which are closest to each bin. These numbers can be plotted
to compare with the corresponding number of reverse imports for the
`data.table` package,

```{r compareLine}
library(ggplot2)
ggplot()+
  theme_bw()+
  geom_line(aes(
    bin, n.packages),
    color="grey50",
    size=2,
    data=bin.log.hist)+
  geom_vline(aes(
    xintercept=rev.imports),
    data=dt.counts)+
  geom_text(aes(
    rev.imports*0.9, Inf, 
    label=sprintf(paste(
      "data.table rev. imports=%d",
      "rank=%d/%d packages",
      "%.2f%% of packages have",
      "larger rev. dep. count",
      sep="\n"),
      rev.imports, rank, nrow(pkg.dt), prop.bigger*100)),
    data=dt.counts,
    hjust=1, 
    vjust=1.1)+
  facet_grid(. ~ dep.type, labeller=label_both)+
  scale_y_log10(paste(
    "Number of packages with that",
    "reverse dependency count (log scale)",
    sep="\n"))+
  scale_x_log10(paste(
    "Reverse dependency count",
    "(number of other packages which import, log scale)"))
```

The figure above shows that `data.table` is ranked near the top, when
comparing with other CRAN packages in terms of number of reverse
imports.

## Appendix: rolling join in original or log space?

In the code above, to compute a histogram that we want to display in
the log space, we did a rolling join in the log space, which results
in symmetric histogram bins. If we did the join in the original space,
then the histogram bins would have been asymmetric, as the code below
shows,

```{r theoretical-bin-size, fig.height=1}
bin.tall <- melt(
  bin.dt[bin>0], 
  measure.vars=c("log10.bin", "bin"), 
  id.vars=c("Bin", "log10.bin"))
bin.tall[, next.break := c(value+c(diff(value)/2,NA)), by=variable]
bin.tall[, next.log10 := ifelse(
  variable=="bin", log10(next.break), next.break)]
bin.tall[, prev.log10 := c(NA, next.log10[-.N]), by=variable]
bin.tall[, distance := ifelse(variable=="bin", "original", "log")]
bin.not.na <- bin.tall[!(is.na(prev.log10)|is.na(next.log10))]
ggplot()+
  geom_segment(aes(
    prev.log10, distance,
    xend=next.log10, yend=distance),
    data=bin.not.na)+
  geom_point(aes(
    log10.bin, distance),
    data=bin.not.na)+
  facet_grid(. ~ Bin, labeller=label_both, scales="free")+
  scale_x_continuous(breaks=seq(log10.min, log10.max, by=0.1))
```

The figure above shows the center of each histogram bin as a dot, and
the min/max extent of each bin as a line segment. It is clear that
using original space distances for the join results in bins which are
asymmetric, in the sense that the bin will count more data which is
larger than the bin center.

Another way of looking at it is empirically, in terms of the reverse
imports data,

```{r histLog}
bin.log.join[, log10.diff := ifelse(
  log10.rev.imports == -Inf, 0, log10.rev.imports-log10.bin)]
bin.width <- 0.05
ggplot()+
  geom_histogram(aes(
    log10.diff, after_stat(ncount)),
    binwidth=bin.width,
    data=bin.log.join)+
  scale_x_continuous(breaks=seq(-1, 1, by=0.2))+
  geom_point(aes(x,y),data=data.table(x=0,y=0))+
  facet_grid(dep.type ~ Bin, labeller=label_both)
```

The histogram above shows the distribution of differences between the
actual number of reverse imports and the corresponding bin
center. Most of the differences fall between -0.2 and 0.2, which is to
be expected, since `bin.dt` used a bin size of 0.5 on the log scale.

What if we did the same computation using the original distance scale?

```{r histOriginal}
bin.join <- bin.dt[dep.type.counts, .(
  dep.type, Package, Bin,
  log10.bin=x.log10.bin, 
  rev.imports),
  roll="nearest", 
  on=.(bin=rev.imports)
  ][, log10.rev.imports := log10(rev.imports)
  ][, log10.diff := ifelse(
    log10.rev.imports == -Inf, 0, log10.rev.imports-log10.bin)]
ggplot()+
  geom_histogram(aes(
    log10.diff, after_stat(ncount)),
    binwidth=bin.width,
    data=bin.join)+
  scale_x_continuous(breaks=seq(-1, 1, by=0.2))+
  geom_point(aes(x,y),data=data.table(x=0,y=0))+
  facet_grid(dep.type ~ Bin, labeller=label_both)
```

The bin assignments above were computed using a rolling join in the
original reverse import count space (not the log space), so each
histogram above is skewed to the right (the are more larger
differences in each bin), as expected based on our theoretical
analysis above which showed the asymmetric bins.

Finally we can compare the two histograms in the same plot. First we
combine the two data sets created from rolling joins,

```{r}
joinDT <- function(DT, distance, sign){
  DT[, .(distance, sign, log10.diff, dep.type, Bin)]
}
(bin.both.join <- rbind(
  joinDT(bin.join, "original", -1),
  joinDT(bin.log.join, "log", sign=1)))
```

Then we compute histograms ourselves in the code below using the
`hist` function (instead of using `geom_histogram` as we did in the
code above),

```{r}
max.abs.diff <- 0.4
breaks.vec <- seq(-max.abs.diff, max.abs.diff, by=bin.width)
(bin.both.hist <- bin.both.join[, {
  hlist <- hist(log10.diff, breaks.vec, plot=FALSE)
  with(hlist, data.table(log10.diff=mids, ncount=counts/max(counts)))
}, by=.(distance, sign, dep.type, Bin)])
```

The table above contains histograms of differences between actual
numbers of reverse imports, and the corresponding bin centers. The
`ncount` column is normalized between 0 and 1 so that the histograms
can be displayed on a common scale. In the code below we also compute
the mean of the differences, to see if there is any skew to larger
values than the bin center, as we would expect.

```{r}
(bin.both.stats <- dcast(
  bin.both.join,
  distance + sign + dep.type + Bin ~ .,
  list(length, mean),
  value.var = "log10.diff"))
```

The table above has number of packages and mean difference columns,
for every bin/histogram to plot, which we show below,

```{r histCompare}
ggplot()+
  theme(legend.position=c(0.95, 0.2))+
  geom_bar(aes(
    log10.diff, sign*ncount, fill=distance),
    stat="identity",
    data=bin.both.hist)+
  geom_text(aes(
    0, sign, label=sprintf(
      "%d packages", log10.diff_length)),
    data=bin.both.stats)+
  geom_point(aes(
    x,y),
    size=4,
    data=data.table(x=0,y=0))+
  geom_point(aes(
    log10.diff_mean, 0, fill=distance),
    data=bin.both.stats,
    size=3,
    shape=21)+
  facet_grid(dep.type ~ Bin, labeller=label_both)
```
