---
layout: post
title: R Package Release History
description: Extracting and plotting data from CRAN web site 
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2022-11-02-release-history/"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=7,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
options(width=200)
if(FALSE){
  knitr::knit("2022-11-02-release-history.Rmd")
}
```

For a recent grant proposal submission to the National Science
Foundation POSE program, I wanted to make a figure that shows the
release history of the `data.table` R package.

## Download Archive web page

First, we create a releases directory to cache the
downloaded web pages,

```{r}
releases.dir <- "~/grants/2022-NSF-POSE/releases"
dir.create(releases.dir, showWarnings = FALSE)
```

We can download an Archive web page via the code below,

```{r}
Archive <- "https://cloud.r-project.org/src/contrib/Archive/"
get_Archive <- function(Package){
  pkg.html <- file.path(releases.dir, paste0(Package, ".html"))
  if(!file.exists(pkg.html)){
    u <- paste0(Archive, Package)
    download.file(u, pkg.html)
  }
  readLines(pkg.html)
}
(Archive.data.table <- get_Archive("data.table"))
```

The output above shows that the Archive web page has a regular
structure.

## Parsing with a regular expression

The Archive web page can be parsed via a regular expression,

```{r}
naive.pattern <- list(
  "_",
  version="[0-9.]+",
  "[.]tar[.]gz</a>\\s+",
  date.str=".*?",
  "\\s")
```

The code above specifies a regular expression:

* `"_"` means to start by matching an underscore,
* `version="[0-9.]+"` means to match one or more digits or dots, and
  output them in the `version` column,
* `"[.]tar[.]gz</a>\\s+"` means to match the `.tar.gz` file name
  suffix, the closing `</a>` tag, and then one or more white space
  characters,
* `date.str=".*?"` means to match zero or more characters (non-greedy,
  as few as possible), and output them in the `date.str` column,
* `"\\s"` means to match one white space character.

The end result is a table with one row for each matched package
version, and one column for each of the named arguments:

```{r}
(naive.dt <- nc::capture_all_str(Archive.data.table, naive.pattern))
```

Above the table shows all matches for the given pattern. How do we
know that it has captured all the patterns? We can use a simpler
pattern to get all of the packages,

```{r}
(tar.gz.vec <- grep("tar[.]gz", Archive.data.table, value=TRUE))
```

The output above shows 56 lines, which means there were some that were
not captured by the regex. Which ones were they?

```{r}
(missed.subjects <- grep(
  nc::var_args_list(naive.pattern)$pattern, 
  tar.gz.vec, value=TRUE, invert=TRUE))
```

The output above indicates 3 missed subjects. Another way to determine
we missed them would be to use that pattern as follows,

```{r err, error=TRUE, purl=TRUE}
nc::capture_first_vec(tar.gz.vec, naive.pattern)
tar.gz.vec[c(5,6,7)]
```

Below we revise the pattern so that it matches all of the tar files,
including those with a hyphen in the version,

```{r}
revised.pattern <- list(
  "_",
  version="[0-9.-]+",
  "[.]tar[.]gz</a>\\s+",
  date.str=".*?",
  "\\s")
nc::capture_first_vec(tar.gz.vec, revised.pattern)
```

## Analyze several packages for comparison

The code below defines a set of four packages for which we would like
to analyze the release history (tidyverse packages for comparison).

```{r}
library(data.table)
compare.pkg.dt <- rbind(
  data.table(project="tidyverse", Package=c("readr","tidyr","dplyr")),
  data.table(project="deprecated", Package=c("reshape2", "plyr")),
  data.table(project="data.table", Package="data.table"))
```

In the code below, we do the same thing for each package, 

```{r}
(release.dt <- compare.pkg.dt[, {
  Archive.pkg <- get_Archive(Package)
  nc::capture_all_str(Archive.pkg, revised.pattern)
}, by=names(compare.pkg.dt)])
```

The result above is a data table with one row for each package
version. Note that the code set `by` to all column names, so that the
code is run for each row/package.

## Add columns for plotting

For plotting we add a few more columns,

```{r}
release.dt[, `:=`(
  IDate = as.IDate(date.str),
  year = as.integer(sub("-.*", "", date.str)),
  package = factor(Package, compare.pkg.dt$Package),
  Project = paste0('\n', project))]
setkey(release.dt, Project, Package, IDate)
release.dt
```

To explain the new columns above,

* `IDate` is for the date to display on the X axis,
* `year` is for labeling the first released version each year,
* `package` is for displaying the Y axis in a particular order
  (defined by the factor levels),
* `Project` is for the facet/panel titles (newline so that minimal
  vertical space is used).
  
## Basic plot

The code below creates a basic version history plot,

```{r points}
library(ggplot2)
(gg.points <- ggplot()+
  theme(
    axis.text.x=element_text(hjust=1, angle=40))+
  facet_grid(Project ~ ., labeller=label_both, scales="free")+
  geom_point(aes(
    IDate, package),
    shape=1,
    data=release.dt)+
  coord_cartesian(expand=5)+
  scale_x_date("Date", breaks="year"))
```

The plot above shows a point for every release to CRAN, so you can see
the distribution of releases over time.

## Add direct labels

Before plotting we make a new table which contains only the first
release of `data.table` in each year (for direct labels),

```{r}
(labeled.releases <- release.dt[Package=="data.table", .SD[1], by=year])
```

```{r points-labels}
gg.points+
  directlabels::geom_dl(aes(
    IDate, package, label=paste0(year, "\n", version)),
    method=list(
      cex=0.7,
      directlabels::polygon.method(
        "top", offset.cm=0.2, custom.colors=list(
          colour="white",
          box.color="black",
          text.color="black"))),
    data=labeled.releases)
```
  
The plot above shows a label for the first version released each year.

## Releases per year

One way to compute releases per year would be to add up the total
number of releases, then divide by the number of years,

```{r release-history}
(overall.stats <- dcast(
  release.dt, 
  project + Package ~ ., 
  list(min,max,length), 
  value.var="year"
)[, releases.per.year := year_length/(year_max-year_min+1)][])
```

Another way to do it would be to compute the number of releases in
each year since the release of the package. To do that we first
compute, for each package, a set of years for which we want to count
releases.

```{r}
(max.year <- max(release.dt$year))
(years.since.release <- release.dt[, .(
  year=seq(min(year), max.year)
), by=.(Project, project, Package, package)])
```

Then we can do a join and summarize to count the number of releases in
each year, for each package,

```{r}
(releases.per.year <- release.dt[years.since.release, .(
  N=as.numeric(.N)
), on=.NATURAL, by=.EACHI])
```

Note that `on=.NATURAL` above means to join on the common columns
between the two tables, and `by=.EACHI` means to compute a summary for
each value specified in `i` (the first argument in the square bracket).
We can plot these data as a heat map via

```{r heatMap}
ggplot()+
  theme_bw()+
  theme(panel.spacing=grid::unit(0, "lines"))+
  geom_tile(aes(
    year, package, fill=log(N+1)),
    data=releases.per.year)+
  geom_text(aes(
    year, package, label=N),
    data=releases.per.year)+
  facet_grid(Project ~ ., labeller=label_both, scales="free", space="free")+
  scale_fill_gradient("releases\n(log scale)", low="white", high="red")+
  scale_x_continuous(breaks=seq(2006, 2022, by=2))+
  coord_cartesian(expand=FALSE)
```

The heat map above shows a summarized display of the release data we
saw earlier in the dot plot.

Next, we can apply a list of summary functions over all of the yearly
counts, for each package, via

```{r}
(per.year.stats <- dcast(
  releases.per.year,
  project + Package ~ .,
  list(min, max, mean, sd, length),
  value.var = "N"))
```

Finally, the code below creates a table to compare the two different ways of
computing the number of releases per year,

```{r}
per.year.stats[overall.stats, .(
  Package, 
  overall.mean=N_mean, 
  mean.per.year=releases.per.year
), on="Package"]
```

The table above show similar numbers for the two methods of computing
the number of releases per year.

## Conclusion

We have shown how to download CRAN package release data, how to parse
the web pages using the `nc` package and a regular expression, how to
summarize/analyze using `data.table`, and how to visualize using
`ggplot2`.
