---
layout: post
title: Interpretable learning algorithms with built-in feature selection
description: Regularized linear model and decision tree
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2023-11-30-glmnet-interpretation"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path,"/"),
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

Machine learning algorithms input a train data set, and output a
prediction function. This post is about interpreting that prediction
function, in terms of what input features in the data are used to
compute predictions. 

## Introduction to model interpretation

Most machine learning algorithms output a prediction function that
uses all of the input features in the train data set. In the special
case of feature selection algorithms, a subset of input features is
used in the prediction function. For example, the L1 regularized
linear learning algorithm (R package glmnet) outputs a
coefficient/weight vector with some values set to zero. We can
therefore say that the model is interpretable in terms of the
different input feature subsets:

* For the features with weights equal to zero, these features are
  completely ignored for the purposes of prediction (non-important
  subset of features).
* For the features with weights not equal to zero, these features are
  used to compute predictions (important subset of features).

In the next sections, we explain how to compute and interpret this
algorithm using base R.

## Data simulation

For the purposes of simulation, we use the simulated data below:

```{r}
N <- 3000
library(data.table)
set.seed(1)
n.features <- 9
(full.dt <- data.table()[
, paste0("x",1:n.features) := replicate(n.features, rnorm(N), simplify=FALSE)
][
, label := factor(ifelse(x1*2+x2-x3-x4*2+rnorm(N) < 0, "not spam", "spam"))
][])
table(full.dt$label)
```

We can imagine a spam filtering system, with training data for which each row in the table above represents a message which has been labeled as spam or not.
In the table above, there are two sets of features:

* `x1` to `x4` are used to define the output `label` (and should be used in the best prediction function)
* other features are random noise (should be ignored by the best prediction function)
  
In the next section, we run the L1 regularized linear learning
algorithm on these data, along with another interpretable algorithm
(decision tree).

## mlr3 training

To use the mlr3 framework on our simulated data, we begin by
converting the data table to a task in the code below,

```{r}
(task.classif <- mlr3::TaskClassif$new(
  "simulated", full.dt, target="label"
)$set_col_roles("label", c("target", "stratum")))
```

The output above shows that we have created a task named simulated,
with target column named label, and with several features (x1
etc). The output also indicates the label column is used as a stratum,
which means that when sampling, the proportion of each label in the
subsample should match the proportion in the total data.

Below we create a resampling object that will vary the size of the train set,

```{r}
size_cv <- mlr3resampling::ResamplingVariableSizeTrainCV$new()
size_cv$param_set$values$min_train_data <- 10
size_cv$param_set$values$random_seeds <- 4
size_cv
```

The output above indicates the resampling involves 3 cross-validation
fols, 20 min train data, 4 random seeds, and 5 train sizes. All of
these choices are arbitrary, and do not have a large effect on the end
results. Exercise for the reader: play with these values, re-do the
computations, and see if you get similar results. (you should!)

Below we define a list of learning algorithms, and note the
`cv_glmnet` learner uses cross-validation, with the given number of
folds (below 6), to select the optimal degree of L1 regularization
(which maximizes prediction accuracy). Note that this `nfolds`
parameter controls the subtrain/validation split, and is different
from the `folds` parameter of `size_cv` (which controls the train/test
split, useful for comparing prediction accuracy of learning
algorithms).

```{r}
cv_glmnet <- mlr3learners::LearnerClassifCVGlmnet$new()
cv_glmnet$param_set$values$nfolds <- 6
(learner.list <- list(
  cv_glmnet,
  mlr3::LearnerClassifRpart$new(),
  mlr3::LearnerClassifFeatureless$new()))
```

The output above shows a list of three learning algorithms. 

* `cv_glmnet` is the L1 regularized linear model, which will set some
  weights to zero (selecting the other features).
* `rpart` is another learning algorithm with built-in feature
  selection, which will be discussed below.
* `featureless` is a baseline learning algorithm which always predicts
  the most frequent label in the train set. This should always be run
  for comparison with the real learning algorithms (which will be more
  accurate if they have learned some non-trivial relationship between
  inputs/features and output/target).
  
Below we define a benchmark grid, which combines our task, with
learners, and the resampling,

```{r}
(bench.grid <- mlr3::benchmark_grid(
  task.classif,
  learner.list,
  size_cv))
```

The output above is a table with one row for each combination of task,
learner, and resampling. 

Below we first define a future plan to do the computations in
parallel, then set log threshold to reduce output, then compute the
benchmark result.

```{r}
if(require(future))plan("multisession")
lgr::get_logger("mlr3")$set_threshold("warn")
(bench.result <- mlr3::benchmark(
  bench.grid, store_models = TRUE))
```

The output above shows the number of resampling iterations computed.

## interpreting prediction error rates on test set

The code below computes scores (test error), for each resampling iteration.

```{r}
bench.score <- mlr3resampling::score(bench.result)
bench.score[1]
```

The output above shows the result of one resampling
iteration. Important columns include

* `train_size`, number of samples in train set.
* `classif.ce`, test error (mis-classification rate).
* `algorithm`, learning algorithm.
* `test.fold`, test fold number in cross-validation.
* `seed`, random seed used to determine sampling order through train set.

Below we plot the results,

```{r testError}
train_min_size_vec <- unique(bench.score[["train_min_size"]])
library(animint2)
ggplot()+
  scale_x_log10(breaks=train_min_size_vec)+
  scale_y_continuous(
    "Classification error on test set")+
  geom_line(aes(
    train_min_size, classif.ce,
    group=paste(algorithm, seed),
    color=algorithm),
    shape=1,
    data=bench.score)+
  geom_point(aes(
    train_min_size, classif.ce, color=algorithm),
    shape=1,
    data=bench.score)+
  facet_grid(
    test.fold~task_id,
    labeller=label_both)
```

The figure above has a panel for each test fold in cross-validation.
There is a line for each algorithm, and for each random seed.  The
plot is test error as a function of train size, so we can see how many
samples are required to learn a reasonable prediction function.  It is
clear that a small number of samples (20) is not sufficient for either
learning algorithm, and a large number of samples (2000) is enough to
learn good predictions (with significantly smaller error rate than
featureless). Interestingly, the linear model is actually more
accurate than the decision tree, for intermediate and large data sizes.
This makes sense, because `label` was defined using a linear function.

## interpreting linear model

In this section we show how to interpret the learned linear models, in
terms of the weights. First we consider the subset of score table rows
which correspond to the linear model. Then we loop over each row,
computing the weight vector learned in each train/test split.
We then combine the learned weights together in a single data table.

```{r}
library(glmnet)
glmnet.score <- bench.score[algorithm=="cv_glmnet"]
weight.dt.list <- list()
levs <- grep("^x", names(full.dt), value=TRUE)
for(score.i in 1:nrow(glmnet.score)){
  score.row <- glmnet.score[score.i]
  fit <- score.row$learner[[1]]$model
  weight.mat <- coef(fit)[-1,]
  weight.dt.list[[score.i]] <- score.row[, .(
    test.fold, seed, train_min_size,
    weight=as.numeric(weight.mat),
    variable=factor(names(weight.mat), levs))]
}
(weight.dt <- rbindlist(weight.dt.list))
```

The output above shows one row for each weight learned in each
train/test split. Most weights are zero (not used for prediction), due
to L1 regularization.
We use the code below to visualize these weights.

```{r linearWeights}
weight.non.zero <- weight.dt[weight!=0]
ggplot()+
  facet_grid(test.fold ~ seed, labeller=label_both)+
  scale_y_discrete(breaks=levs,drop=FALSE)+
  geom_tile(aes(
    train_min_size, variable, fill=weight),
    data=weight.non.zero)+
  scale_fill_gradient2()+
  scale_x_log10(breaks=train_min_size_vec)+
  theme(axis.text.x=element_text(angle=60, hjust=1))
```

The heat map above shows a tile for each seed, train size, test fold,
and variable. Missing tiles (grey background) indicate zero weights
(not used for prediction. Recall that in our simulated data, there was
only one signal feature (x0) and the others are noise that should be
ignored. It is clear that at small train sizes, there are some false
positive non-zero weights, and there are also false negatives (weight
for x0 should not be zero). For large train sizes, the L1
regularization does a good job of selecting only the important
variable (x0 has negative weight, and others have zero weight).

Typically positive weights mean that larger feature values mean more
likelihood of being classified as positive (and negative
weights/smaller feature values would be the opposite), but mlr3 seems
to invert the glmnet weights, [which I reported as an
issue](https://github.com/mlr-org/mlr3learners/issues/281).

Another method which I typically use for interpreting L1 regularized
linear models involves counting folds/splits with non-zero weight, for each
variable. To do that we can use the code below,

```{r weightsForEachSeedSize}
ggplot()+
  scale_y_discrete(breaks=levs,drop=FALSE)+
  theme_bw()+
  geom_vline(xintercept=0, color="grey50")+
  geom_point(aes(
    weight, variable),
    shape=1,
    data=weight.non.zero)+
  facet_grid(seed ~ train_min_size, labeller=label_both)
```

The plot above shows a point for each non-zero linear model weight,
with one panel for each seed and train size. There is a vertical grey
line to emphasize a weight value of zero. It is clear that large train
sizes result in three points/folds with non-zero weights, for the
signal feature `x0`, and zero weights for the other noise features.

We can also use the number of folds with non zero weights as a metric
for variable importance, as we compute in the code below.

```{r facetNonZero}
(one.panel <- weight.non.zero[
  train_min_size==64 & seed==2
][
, non.zero.folds := .N, by=variable
])
var.ord.dt <- one.panel[, .(
  mean.weight=mean(weight)
), by=.(variable, non.zero.folds)
][order(-non.zero.folds, -abs(mean.weight))]
var.ord.levs <- paste(var.ord.dt$variable)
one.panel[, Variable := factor(variable, var.ord.levs)]
ggplot()+
  theme_bw()+
  geom_vline(xintercept=0, color="grey50")+
  geom_point(aes(
    weight, Variable),
    shape=1,
    data=one.panel)+
  facet_grid(non.zero.folds ~ ., scales="free", space="free")
```

The plot above has a panel for each value of `non.zero.folds` -- the
variables that appear in the larger panel numbers are more important
(have non-zero weights in more folds).
Also, within each panel, the most important variables (with largest absolute weight) appear near the bottom.

## interpreting decision tree

Another machine learning algorithm with built-in feature selection is
the decision tree, which we can interpret using the code below.

```{r}
rpart.score <- bench.score[algorithm=="rpart"]
decision.dt.list <- list()
for(rpart.i in 1:nrow(rpart.score)){
  rpart.row <- rpart.score[rpart.i]
  rfit <- rpart.row$learner[[1]]$model
  decision.dt.list[[rpart.i]] <- rpart.row[, .(
    test.fold, seed, train_min_size,
    rfit$frame
  )][var!="<leaf>"]
}
(decision.dt <- rbindlist(decision.dt.list))
```

The code above examines the splits which are used in each decision
tree, and outputs a table above with one row per split used.  The code
below computes a table with one row per variable, with additional
columns splits and samples to measure importance.

```{r}
(var.dt <- decision.dt[, .(
  samples=sum(n),
  splits=.N
), by=.(test.fold, seed, train_min_size, variable=factor(var, levs))])
```

The code below computes the proportion of samples used in each split,
which is another measure of importance.

```{r}
var.dt[
, split.sample.prop := samples/sum(samples)
, by=.(test.fold, seed, train_min_size)
][]
```

The code below makes a heatmap with a tile for each variable which was
used in at least one split of the learned decision tree. Larger values
of `split.sample.prop` (more red) indicate variables which are more
important (used with more samples).

```{r rpartHeatmap}
ggplot()+
  facet_grid(test.fold ~ seed, labeller=label_both)+
  geom_tile(aes(
    train_min_size, variable, fill=split.sample.prop),
    data=var.dt)+
  scale_y_discrete(breaks=levs,drop=FALSE)+
  scale_fill_gradient(low="white", high="red")+
  scale_x_log10(breaks=train_min_size_vec)+
  theme(axis.text.x=element_text(angle=60, hjust=1))
```

Finally we make an analogous plot to the linear model below.

```{r rpartNonZero}
(one.rpart <- var.dt[
  train_min_size==632 & seed==4
][
, non.zero.folds := .N, by=variable
])
rpart.ord.dt <- one.rpart[, .(
  mean.prop=mean(split.sample.prop)
), by=.(variable, non.zero.folds)
][order(-non.zero.folds, -mean.prop)]
rpart.ord.levs <- paste(rpart.ord.dt$variable)
one.rpart[, Variable := factor(variable, rpart.ord.levs)]
ggplot()+
  theme_bw()+
  geom_point(aes(
    split.sample.prop, Variable),
    shape=1,
    data=one.rpart)+
  facet_grid(non.zero.folds ~ ., scales="free", space="free")
```

The plot above shows variable importance in the learned rpart model
(decision tree).  It shows the mean proportion of samples used in each
split of the decision tree, for each variable, with one point for each
fold in which that variable appeared in the decision tree (un-used
variables do not appear in the plot).  The number in the panel shows
the number of folds for which this variable was used in the decision
tree. The most important variables are sorted to the bottom of the
plot (and x1 and x4 correctly appear there).

## Conclusions

We have shown how to interpret two kinds of feature selection machine
learning algorithms, L1 regularized linear models and decision trees,
after having learned them using the mlr3 framework in R.

## Session info

```{r}
sessionInfo()
```


