---
layout: post
title: Historical reverse imports
description: Analysis of R package usage over time
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2022-05-13-rev-imp-history/"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=7)
if(FALSE){
  knitr::knit("2023-02-06-mran-meta-backup.Rmd")
}
```

The Microsoft R Application Network (MRAN) [time
machine](https://mran.microsoft.com/documents/rro/reproducibility#timemachine)
keeps daily CRAN snapshots, so we can access historical packages.rds
files going back to its inception on September 17th, 2014. However it
is going offline in July 2023, so here we explain how to make a
backup.

```{r}
library(data.table)
get_packages <- function(date){
  date.str <- paste(date)
  date.dir <- file.path("~/R/dt-deps-time", date.str)
  dir.create(date.dir,showWarnings=FALSE,recursive=TRUE)
  packages.rds <- file.path(date.dir, "packages.rds")
  tryCatch({
    if(!file.exists(packages.rds)){
      u <- paste0(
        "https://cran.microsoft.com/snapshot/",
        date.str,
        "/web/packages/packages.rds")
      print(packages.rds)
      download.file(u, packages.rds)
    }
    packages <- readRDS(packages.rds)
    data.table(packages)
  }, error=function(e){
    NULL
  })
}
pkg.dt <- get_packages("2014-09-17")
names(pkg.dt)
(rev.imports.str <- pkg.dt["data.table", on="Package"][["Reverse imports"]])
```

The code below downloads meta-data for every day,

```{r}
date.vec <- seq(as.IDate("2014-09-17"), as.IDate(Sys.time()), by="day")
date.pkgs.list <- list()
for(date.i in seq_along(date.vec)){
  date <- date.vec[[date.i]]
  pkg.dt <- get_packages(date)
  value <- if(is.data.table(pkg.dt)){
    dim(pkg.dt)
  }else{
    c(0,0)
  }
  date.pkgs.list[[paste(date)]] <- data.table(
    date,
    variable=c("nrow","ncol"),
    value)
}
(date.pkgs <- rbindlist(date.pkgs.list))
```

Finally we can plug these count data into a ggplot,

```{r}
library(ggplot2)
ggplot()+
  theme_bw()+
  facet_grid(variable ~ ., scales="free")+
  geom_point(aes(
    date, value),
    shape=21,
    data=date.pkgs)
```

We can see in the figure above that the number of packages (nrow) has
been increasing over time until 2022, when there was a period where
more packages were removed than packages added. Also we see that the
number of fields/features per package (ncol) has generally increased
over time, from 41 in 2014, to more than 60 currently. (the one
exception is when a field was removed in 2019)

