---
layout: post
title: Weighted binary classification
description: An interesting real-world example
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2023-12-30-weighted-binary-classification"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this blog post is to explain weighted binary classification.

### Introduction



### Moonwalking with Einstein, by Joshua Foer, The Penguin Press (2011), pages 50 to 52.

"From the perspective of the poultry farmer, male chickens are
useless. They can't lay eggs, their meat is stringy, and they're
ornery to the hens that do all the hard work of putting food on our
tables."

"The sooner they can be disposed of---often they're ground into animal
feed---the better. But a costly problem has vexed egg farmers for
millennia: It's virtually impossible to tell the difference between
male and female chickens until they're four to six weeks old, when
they begin to grow distinctive feathers and secondary sex
characteristics like the rooster's comb. Until then, they're all just
indistinguishable fluff balls that have to be housed and fed---at
considerable expense."

TDH: karyotype sexing is technically possible but expensive. 

"Somehow it took until the 1920s before anyone figured out a solution to this costly dilemma. The momentous discovery was made by a group of Japanese veterinary scientists, who realized that just inside the chick's rear end there is a constellation of folds, marks, spots, and bumps that to the untrained eye appear arbitrary, but when properly read, can divulge the sex of a day-old bird. When this discovery was unveiled at the 1927 World Poultry Congress in Ottawa, it revolutionized the global hathery industry and eventually lowered the price of eggs worldwide."

"Chicken sexing is a delicate art, requiring Zen-like concentration and a brain surgeon's dexterity. The bird is cradled in the left hand and given a gentle squeeze...the sexer flips the bird over and parts a small flap on its hindquarters to expose the cloaca, a tiny vent where both the genitals and anus are situated, and peers deep inside. In the simple cases---the ones that the sexer can actually explain---he's looking for a barely perceptible protuberance called the "bead," about the size of a pinhead. If the bead is convex, the bird is a boy, and gets thrown to the left; concave or flat and it's a girl, sent down a chute to the right...But it roughly 80 percent of the chicks, the bead is not obvious and there is no single distinguishing trait the sexer can point to."

TDH: multi-variate decision rule.

"By some estimates there are as many as a thousand different vent configurations that a sexer has to learn to become competent. The job is made even more difficult by the fact that the sexer has to diagnose the bird with just a glance. There is no time for conscious reasoning. If he hesistates for even a couple seconds, his grip on the bird can cause a pullet's vent to swell to look unquestionably like a cockerel's."

TDH: overlap of classes in feature space (not separable).

"Mistakes are costly. In the 1960s, one hatchery paid its sexers a penny for each correctly sexed chick and deducted 35 cents for each one they got wrong."

TDH: female/pullet/positive, male/cockerel/negative, 35 cents is cost
of false positive, which is when the sexer mistakenly sends a
male/cockerel/negative down the right/female/pullet/positive chute.

"The best in the business can sex 1,200 chicks an hour with 98 to 99
percent accuracy. In Japan, a few superheroes of the industry have
learned how to double clutch the chicks and sex them two at a time, at
the rate of 1,700 per hour."

TDH: compute money.

```{r}
library(data.table)
(rate.per.hour.dt <- CJ(
  chicks=c(1200, 1700),
  accuracy.rate=c(50,98,99)/100
)[, `:=`(
  chicks.correct=chicks*accuracy.rate,
  chicks.incorrect=chicks*(1-accuracy.rate)
)][
, cents := chicks.correct-35*chicks.incorrect
][])
```

The [wikipedia page about USA minimum
wage](https://en.wikipedia.org/wiki/Minimum_wage_in_the_United_States)
says that 1968 minimum wage was $1.60, so by my estimates above,
chicken sexers were making decent wages. But my estimates are not
consistent with "five hundred dollars a day" wages for graduates of
the Nippon Chick Sexing School.

"What makes chicken sexing such a captivating subject...is that even
the best professional sexers can't decribe how they determine gender
in the toughest, most ambiguous cases. Their art is inexplicable. They
say that within three seconds they just "know" whether a bird is a boy
or girl, but they can't say how they know."

TDH: explicable ML.

### Genetic basis of chicken sex determination 

Discussed on [wikipedia
ZW_sex-determination_system](https://en.wikipedia.org/wiki/ZW_sex-determination_system).

* Males are ZZ
* Females are ZW
* Ovum determines sex. Why? Every male contributes Z, but female
  contributes 50% Z, 50% W.
* In this system, half of offspring are male, half are female.

### Modern references 

Discussed on [wikipedia In-ovo
sexing](https://en.wikipedia.org/wiki/In-ovo_sexing). Companies in
Europe are now sexing eggs prior to hatching, which is touted as more
humane and less costly.

