---
layout: post
title: CRAN Meta-data
description: Backing up MRAN
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2023-02-06-mran-meta-backup/"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=7)
if(FALSE){
  knitr::knit("2023-02-06-mran-meta-backup.Rmd")
}
```

Recently I submitted a grant to the National Science Foundation, POSE
program, about expanding the open-source ecosystem around the R
package `data.table`. In the grant application, I made some figures
that show how many other R packages depend on `data.table`, and I
showed trends over time, thanks to the historical CRAN meta-data which
is available on the The Microsoft R Application Network (MRAN) [time
machine](https://mran.microsoft.com/documents/rro/reproducibility#timemachine),
which keeps daily CRAN snapshots, so we can access historical
packages.rds files going back to its inception on September
17th, 2014. However it is going offline in July 2023, so here we
explain how to make a backup.

```{r}
library(data.table)
get_packages <- function(date){
  date.str <- paste(date)
  date.dir <- file.path("~/R/dt-deps-time", date.str)
  dir.create(date.dir,showWarnings=FALSE,recursive=TRUE)
  packages.rds <- file.path(date.dir, "packages.rds")
  tryCatch({
    if(!file.exists(packages.rds)){
      u <- paste0(
        "https://cran.microsoft.com/snapshot/",
        date.str,
        "/web/packages/packages.rds")
      print(packages.rds)
      download.file(u, packages.rds)
    }
    packages <- readRDS(packages.rds)
    data.table(packages)
  }, error=function(e){
    NULL
  })
}
pkg.dt <- get_packages("2014-09-17")
names(pkg.dt)
(rev.imports.str <- pkg.dt["data.table", on="Package"][["Reverse imports"]])
```

The code below downloads meta-data for every day,

```{r}
date.vec <- seq(as.IDate("2014-09-17"), as.IDate(Sys.time()), by="day")
date.pkgs.list <- list()
for(date.i in seq_along(date.vec)){
  date <- date.vec[[date.i]]
  pkg.dt <- get_packages(date)
  value <- if(is.data.table(pkg.dt)){
    dim(pkg.dt)
  }else{
    c(0,0)
  }
  date.pkgs.list[[paste(date)]] <- data.table(
    date,
    variable=c("nrow","ncol"),
    value)
}
(date.pkgs <- rbindlist(date.pkgs.list))
```

Finally we can plug these count data into a ggplot,

```{r dimTimeSeries}
library(ggplot2)
ggplot()+
  theme_bw()+
  facet_grid(variable ~ ., scales="free")+
  geom_point(aes(
    date, value),
    shape=21,
    data=date.pkgs)
```

We can see in the figure above that the number of packages (nrow) has
been increasing over time until 2022, when there was a period where
more packages were removed than packages added. Also we see that the
number of fields/features per package (ncol) has generally increased
over time, from 41 in 2014, to more than 60 currently. (the one
exception is when a field was removed in 2019)

UPDATE 23 Feb 2023: I run a cron job every day on NAU Monsoon, and you
can download the CRAN meta-data using:

```{r}
download.file(
  "https://rcdata.nau.edu/genomic-ml/cran-meta/data/2023/02/23/packages.rds",
  "packages.rds")
packages.mat <- readRDS("packages.rds")
nrow(packages.mat)
```

The output above shows the number of packages available on the CRAN,
as of 23 Feb 2023.
