---
layout: post
title: Count unique students
description: Regex and data table summarization
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2023-08-29-count-unique/"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=7)
if(FALSE){
  knitr::knit("2023-08-29-count-unique.Rmd")
}
```

The goal of this post is to show how to count the number of unique
google summer of code students I have mentored. The data source is my
[Teaching web
page](https://github.com/tdhock/tdhock.github.io/blob/master/_pages/teaching.md),
source code shown below:

```{r}
projects.string <- "
I have mentored the
following students in coding free/open-source software.
- [Arthur Pan](https://github.com/ampurrr), 2023, polars in R.
- [Jocelyne Chen](https://github.com/ampurr), 2023, [animint2
  documentation and bug fixes](https://gsoc.ampurr.com).
- [Yufan Fei](https://github.com/Faye-yufan), 2023, animint2:
  interactive grammar of graphics.
- [Yufan Fei](https://github.com/Faye-yufan), 2022, [animint2: interactive
  grammar of
  graphics](https://github.com/Faye-yufan/gsoc22-animint/blob/master/README.md).
- [Fabrizio Sandri](https://github.com/FabrizioSandri), 2022,
  [RcppDeepState: github action for fuzz testing C++ code in R
  packages](https://fabriziosandri.github.io/gsoc-2022-blog/summary/2022/09/08/gsoc-summary.html).
- [Daniel Agyapong](https://github.com/EngineerDanny), 2022, [Rperform
  github action for performance testing R
  packages](https://engineerdanny.github.io/GSOC22-RPerform-Blog/).
- [Anirban Chetia](https://github.com/Anirban166), 2021, [directlabels
  improvements](https://github.com/Anirban166/directlabels).
- [Diego Urgell](https://github.com/diego-urgell), 2021,
  [BinSeg](https://github.com/diego-urgell/BinSeg) efficient C++
  implementation of binary segmentation.
- [Mark Nawar](https://github.com/Mark-Nawar), 2021, [re2r back on CRAN](https://github.com/rstats-gsoc/gsoc2021/wiki/re2r-back-on-CRAN)
- [Sanchit Saini](https://github.com/sanchit-saini), 2020,
  [rtracklayer R package
  improvements](https://github.com/rstats-gsoc/gsoc2020/wiki/rtracklayer-improvements).
- [Himanshu Singh](https://github.com/lazycipher), 2020, [animint2:
  interactive grammar of
  graphics](https://github.com/tdhock/animint2).
- [Julian Stanley](https://github.com/julianstanley), 2020, [Graphical
  User Interface for gfpop R
  package](https://github.com/julianstanley/gfpop-gui).
- [Anirban Chetia](https://github.com/Anirban166), 2020,
  [testComplexity R
  package](https://github.com/Anirban166/testComplexity).
- [Anuraag Srivastava](https://github.com/as4378), 2019, [Optimal
  Partitioning algorithm and opart R
  package](https://github.com/as4378/opart).
- [Avinash Barnwal](https://github.com/avinashbarnwal/), 2019, [AFT
  and Binomial loss functions for
  xgboost](https://github.com/avinashbarnwal/GSOC-2019).
- [Aditya Sam](https://github.com/theadityasam/), 2019, [Elastic net regularized interval regression and iregnet R package](https://theadityasam.github.io/GSOC2019/).
- [Alan Williams](https://github.com/aw1231), 2018,
  [SegAnnDB: machine learning system for DNA copy number analysis](https://github.com/tdhock/SegAnnDB),
  [blog](https://medium.com/alans-gsoc-blog/work-product-a1080d175160).
- [Vivek Kumar](https://github.com/vivekktiwari), 2018,
  [animint2: interactive grammar of graphics](https://github.com/tdhock/animint2),
  [blog](https://vivekktiwari.github.io/gsoc18/).
- [Johan Larsson](https://github.com/jolars), 2018,
  [sgdnet: SAGA algorithm for sparse linear models](https://github.com/jolars/sgdnet).
- [Marlin Na](https://github.com/Marlin-Na), 2017,
  [TnT: interactive genome browser](https://github.com/Marlin-Na/TnT).
- [Rover Van](https://github.com/RoverVan), 2017, [iregnet: regularized interval regression](https://github.com/anujkhare/iregnet).
- [Abhishek Shrivastava](https://github.com/abstatic), 2016,
  [SegAnnDB: interactive system for labeling and machine learning in genomic data](https://github.com/tdhock/SegAnnDB).
- [Faizan Khan](https://github.com/faizan-khan-iit), 2016--2017, [animint: interactive grammar of graphics](https://github.com/tdhock/animint).
- [Anuj Khare](https://github.com/anujkhare), 2016, [iregnet: regularized interval regression](https://github.com/anujkhare/iregnet).
- [Qin Wenfeng](https://github.com/qinwf), 2016, [re2r: regular expressions](https://github.com/qinwf/re2r).
- [Akash Tandon](https://github.com/analyticalmonk), 2016, [Rperform: performance testing R packages](https://github.com/analyticalmonk/Rperform).
- [Ishmael Belghazi](https://github.com/IshmaelBelghazi), 2015, [bigoptim: stochastic average gradient algorithm](https://github.com/IshmaelBelghazi/bigoptim).
- [Kevin Ferris](https://github.com/kferris10), 2015, [animint: interactive grammar of graphics](https://github.com/tdhock/animint).
- [Tony Tsai](https://github.com/caijun), 2015, [animint: interactive grammar of graphics](https://github.com/tdhock/animint).
- [Carson Sievert](https://github.com/cpsievert), 2014, [animint: interactive grammar of graphics](https://github.com/tdhock/animint).
- [Susan VanderPlas](https://github.com/srvanderplas), 2013, [animint: interactive grammar of graphics](https://github.com/tdhock/animint).
"
```

The markdown code above has a regular structure: newline, dash, space,
open square bracket, then student name. Later on there is a comma,
space, then a year for the GSOC project. We can convert the text
string above to a data table with columns for name and year, using the
regular expression R code below.

```{r}
(projects.dt <- nc::capture_all_str(
  projects.string,
  "\n- \\[",
  name="[^]]+",
  ".*?, ",
  year="[0-9]+", as.integer))
```

One way to get unique students is to use `by` inside of data table
square brackets, as below:

```{r}
projects.dt[, .(projects=.N), by=.(student=name)]
```

Another way to do that is to `dcast`, which is shown below,

```{r}
data.table::dcast(
  projects.dt,
  name ~ .,
  list(length, min, max),
  value.var="year"
)[, year_range := ifelse(
  year_min==year_max,
  year_min,
  sprintf("%d-%d", year_min, year_max)
)][order(year_range)]
```

The table above has 29 rows, one for each GSOC student I have
mentored. There are a total of 31 projects; Anirban Chetia and Yufan
Fei each did two consecutive years of GSOC.

Note that the warning above about NAs is a false positive, [that I
proposed to
remove](https://github.com/Rdatatable/data.table/issues/5512).

Exercise for the reader: add some code above to download the markdown
text string of data to parse (`projects.string`), rather than defining
it directly in R code.
