---
layout: post
title: Collaborations not allowed
description: Parsing a web page with regex
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2024-09-09-collaborations-not-allowed"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=2)
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to explore what countries and institutions are not allowed in collaborations with researchers in Canada.

## Introduction 

When applying for grants at my new job at Université de Sherbrooke in
Canada, I was asked to agree to not share research "secrets" with a
certain list of organizations, on a web page.  This is silly to ask of
most researchers, who publish in peer-reviewed journals/conferences
that are more or less publicly accessible for anyone. If anyone can
access my research, how can I prevent sharing with these
organizations? This makes me think of the McCarthy-era communist witch
hunt in the USA in the 1950s. Are we going to soon have to swear
loyalty, or renounce a certain idealogy, to continue our jobs at the
university? I find this dubious.

## Simple example

A good example of what the data look like is below:

```{r}
some.html <- '
	<li><strong>A.A. Kharkevich Institute for Information Transmission Problems, IITP, Russian Academy of Sciences</strong> (Russia)</li>
	<li><strong>Academy of Military Medical Sciences</strong> (People’s Republic of China)
	<p>Known alias(es): AMMS</p>
	</li>
	<li><strong>Academy of Military Science</strong> (People’s Republic of China)
	<p>Known alias(es): AMS</p>
	</li>
	<li><strong>Aerospace Research Institute</strong> (Iran)
	<p>Known alias(es): ARI</p>
	</li>
'
```

We parse name in each block using the regex below,

```{r}
name.pattern <- list(
  "\n\t<li><strong>",
  name=".*?",
  "</strong>")
nc::capture_all_str(
  some.html,
  name.pattern)
```

The table above has four rows and one column.
We can add another column for country via the code below.

```{r}
country.pattern <- list(
  " [(]",
  country=".*?",
  " *[)]")
nc::capture_all_str(
  some.html,
  name.pattern,
  country.pattern)
```

The table above has two columns.
To add a final column for aliases, we use the code below.

```{r}
aliases.pattern <- list(
  "\n\t<p>Known alias[(]es[)]: ",
  aliases=".*?",
  "</p>")
nca.pattern <- list(
  name.pattern,
  country.pattern,
  nc::quantifier(aliases.pattern, "?"))
nc::capture_all_str(some.html, nca.pattern)
```

## Download and parse

```{r}
orgs.html <- "../assets/2024-09-09-canada-list-of-risky-research-orgs.html"
if(!file.exists(orgs.html)){
  u <- "https://science.gc.ca/site/science/en/safeguarding-your-research/guidelines-and-tools-implement-research-security/sensitive-technology-research-and-affiliations-concern/named-research-organizations"
  download.file(u, orgs.html)
}
orgs.lines <- readLines(orgs.html)
(n.strong <- length(grep("<strong>",orgs.lines)))
orgs.dt <- nc::capture_all_str(orgs.lines, nca.pattern)
nrow(orgs.dt)
```

The number of rows above seems to agree with the number of `<strong>` tags (simpler pattern).
Below we check the number of aliases.

```{r}
sum(orgs.dt$aliases!="")
aliases.lines <- grep("alias(es)", orgs.lines, fixed=TRUE, value=TRUE)
length(aliases.lines)
```

Above it looks like there were some aliases not parsed. Which ones?

```{r}
alias.dt <- nc::capture_first_vec(paste0("\n",aliases.lines), aliases.pattern)
alias.dt[!orgs.dt,.(aliases40=substr(aliases,1,40)),on="aliases"]
```

### Trying again

Looking at the ones that did not match, it seems that there are some empty lines which are optional.

```{r}
odd.html <- '
	<li><strong>Center for High Pressure Science and Technology Advanced Research</strong> (People’s Republic of China)

	<p>Known alias(es): HPSTAR; Beijing High Voltage Science Research Center</p>
	</li>
	<li><strong>Engineering University of the CAPF</strong> (People’s Republic of China)

	<p>Known alias(es): PAP Engineering University</p>
	</li>
	<li><strong>Explosion and Impact Technology Research Centre</strong> (Iran)
	<p>Known alias(es): Research Centre for Explosion and Impact; METFAZ</p>
	</li>
	<li><strong>Institute of NBC Defense</strong> (People’s Republic of China)</li>
'
aliases.plus.pattern <- list(
  "\n+\t<p>Known alias[(]es[)]: ", #added a +
  aliases=".*?",
  "</p>")
nca.plus.pattern <- list(
  name.pattern,
  country.pattern,
  nc::quantifier(aliases.plus.pattern, "?"))
nc::capture_all_str(some.html, nca.plus.pattern)
nc::capture_all_str(odd.html, nca.plus.pattern)
```

Results above look great! Let's try it again below.


```{r}
plus.dt <- nc::capture_all_str(orgs.lines, nca.plus.pattern)
nrow(plus.dt)
n.strong
sum(plus.dt$aliases!="")
length(aliases.lines)
```

Numbers agreeing in the output above indicate that the data were parsed correctly.

## Country analysis

```{r}
plus.dt[, .(organizations=.N), by=country]
```

The output above indicates there are just three countries with organizations to avoid.

## Aliases

How many aliases per organization?

```{r}
(a.dt <- plus.dt[
, alias.list := strsplit(aliases,split="; ")
][
, .(alias=alias.list[[1]]), by=name
])
tibble::tibble(plus.dt[
, n.alias := sapply(alias.list, length)
]) # for nice print.
## another way to get the count is via .N in join:
a.dt[plus.dt, .(.N=.N, n.alias), on='name', by=.EACHI]
```

## Bonus: French

Below we combine the different parts of the regex code above into a
single regex code block below, with some additions specific to the
French version of the web page.

```{r}
## https://science.gc.ca/site/science/fr/protegez-votre-recherche/lignes-directrices-outils-pour-mise-oeuvre-securite-recherche/recherche-technologies-sensibles-affiliations-preoccupantes/organisations-recherche-nommees
french.html <- "
	<li><strong>Académie des transports militaires de l'armée <em>[Army Military Transportation Academy]</em></strong> (République populaire de Chine)
	<p>Alias connu(s) : <em>Academy of Military Transportation</em></p>
	</li>
	<li><strong>Académie d'infanterie de l'armée <em>[Army Infantry Academy]</em></strong> (République populaire de Chine)
	<p>Alias connu(s) : <em>Army Infantry Academy of PLA; PLA Army Infantry Academy; Nachang Army Academy</em></p>
	</li>
	<li><strong>Académie du service militaire <em>[Army Service Academy]</em></strong> (République populaire de Chine)
	<p>Alias connu(s) : <em>PLA Army Service Academy</em></p>
	</li>
	<li><strong>Académie militaire d'artillerie et de la défense aérienne <em>[Army Academy of Artillery and Air Defense]</em></strong> (République populaire de Chine)</li>
	<li><strong>Académie militaire de défense des frontières et des côtes <em>[Army Academy of Border and Coastal Defense]</em></strong> (République populaire de Chine)</li>
	<li><strong>Académie militaire des forces blindées <em>[Army Academy of Armored Forces]</em></strong> (République populaire de Chine)
	<p>Alias connu(s) : <em>Army Armored Forces Academy; Armored Forces Engineering Academy Académie navale de Dalian (Dalian)</em></p>
	</li>
	<li><strong>Académie navale de Dalian <em>[Dalian Naval Academy]</em></strong> (République populaire de Chine)
	<p>Alias connu(s) : <em>PLA Dalian Naval Academy</em></p>
	</li>
	<li><strong>Association professionnelle de l'Organisation autonome à but non lucratif des concepteurs de systèmes informatiques <em>[Autonomous Noncommercial Organization Professional Association of Designers of Data Processing Systems]</em></strong> (Russie)
	<p>Alias connu(s) : <em>ANO PO KSI</em></p>
	</li>
"
french.dt <- nc::capture_all_str(
  french.html,
  "\n\t<li><strong>",
  french=".*?",
  " <em>\\[",
  english=".*?",
  "\\]</em></strong> [(]",
  pays=".*?",
  " *[)]",
  nc::quantifier(
    "\n+\t<p>Alias connu[(]s[)] : <em>", 
    aliases=".*?",
    "</em>",
    "?"
  )
)
abbrev <- function(x,m)paste0(substr(x,1,m),"…")
(french.aliases <- french.dt[, let(
  alias.list = strsplit(aliases,split="; "),
  français = abbrev(french,25),
  English = abbrev(english,20)
)][
, .(alias=alias.list[[1]]), by=français
])
```

The result table above has one row for each alias.
The code below produces an table with one row per organisation.

```{r}
french.aliases[french.dt, .(
  English,
  pays,  
  Alias=if(.N)paste(abbrev(alias,10), collapse="; ") else ""
), by=.EACHI, on="français"]
```

Exercise for the reader: download the French web page, and convert it to a table using this new regex code. Do you get the same number of organisations as with the English version?

## Conclusions

We used regular expressions to help us understand that Canada does not
want researchers collaborating with certain organizations in three
countries: Russia, Iran, China.

## Session info

```{r}
sessionInfo()
```
