---
layout: post
title: Overhead of auto-grad in torch
description: Comparison with explicit gradients
---

```{r Ropts, echo=FALSE, results='hide'}
repo.dir <- normalizePath("..")
post.id <- "2024-10-30-mlr3torch"
fig.path <- paste0(file.path(repo.dir, "assets", "img", post.id), "/")
dir.create(fig.path, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=10, ## TODO python figures wider? look at prev issue.
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
conda.env <- "2023-08-deep-learning"
conda.env <- "torch-aum"
RETICULATE_PYTHON <- sprintf(if(.Platform$OS.type=="unix")
  ##"/home/tdhock/.local/share/r-miniconda/envs/%s/bin/python"
  "/home/tdhock/miniconda3/envs/%s/bin/python"
  else "~/AppData/Local/Miniconda3/envs/%s/python.exe", conda.env)
Sys.setenv(RETICULATE_PYTHON=RETICULATE_PYTHON)
##reticulate::use_condaenv(dirname(RETICULATE_PYTHON), required=TRUE)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to show how to use the mlr3torch package in R
with AUM (Area Under Min of False Positive and False Negative rates,
our newly Proposed surrogate loss for ROC curve optimization).

## Introduction 

Last Friday I gave a talk at [MILA](https://mila.quebec/en), [Two new
algorithms for scientific applications of machine
learning](https://github.com/tdhock/two-new-algos-sci-ml/tree/main?tab=readme-ov-file#title-abstract-slides).
The first algorithm that I discussed is [SOAK: Same/Other/All K-fold
cross-validation for estimating similarity of patterns in data
subsets](https://arxiv.org/abs/2410.08643). One simple
demonstration of the algorithm involves three image classification benchmark data sets. Code below adapted from [my github repo](https://github.com/tdhock/cv-same-other-paper/blob/main/data_Classif_MNIST_other.R):

```{r}
other.name.vec <- c("EMNIST", "FashionMNIST")
data.name.vec <- c(other.name.vec, "MNIST")
```

First we download those data sets.

```{r}
prefix <- "https://rcdata.nau.edu/genomic-ml/cv-same-other-paper/data_Classif/"
data_Classif <- "~/projects/cv-same-other-paper/data_Classif"
options(timeout = 600)#seconds
for(data.name in data.name.vec){
  data.csv <- paste0(data.name, ".csv")
  local.csv <- file.path(data_Classif, data.csv)
  if(!file.exists(local.csv)){
    remote.csv <- paste0(prefix, data.csv)
    download.file(remote.csv, local.csv)
  }
}
```

Figure 5 from that paper shows that

https://rcdata.nau.edu/genomic-ml/cv-same-other-paper/data_Classif/

https://cran.r-project.org/web/packages/mlr3torch/readme/README.html

https://mlr3torch.mlr-org.com/articles/pipeop_torch.html

## Conclusions

TODO

## Session info

```{r}
sessionInfo()
```
