s
---
layout: post
title: Writing comprehensible tests
description: Documenting key code magic numbers in animint2 tests
---

[animint2](https://github.com/tdhock/animint2) is an R package
for animated, interactive data visualization.
It uses R ggplot code, which makes it possible to define a wide range of web-based data visualizations, as can be seen in the [gallery](https://animint.github.io/gallery/).
The goal of this post is to explain some changes in the framework used for testing `animint2`.

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2024-08-14-chromote-key-codes"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

## Background about animint2

Since `animint2` creates web-based data visualizations, we can test the software by using a web browser to render, and examining the output 
TODO headless browser, RSelenium, chromote.

### Old test

```r
test_that("top widget adds/remove points", {
  expect_equal(get_circles(), list(10, 10))
  remDr$sendKeysToActiveElement(list(key="backspace"))
  sendBackspace()
  expect_equal(get_circles(), list(5, 10))
  remDr$sendKeysToActiveElement(list(key="backspace"))
  sendBackspace()
  expect_equal(get_circles(), list(0, 10))
  remDr$sendKeysToActiveElement(list("a", key="enter"))
  send("a")
tdhock marked this conversation as resolved.
  expect_equal(get_circles(), list(5, 10))
  remDr$sendKeysToActiveElement(list("b", key="enter"))
  send("b")
  expect_equal(get_circles(), list(10, 10))
})
```

### Siddesh's first proposition

From https://github.com/animint/animint2/blob/d2b5de7cd69bd22e50bbc6da88b528f89828c598/tests/testthat/test-renderer1-knit-print.R

```r
sendKey <- function(key, code, keyCode) {
  remDr$Input$dispatchKeyEvent(type = "keyDown", key = key, code = code, windowsVirtualKeyCode = keyCode, nativeVirtualKeyCode = keyCode)
  remDr$Input$dispatchKeyEvent(type = "keyUp", key = key, code = code, windowsVirtualKeyCode = keyCode, nativeVirtualKeyCode = keyCode)
}
sendBackspace <- function() {
  if (remDr$browserName == "chromote") {
    sendKey("Backspace", "Backspace", 8)
  } else {
    remDr$sendKeysToActiveElement(list(key="backspace"))
  }
  Sys.sleep(0.5)
}
sendA <- function() {
  if (remDr$browserName == "chromote") {
    remDr$Input$insertText(text = "a")
    sendKey("Enter", "Enter", 13)
  } else {
    remDr$sendKeysToActiveElement(list("a", key="enter"))
  }
  Sys.sleep(0.5)
}
sendB <- function() {
  if (remDr$browserName == "chromote") {
    remDr$Input$insertText(text = "b")
    sendKey("Enter", "Enter", 13)
  } else {
    remDr$sendKeysToActiveElement(list("b", key="enter"))
  }
  Sys.sleep(0.5)
}```

### Siddesh's proposed fix

From https://github.com/animint/animint2/blob/2a633480122337d98888b8686a6f81b699828ad9/tests/testthat/helper-HTML.R#L83

```r
sendKey <- function(key) {
  stopifnot(is.character(key))
  #The key codes in the list below are adopted from Windows Virtual keycode standards
  #https://learn.microsoft.com/en-us/windows/win32/inputdev/virtual-key-codes
  # VK_BACK->Backspace, VK_RETURN->Enter, VK_DOWN->ArrowDown
  # we use the corresponding decimal values of the key codes given in hex value in the above link
  key2code <- c(
    Backspace=8,
    Enter=13,
    ArrowDown=40)
  for (type in c("keyDown", "keyUp")) {
    remDr$Input$dispatchKeyEvent(
	  type = type, key = key, code = key, 
	  windowsVirtualKeyCode = key2code[[key]], 
	  nativeVirtualKeyCode = key2code[[key]])
  }
}
```

### My proposed new test

From https://github.com/animint/animint2/blob/053dde95b5be7f2fced0741cc71a1ddb8b900d19/tests/testthat/helper-HTML.R

```r
sendBackspace <- function() {
  sendKey("Backspace")
  Sys.sleep(0.5)
}
send <- function(alphabet) {
  remDr$Input$insertText(text = alphabet)
  sendKey("Enter")
  Sys.sleep(0.5)
}
test_that("top widget adds/remove points", {
  expect_equal(get_circles(), list(10, 10))
  sendBackspace()
  expect_equal(get_circles(), list(5, 10))
  sendBackspace()
  expect_equal(get_circles(), list(0, 10))
  send("a")
  expect_equal(get_circles(), list(5, 10))
  send("b")
  expect_equal(get_circles(), list(10, 10))
})
```

```r 
### The hex codes come from
### https://developer.mozilla.org/en-US/docs/Web/API/UI_Events/Keyboard_event_key_values
key2hex_code <- c(
  Backspace="08",
  Enter="0D",
  ArrowDown="28")
### https://chromedevtools.github.io/devtools-protocol/tot/Input/ says
### that dispatchKeyEvent() requires DOM key codes (in decimal) for
### the windowsVirtualKeyCode and nativeVirtualKeyCode arguments.
key2dec_code <- structure(
  strtoi(key2hex_code,base=16),
  names=names(key2hex_code))
# Function to send a key event
sendKey <- function(key) {
  stopifnot(is.character(key))
  for (type in c("keyDown", "keyUp")) {
    remDr$Input$dispatchKeyEvent(type = type, key = key, code = key, windowsVirtualKeyCode = key2dec_code[[key]], nativeVirtualKeyCode = key2dec_code[[key]])
  }
}
```

## What about the other hex codes?

```{r}
library(data.table)
code <- function(dom_name, key_name, hex){
  data.table(dom_name, key_name, hex)
}
(animint2.expected <- rbind(
  code("Backspace", "VK_BACK",   "08"),
  code("Enter",     "VK_RETURN", "0D"),
  code("ArrowDown", "VK_DOWN",   "28")))
local.html <- "~/keyboard_event_key_values.html"
if(!file.exists(local.html)){
  remote.url <- "https://developer.mozilla.org/en-US/docs/Web/API/UI_Events/Keyboard_event_key_values"
  download.file(remote.url, local.html)
}
html.lines <- readLines(local.html)
```

Contains entries like:

```{r}
some.html <- '
	<tr>
      <td><code>"Backspace"</code></td>
      <td>
        The <kbd>Backspace</kbd> key. This key is labeled <kbd>Delete</kbd> on
        Mac keyboards.
      </td>
      <td><code>VK_BACK</code> (0x08)</td>
      <td><code>kVK_Delete</code> (0x33)</td>
      <td>
        <code>GDK_KEY_BackSpace</code> (0xFF08)<br><code>Qt::Key_Backspace</code>
        (0x01000003)
      </td>
      <td><code>KEYCODE_DEL</code> (67)</td>
    </tr>
    <tr>
      <td><code>"Enter"</code></td>
      <td>
        The <kbd>Enter</kbd> or <kbd>↵</kbd> key (sometimes labeled
        <kbd>Return</kbd>).
      </td>
      <td><code>VK_RETURN</code> (0x0D)</td>
      <td>
        <code>kVK_Return</code> (0x24)<br><code>kVK_ANSI_KeypadEnter</code>
        (0x4C)<br><code>kVK_Powerbook_KeypadEnter</code> (0x34)
      </td>
      <td>
        <code>GDK_KEY_Return</code> (0xFF0D)<br><code>GDK_KEY_KP_Enter</code>
        (0xFF8D)<br><code>GDK_KEY_ISO_Enter</code> (0xFE34)<br><code>GDK_KEY_3270_Enter</code>
        (0xFD1E)<br><code>Qt::Key_Return</code> (0x01000004)<br><code>Qt::Key_Enter</code>
        (0x01000005)
      </td>
      <td>
        <code>KEYCODE_ENTER</code> (66)<br><code>KEYCODE_NUMPAD_ENTER</code>
        (160)<br><code>KEYCODE_DPAD_CENTER</code> (23)
      </td>
    </tr>
	<tr>
      <td><code>"ArrowDown"</code> [1]</td>
      <td>The down arrow key.</td>
      <td><code>VK_DOWN</code> (0x28)</td>
      <td><code>kVK_DownArrow</code> (0x7D)</td>
      <td>
        <code>GDK_KEY_Down</code> (0xFF54)<br><code>GDK_KEY_KP_Down</code>
        (0xFF99)<br><code>Qt::Key_Down</code> (0x01000015)
      </td>
      <td><code>KEYCODE_DPAD_DOWN</code> (20)</td>
    </tr>
'
```

```{r}
dom_name_pattern <- list(
  '<td><code>"',
  dom_name='[^"]+')
nc::capture_all_str(some.html, dom_name_pattern)
nc::capture_all_str(html.lines, dom_name_pattern)
```

The output above suggests there are 319 lines on the web page which have a DOM code.

```{r}
td_key_name_pattern <- list(
  '      <td><code>',
  key_name="VK_.*?",
  '</code>')
nc::capture_all_str(some.html, td_key_name_pattern)
nc::capture_all_str(html.lines, td_key_name_pattern)
```

The output above suggests there are 318 lines with windows key codes.

```{r}
key_name_pattern <- list(
  '<code>',
  key_name="VK_.*?",
  '</code>')
nc::capture_all_str(some.html, key_name_pattern)
nc::capture_all_str(html.lines, key_name_pattern)
```

The output above is a table with 370 rows, which suggests there may be
more than one `key_name` per `dom_name`.

To find out, we can first try matching all HTML `<tr>` rows:

```{r}
tr_pattern <- list(
  '    <tr>\n      <td><code>',
  in.tr="(?:.*\n)+?",
  '    </tr>')
tr_all <- nc::capture_all_str(
  html.lines, tr_pattern
)[
, first70 := substr(in.tr,1,70)
][]
tr_all[, .(first70)]
```

The output above suggests there are 348 rows which could have a DOM name,
and could contain key name codes. Below we try extracting a DOM name from
each:

```{r}
dom_all <- nc::capture_first_vec(
  tr_all$in.tr,
  '^"',
  dom_name='[^"]+',
  rest="(?:.*\n)*",
  nomatch.error=FALSE)
has.dom <- !is.na(dom_all$dom_name)
tr_all[has.dom, .(first70)]
```

The table above represents the 319 HTML `<tr>` elements which did
contain a DOM which matched our regex. The table below shows the 29
elements which did not match, which seem to be [dead
keys](https://developer.mozilla.org/en-US/docs/Web/API/UI_Events/Keyboard_event_key_values#dead_keycodes_for_linux
) (probably not interesting for testing).

```{r}
tr_all[!has.dom, .(first70)]
```

Below we compute all key codes listed in each HTML row.

```{r}
(key_all <- dom_all[, nc::capture_all_str(
  rest,
  "<code>",
  key_name=".*?",
  "</code> [(]0x",
  hex="[^)]+"
), by=dom_name])
```

The table above has one row for each virtual `key_name` found. Some
`dom_name` values are missing in the table above, because there were
no corresponding `key_name` values found, and below we show two which
do have hex numbers somewhere in the row:

```{r}
dom_all[!key_all, on="dom_name"][grepl("0x", rest)]
```

Above we see several hex codes, in two sound-related keys, which we
can ignore for testing.
Below we subset by the windows keys with a `VK` prefix:

```{r}
(win_all <- key_all[grepl("^VK",key_name)])
```

Below we take the first hex code mentioned:

```{r}
(win_first <- win_all[, .SD[1], by=dom_name])
```

The table above could be used to define a new `key2hex_code` variable,
which would expand the keys supported in `animint2` tests:

```{r}
key2hex_code <- win_first[, structure(hex, names=dom_name)]
dput(key2hex_code)
```

To verify that it is compatible with the ones
already used in the tests, we can further subset to get the keys we
use in existing `animint2` tests:

```{r}
(win_some <- win_first[c("Backspace","Enter","ArrowDown"),on="dom_name"])
identical(win_some, animint2.expected)
```

Above we see that the data computed using regex are consistent with
the data which are already being used in the `animint2` tests.

## Session info

```{r}
sessionInfo()
```
