---
layout: post
title: Nest and un-nest operations
description: Making and expanding list columns
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2024-09-01-nest-unnest-hoist"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=2)
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to explain how `data.table` can be used to
reshape data with list columns.

## Introduction to list columns

In R the main data structure is `data.frame`, which is a 2d object,
like a matrix, but the columns may have different types. Each column
is often an atomic type, which means one of integer, double/numeric,
character, or logical. For example, if you have an integer column,
then each entry in that column is an integer (C int).

```{r}
library(data.table)
(two.rows <- data.table(
  INT=c(1L,NA_integer_),
  NUM=c(Inf,NA_real_),
  CHR=c("foo",NA_character_),
  LOG=c(TRUE,NA)))
```

The above is a data table with two rows.
List columns are useful when you have data which is more complex than a simple atomic type.
To add a list column to the table above (which has two rows), 
we need to make sure it has two entries.

```{r}
two.rows[, `:=`(
  L_INT_LOG=list(1:5, NA),
  L_NUM_LIST=list(c(3.14,NaN,-Inf), list("foo", function(x)x^2, quote(bar)))
)][]
```

The table above has two additional list columns. Each list has two
elements (because the table has two rows). Each element can be an R
object of any type, which we can see using the code below:

```{r}
str(two.rows)
```

The output above shows that
* `L_INT_LOG` is a list of two items, of types integer and logical.
* `L_NUM_CHR` is a list of two items, of types numeric and list. This
  internal/nested list has 3 items, of types character, function, and symbol.
  
## animint JSON example

List columns can be useful when parsing JSON data.

For example, consider data from the animint gallery.

```{r}
meta.csv <- "~/R/gallery/meta.csv"
if(!file.exists(meta.csv)){
  download.file("https://github.com/animint/gallery/raw/gh-pages/meta.csv", meta.csv)
}
meta.dt <- fread(meta.csv)
str(meta.dt)
```

The `meta.dt` above has one row per visualization published in the
[animint gallery](https://animint.github.io/gallery/). Each has a
corresponding JSON plot data file, which we can download using the
code below:

```{r}
meta.dt[, json.list := {
  plot.json <- file.path("~/viz-json", viz_owner_repo, "plot.json")
  if(!file.exists(plot.json)){
    u <- sprintf("https://github.com/%s/raw/gh-pages/plot.json", viz_owner_repo)
    dir.create(dirname(plot.json), showWarnings=FALSE, recursive = TRUE)
    download.file(u, plot.json)
  }
  list(list(RJSONIO::fromJSON(plot.json)))
}, by=.I]
```

The code below examines what names are present in each downloaded JSON file:

```{r}
lapply(meta.dt$json.list, names)
```

### data.table analogs of tidyr functions

`tidyr` provides several functions related to nesting. 
Below we use the wider function, which creates a table with one row for each list element, and one column for each list name.

```{r}
only.json <- meta.dt[,.(json.list)]
tidyr::unnest_wider(only.json, json.list)
```

We see in the table above that there are `<NULL>` in some rows of
`durations` and `time` columns, which is consistent with the `lapply`
output we saw earlier.

The analog in `data.table` would be to use 
* `by=.I` to run some code for each row,
* `json.list[[1]]` to get the inner/nested list out of the list column,
* and `as.data.table` to create a data table with a list column for each inner list element.

```{r}
only.json[, as.data.table(json.list[[1]]), by=.I]
```

The output above shows an error; `data.table` does not want to do this operation for you, because the number of items/columns in each group do not match. 
To get around that, you can explicitly specify what items/columsn you want:

```{r}
(after.wider <- only.json[, with(json.list[[1]], .(
  geoms=list(geoms), title
)), by=.I])
```

So as we see above, the unnest wider operation can be achieved using
plain `data.table` code, although it is more explicit, and less
convenient, than the corresponding `tidyr` code.

Next we do an unnest which expands the list of geoms into one row per geom.

```{r}
only.geoms <- after.wider[,.(title,geoms)]
tidyr::unnest_longer(only.geoms, geoms, "geom_info")
```

The equivalent in `data.table` is similar to the wider approach above:

```{r}
(after.longer <- only.geoms[, .(
  geom_info=geoms[[1]], geom_info_id=names(geoms[[1]])
), by=title])
```

We can again go wider to get info about each geom, such as counts of which geoms are most frequently used below.

```{r}
tidyr::unnest_wider(after.longer, geom_info) 
second.wider <- after.longer[, .(short_title=substr(title,1,20), geom=sapply(geom_info, "[[", "geom"))]
dcast(second.wider, geom ~ ., length)
dcast(second.wider, short_title ~ geom, length)
```

Another way to get the same info is via TODO

```{r}
hoist.result <- tidyr::hoist(meta.dt, json.list, "geoms", "plots", "selectors")
```

```{r}
meta.dt[, geoms := list(json.list[[1]]$geoms), by=.I]
```

```{r}
spec <- unnest::s(
  stack = TRUE,
  unnest::s("geoms"),
  unnest::s("plots"),
      "owner",
      unnest::s("login"))))
unnest.result <- unnest::unnest(gh_repos, spec)
```

```{r}
Rdatatable.members <- gh::gh("/orgs/Rdatatable/members")
Rdatatable.roles <- gh::gh("/orgs/Rdatatable/organization-roles")
str(Rdatatable.members[1:2])
library(data.table)
do.call(rbind, lapply(Rdatatable.members, as.data.table))
dt.list.col <- data.table(L=Rdatatable.members)
tidyr::unnest_wider(dt.list.col, L)
dt.list.col[, L[[1]], by=.I]
hoist.result <- tidyr::hoist(dt.list.col, L, "login", "type")
dt.list.col

data("gh_repos",package="repurrrsive")
length(gh_repos)
repos_tib <- tibble::tibble(repo = gh_repos)
repos_tib_longer <- tidyr::unnest_longer(repos_tib, repo)
hoist.result <- tidyr::hoist(
  repos_tib_longer, repo,
  login = c("owner", "login"),
  name = "name",
  homepage = "homepage",
  watchers = "watchers_count") |>
  dplyr::select(-repo)
spec <- unnest::s(
  stack = TRUE,
  unnest::s(
    stack = TRUE,
    unnest::s("name"),
    unnest::s("homepage"),
    unnest::s("watchers_count", as = "watchers"),
    unnest::s(
      "owner",
      unnest::s("login"))))
unnest.result <- unnest::unnest(gh_repos, spec)

get_repo_row <- function(repo){
  one.row <- with(repo, data.table(name, watchers=watchers_count))
  one.row$login <- repo$owner$login
  one.row$homepage <- if(is.null(repo$homepage))NA_character_ else repo$homepage
  one.row
}
data.table(L=gh_repos)[, {
  data.table(repos.list=L[[1]])[, {
    get_repo_row(repos.list[[1]])
  }, by=.(repo=.I)] ## we still get column named I?
}, by=.(user=.I)]

repo_dt <- data.table(L=gh_repos)[, {
  data.table(Rlist=L[[1]])
}, by=.I]
repo_dt[, {
  get_repo_row(Rlist[[1]])
}, by=.I]
repo_dt

## TODO run on faster computer.
atime.res <- atime::atime(
  setup={
    N_repos <- rep(gh_repos, l=N)
  },
  dt_nested={
    data.table(L=N_repos)[, {
      data.table(repos.list=L[[1]])[, {
        get_repo_row(repos.list[[1]])
      }, by=.(repo=.I)] ## we still get column named I?
    }, by=.(user=.I)]
  }, 
  dt_sequential={
    repo_dt <- data.table(L=N_repos)[, {
      data.table(Rlist=L[[1]])
    }, by=.I]
    repo_dt[, {
      get_repo_row(Rlist[[1]])
    }, by=.I]
  },
  hoist={
    repos_tib <- tibble::tibble(repo = N_repos)
    repos_tib_longer <- tidyr::unnest_longer(repos_tib, repo)
    tidyr::hoist(
      repos_tib_longer, repo,
      login = c("owner", "login"),
      name = "name",
      homepage = "homepage",
      watchers = "watchers_count") |>
      dplyr::select(-repo)
  }, 
  unnest={
    spec <- unnest::s(
      stack = TRUE,
      unnest::s(
        stack = TRUE,
        unnest::s("name"),
        unnest::s("homepage"),
        unnest::s("watchers_count", as = "watchers"),
        unnest::s(
          "owner",
          unnest::s("login"))))
    unnest::unnest(N_repos, spec)
  }
)
plot(atime.res)
```

## Conclusions

TODO

## Session info

```{r}
sessionInfo()
```
