---
title: "Comparing Pandas in Python to data.table in R"
output: html_document
date: "2024-02-22"
---


```{r setup, include = FALSE}
library(data.table)
library(ggplot2)
conda.env <- "torch-aum"
RETICULATE_PYTHON <- normalizePath(sprintf(if(.Platform$OS.type=="unix")
  ##"/home/tdhock/.local/share/r-miniconda/envs/%s/bin/python"
  "/home/tdhock/miniconda3/envs/%s/bin/python"
  else "~/AppData/Local/miniconda3/envs/%s/python.exe", conda.env))
##RETICULATE_PYTHON <- "~/AppData/Local/miniconda3/python.exe"
##Sys.setenv(RETICULATE_PYTHON=RETICULATE_PYTHON)
reticulate::use_condaenv("~/AppData/Local/miniconda3/envs/torch-aum")
reticulate::py_run_string("input_df_pd.to_csv('data.csv', index=False)")
```

```{r}
write.colors <- c(
  "data.table::fwrite" = "#D6604D",
  "pandas::to_csv" = "#BF812D"
)
n.rows <- 100
atime.write.vary.cols <- atime::atime(
  N = as.integer(10^seq(2, 6, by = 0.5)),
  setup = {
    set.seed(1)
    input.vec <- rnorm(n.rows * N)
    input.mat <- matrix(input.vec, n.rows, N)
    input.df <- data.frame(input.mat)
    ##pd <- reticulate::import("pandas")
    ##input_df_pd <- pd$DataFrame(input.df)
  },
  seconds.limit = 0.01,
  "data.table::fwrite" = {
    data.table::fwrite(input.df, tempfile(), showProgress = FALSE)
  },
  "pandas::to_csv" = {
    reticulate::py_run_string("input_df_pd.to_csv('data.csv', index=False)")
  }
)

```





```{r}
refs.write.vary.cols <- atime::references_best(atime.write.vary.cols)
pred.write.vary.cols <- predict(refs.write.vary.cols)

gg.write.dt.pd <- plot(pred.write.vary.cols)+
  theme(text=element_text(size=20))+
  ggtitle(sprintf("Write real numbers to CSV, with pandas in Python and data.table in R, %d x N", n.rows))+
  scale_x_log10("N = number of columns to write")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=write.colors)+
  scale_color_manual(values=write.colors)
```




```{r}
png()
print(gg.write.dt.pd)
ggsave("gg.write.dt.pd.png", gg.write.dt.pd, width = 10, height = 6, units = "in", dpi = 300)
dev.off()
```


# read a Csv files with Pandas and data.table
```{python}
file_path = 'data.csv',
```

```{r}
read.colors <- c(
  "data.table::fread" = "#D6604D",
  "pandas::read_csv" = "#BF812D"
)
n.rows <- 100
seconds.limit <- 1


atime.read <- atime::atime(
  setup = {
    set.seed(1)
    input.vec <- rnorm(n.rows*N)
    input.mat <- matrix(input.vec, n.rows, N)
    input.df <- data.frame(input.mat)
    input.csv <- tempfile()
    fwrite(input.df, "data.csv")
    
    pd <- import("pandas")
    input_df_pd <- pd$DataFrame(input.df)
    #input_df_pd.to_csv('input.csv', index=False)
    
    
  },
  seconds.limit = seconds.limit,
  "data.table::fread" = {
    data.table::fread("data.csv", showProgress = FALSE)
    
    
  },
  "pandas::read_csv" = {
    pd <- import("pandas")
    reticulate::py_run_string("import pandas as pd")
    reticulate::py_run_string("pd.read_csv(file_path)")
    #py$input_df_pd
  }
)
```

import pandas as pd

# Read the CSV file
df = pd.read_csv('path/to/file.csv')

# Display the contents of the DataFrame
print(df)

```{r}
refs.read.vary.cols <- atime::references_best(atime.read)
pred.read.vary.cols <- predict(refs.read.vary.cols)

gg.read.pd <- plot(pred.read.vary.cols)+
  theme(text=element_text(size=20))+
  ggtitle(sprintf("Read real numbers to CSV, with pandas in Python and data.table in R, %d x N", n.rows))+
  scale_x_log10("N = number of columns to write")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=read.colors)+
  scale_color_manual(values=read.colors)
```

```{r}
png()
print(gg.read.pd)
ggsave("gg.read.pd.png", gg.read.pd, width = 10, height = 6, units = "in", dpi = 300)
dev.off()

gg.read.pd

```


# Reshaping data

Wide to long

```{r}
ml.colors <- c(
  "data.table::melt"="#D6604D",
  "pandas::pd.melt" = "#BF812D"
  )
n.folds <- 10
n.rows <- 100
seconds.limit <- 1

ml.reshape.atime <- atime::atime(
  N=as.integer(10^seq(2, 15, by=0.5)),
  setup={
    df <- data.frame(
      id = rep(1:N, each = 2),
      category = rep(c("A", "B"), N),
      value = rnorm(2 * N)
      )
    },
  seconds.limit= seconds.limit,
  
  "data.table::melt" = {
    data.table::melt(data.table(df), id.vars = c("id",  "category"),variable.names="variable", value.name = "value")
  },
  "pandas::pd.melt" = {
    py_df <- reticulate::r_to_py(df)
    pd <- import("pandas")
    pd$melt(py_df, id_vars = c("id", "category"), value_name = "score")
    #reticulate::py_run_string("import pandas as pd")
    #reticulate::py_run_string(pd$melt(py_df, id_vars = c('id', 'category'), value_name = 'score'))
    #melted_df <- py_to_r(melt_df)
    
  #reticulate::py_run_string("import pandas as pd")
  #reticulate::py_run_string(pd$melt(df_pd, id_vars = c("id", "category"), var_name = "variable", value_name = "melted_value"))
  }

  )

```

```{r}

ml.reshape.refs <- atime::references_best(ml.reshape.atime)
ml.reshape.pred <- predict(ml.reshape.refs)
ml.wide2long.pd <- plot(ml.reshape.pred)+
  theme(text=element_text(size=15))+
  ggtitle(sprintf("Reshaping from wide to long panda & data.tableover %dreal numbers, N times", n.folds))+
  scale_x_log10("N = number of Mean,SD,Length to compute")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=ml.colors)+
  scale_color_manual(values=ml.colors)
```

```{r}
png()
print(ml.wide2long.pd)
ggsave("ml.wide2long.pd.png", ml.wide2long.pd, width = 10, height = 6, units = "in", dpi = 300)
dev.off()

ml.wide2long.pd
```





```{r}
ml.wide2long.atime <- atime::atime(
  N=as.integer(10^seq(2, 15, by=0.5)),
  setup={
    df <- data.frame(
      id = rep(1:N, each = 2),
      category = rep(c("A", "B"), N),
      value = rnorm(2 * N)
    )
  },
  seconds.limit= seconds.limit,
  "data.table::melt" = {
    data.table::melt(data.table(df), id.vars = c("id",  "category"),variable.names="variable", value.name = "value")
  },
  "pandas::pd.melt" = {
    pd <- import("pandas")
    reticulate::py_run_string("import pandas as pd")
    df_pd <- r_to_py(df)
    reticulate::py_run_string(pd$melt(df_pd, id_vars = c("id", "category"), var_name = "variable", value_name = "melted_value"))
  }
)
```






```{r}

ml.colors <- c(
  "data.table::melt" = "#D6604D",
  "pandas::pd.melt" = "#BF812D"
)

ml.wide2long.atime <- atime::atime(
  N=as.integer(10^seq(2, 15, by=0.5)),
  setup={
    df <- data.frame(
      id = rep(1:N, each = 2),
      category = rep(c("A", "B"), N),
      value = rnorm(2 * N)
    )
  },
  seconds.limit= seconds.limit,
  "data.table::melt" = {
    data.table::melt(data.table(df), id.vars = c("id",  "category"),variable.names="variable", value.name = "value")
  },
  "pandas::pd.melt" = {
    pd <- import("pandas")
    reticulate::py_run_string("import pandas as pd")
    df_pd <- r_to_py(df)
    reticulate::py_run_string(pd$melt(df_pd, id_vars = c("id", "category"), var_name = "variable", value_name = "melted_value"))
                              
    #reticulate::py_run_string(pd$melt(df_pd, id_vars = c("id", "category"), var_name = "variable", value_name = "value"))
    #reticulate::py_run_string(pd$pd.melt(df_pd, id_vars = c("id", "category"), var_name = "variable", value_name = "value"))
  }
)
```





```{r,warning=FALSE,message=FALSE}

ml.colors <- c(
  "data.table::melt" = "#D6604D",
  "pandas::pivot_table" = "#BF812D"
)

n.folds <- 10
n.rows <- 100
seconds.limit <- 10

ml.long2wide.atime <- atime::atime(
  N=as.integer(10^seq(2, 7, by=0.2)),
  
  setup={
    df <- data.frame(
      id = rep(1:N, each = 2),
      category = rep(c("A", "B"), N),
      value = rnorm(2 * N)
      )
    },
  seconds.limit= seconds.limit,
  "data.table::dcast" = {
    data.table::dcast(data.table(df), id ~ category, value.var = "value")
  },
  "pandas::pivot_table" = {
    py_df <- reticulate::r_to_py(df)
    pd <- import("pandas")
    pd$pivot_table(py_data, values = "value", index = "id", columns = "category")
  }
  )

```



```{r,warning=FALSE,message=FALSE}
ml.long2wide.refs <- atime::references_best(ml.long2wide.atime)
ml.long2wide.pred <- predict(ml.long2wide.refs)
ml.long2wide <- plot(ml.long2wide.pred)+
  theme(text=element_text(size=20))+
  ggtitle(sprintf("Reshaping from long to wide over %dreal numbers, N times", n.folds))+
  scale_x_log10("N = number of Mean,SD,Length to compute")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=ml.colors)+
  scale_color_manual(values=ml.colors)

```

```{r}
#print(ml.long2wide)
ggsave("ml.long2wide.png", ml.long2wide, width = 10, height = 6, units = "in", dpi = 300)
#dev.off()

#ml.long2wide
```

























pd <- import("pandas")
    reticulate::py_run_string("import pandas as pd")
    reticulate::py_run_string("pd.read_csv(file_path)")

library(reticulate)

ml.wide2long.atime <- atime::atime(
  N = as.integer(10^seq(2, 15, by = 0.5)),
  setup = {
    df <- data.frame(
      id = rep(1:N, each = 2),
      category = rep(c("A", "B"), N),
      value = rnorm(2 * N)
    )
    df
  },
  seconds.limit = seconds.limit,
  "data.table::melt" = {
    data.table::melt(data.table(df), id.vars = c("id", "category"), variable.name = "variable", value.name = "value")
  },
  "pandas::pd.melt" = {
    pd <- import("pandas")
    df_pd <- r_to_py(df)
    pd$pd.melt(df_pd, id_vars = c("id", "category"), var_name = "variable", value_name = "value")
  }
)

# Summarizing by group

```{r}
ml.colors <- c(
  "dplyr::summarise"="#00FF00",
  "[.data.table"="#D6604D",
  "stats::aggregate"="#4DB3E6",
  "summarytools::descr"="#00BFC4",
  "psych::describe"="#FFA500",
  "plyr::ddply"="#FF0000FF")
  
options(dplyr.summarise.inform=FALSE)
n.folds <- 10
ml.atime <- atime::atime(
  N=as.integer(10^seq(2, 7, by=0.5)),
  setup={
    loss.dt <- data.table(
      name="loss", 
      fold=rep(1:n.folds, each=2*N),
      loss=rnorm(2*N*n.folds),
      set=rep(c("subtrain","validation"),each=N),
      epoch=1:N,
      key=c("set","epoch","fold"))
  },
  seconds.limit=seconds.limit,
  "[.data.table"={
    loss.dt[, .(
      loss_length=.N,
      loss_mean=mean(loss),
      loss_sd=sd(loss)
    ), by=.(set, epoch)]
  },
  "stats::aggregate"={
    res <- stats::aggregate(
      loss ~ set + epoch, 
      loss.dt, 
      function(values)list(c(
        loss_length=length(values),
        loss_mean=mean(values), 
        loss_sd=sd(values))))
    data.frame(
      subset(res, select=-loss), 
      do.call(rbind, res$loss))
  },
  "plyr::ddply"={
    ddply(loss.dt, c("set", "epoch"), summarize,
          loss_length = length(loss),
          loss_mean = mean(loss),
          loss_sd = sd(loss))
  }
)
```






































```{r}
write.colors <- c(
  "data.table::fwrite" = "#D6604D",
  "pandas::to_csv" = "#BF812D"
)
n.rows <- 100
atime.write.vary.cols <- atime::atime(
  N = as.integer(10^seq(2, 6, by = 0.5)),
  setup = {
    set.seed(1)
    input.vec <- rnorm(n.rows * N)
    input.mat <- matrix(input.vec, n.rows, N)
    input.df <- data.frame(input.mat)
    py_run_string("import pandas as pd")
    py_set_attr(x = py, name = "input_mat_pd", value = input.mat)  
    py_run_string("input_df_pd = r_to_py(input_mat_pd)")
    input_df_pd = r_to_py(py$input_mat_pd)
  },
  seconds.limit = 0.01,
  "data.table::fwrite" = {
    data.table::fwrite(input.df, tempfile(), showProgress = FALSE)
  },
  "pandas::to_csv" = {
    py_run_string("input_df_pd.to_csv(file_path, index=False)")
  }
)
```



```{r}
refs.write.vary.cols <- atime::references_best(atime.write.vary.cols)
pred.write.vary.cols <- predict(refs.write.vary.cols)

gg.write.dt.pd <- plot(pred.write.vary.cols)+
  theme(text=element_text(size=20))+
  ggtitle(sprintf("Write real numbers to CSV, with pandas in Python and data.table in R, %d x N", n.rows))+
  scale_x_log10("N = number of columns to write")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=write.colors)+
  scale_color_manual(values=write.colors)
```

```{r}
png()
print(gg.write.dt.pd)
ggsave("gg.write.dt.pd.png", gg.write.dt.pd, width = 10, height = 6, units = "in", dpi = 300)
dev.off()

```



# Reading  csv Files with Pandas in Python and data.table in R

```{python}
import pandas as pd
import numpy as np

# Read the CSV file using pandas
input_df_pd = pd.read_csv(file_path)

# Convert the pandas DataFrame to a numpy array
input_mat_pd = input_df_pd.values

# Convert the numpy array to a list
input_vec_pd = input_mat_pd.flatten().tolist()

# Convert the list to a data.table object in R
input_df_r = reticulate::r.data.frame(input_vec_pd, stringsAsFactors=FALSE)
```


```{r}
file_path = 'data.csv'

write.colors <- c(
  "data.table::fread" = "#D6604D",
  "pandas::read_csv" = "#BF812D"
)

atime.read <- atime::atime(
  setup = {
    input.vec <- rnorm(n_rows * N)
    input.mat <- matrix(input.vec, n_rows, N)
    input.df <- data.frame(input.mat)
    input.csv <- tempfile()
    fwrite(input.df, input.csv)
    
  },
  seconds.limit = seconds.limit,
  "data.table::fread" = {
    data.table::fread(input.csv, showProgress = FALSE)
  },
  "pandas::read_csv" = {
    py_run_string("input_df_pd = pd.read_csv('", file_path, "')")
    py$input_df_pd
  }
)
```


```{r}
# Read the CSV file using data.table

write.colors <- c(
  "data.table::fread" = "#D6604D",
  "pandas::read_csv" = "#BF812D"
)

atime.read <- atime::atime(
  setup = {
    input.vec <- rnorm(n.rows*N)
    input.mat <- matrix(input.vec, n.rows, N)
    input.df <- data.frame(input.mat)
    input.csv <- tempfile()
    fwrite(input.df, input.csv)
    
  },
  seconds.limit = seconds.limit,
  "data.table::fread"={
    data.table::fread(input.csv, showProgress = FALSE)
  },
  "pandas::read_csv" = {
    py_run_string(paste0("input_df_pd = pd.read_csv('", file_path, "')"))
    py$input_df_pd
  }
)
```


```{r}
refs.read.vary.cols <- atime::references_best(atime.read.vary.cols)
pred.read.vary.cols <- predict(refs.read.vary.cols)

gg.read.pd <- plot(pred.read.vary.cols)+
  theme(text=element_text(size=20))+
  ggtitle(sprintf("Read real numbers to CSV, with pandas in Python and data.table in R, %d x N", n.rows))+
  scale_x_log10("N = number of columns to write")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=read.colors)+
  scale_color_manual(values=read.colors)
```

```{r}
png()
print(gg.read.pd)
ggsave("gg.read.pd.png", gg.read, width = 10, height = 6, units = "in", dpi = 300)
dev.off()

gg.read.pd

```














```{r}
write.colors <- c(
  "data.table::fread" = "#D6604D",
  "pandas::read_csv" = "#BF812D"
)


atime.read.vary.cols <- atime::atime(
  N = as.integer(10^seq(2, 6, by = 0.5)),
  setup = {
    input.vec <- rnorm(n.rows*N)
    input.mat <- matrix(input.vec, n.rows, N)
    input.df <- data.frame(input.mat)
    input.csv <- tempfile()
    fwrite(input.df, input.csv)
    
  },
  seconds.limit = seconds.limit,
  "data.table::fread"={
    data.table::fread(input.csv, showProgress = FALSE)
  },
  "pandas::read_csv" = {
    py_run_string("
    ")
  }
)


```


```{r}
refs.read.vary.cols <- atime::references_best(atime.read.vary.cols)
pred.read.vary.cols <- predict(refs.read.vary.cols)

gg.read <- plot(pred.read.vary.cols)+
  theme(text=element_text(size=20))+
  ggtitle(sprintf("Read real numbers to CSV, with pandas in Python and data.table in R, %d x N", n.rows))+
  scale_x_log10("N = number of columns to write")+
  scale_y_log10("Computation time (seconds)
median line, min/max band
over 10 timings")+
  facet_null()+
  scale_fill_manual(values=read.colors)+
  scale_color_manual(values=read.colors)
```

```{r}
print(gg.read)
ggsave("gg.read.png", gg.read, width = 10, height = 6, units = "in", dpi = 300)
dev.off()

gg.read
```

