---
layout: post
title: Visualizing prediction error
description: And clearly showing differences between algorithms
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2024-08-30-viz-pred-err"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=2)
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

In machine learning papers, we often need to compare the prediction
error/accuracy of different algorithms. This post explains how to do
that using data visualizations that are easy to read/interpret.

## Example: gradient descent learning of binary classification models

Classification is a machine learning problem that has been widely
studied for the last few decades. Binary classification is the special
case when there are two possible classes to predict (spam vs normal
email, cat vs dog in images, etc). To evaluate the prediction accuracy
of learned binary classification models, we often use the Area Under
the ROC Curve, because it allows fair comparison, even when the
distribution of labels is unbalanced (for example, 1% positive/spam
and 99% negative/normal email).

In our recent [JMLR'23](https://jmlr.org/papers/v24/21-0751.html)
paper, we proposed the AUM loss function, which can be used in
gradient descent learning algorithms, to optimize ROC curves.
Recently I did a computational experiment to compare this loss
function to others, via the following setup.

* We were motivated by the following question: Can gradient descent
  using the AUM loss result in faster computation of a model with good
  generalization properties? (large AUC on held-out data)
* We wanted to compare the AUM loss to the standard
  Logistic/Cross-Entropy loss used in classification, as well as the
  all pairs squared hinge loss which is a popular relaxation of the
  Mann-Whitney U statistic (and therefore a surrogate for ROC-AUC, for
  more info see [my paper with Kyle
  Rust](https://arxiv.org/abs/2302.11062)).
* We analyzed four different image classification data sets, in which
  each had 10 classes. So in each data set we converted to a binary
  problem, by using the first class (0) as the negative/0 class, and
  using all of the other classes as the positive/1 class. So each data
  set had about 10% negative and 90% positive labels.
* Data sets had different numbers of features, so were down-sampled to
  different sizes, in order to get train times which were similar
  between data sets. For example STL10 had the largest number of
  features (27648), so had the smallest train set; MNIST with only 784
  features had the largest train set.
* Source code used to compute the result, for a given loss function
  and learning rate, is in
  [data_Classif.py](https://github.com/tdhock/max-generalized-auc/blob/master/data_Classif.py)
* We tried a range of learning rates, `10^seq(-4,5)`, and three different loss functions, as can be seen in [data_Classif_batchtools.R](https://github.com/tdhock/max-generalized-auc/blob/master/data_Classif_batchtools.R).
* For each algorithm, data set, random seed, and learning rate, we
  used torch to initialize a random linear model, and then did 100000
  epochs of gradient descent learning with constant step size (learning rate).
* Then for each algorithm, data set, and random seed, we select only
  the epoch/iteration and step size that achieved the max AUC on the
  validation set.

The results can be read from
[data_Classif_batchtools_best_valid.csv](/assets/data_Classif_batchtools_best_valid.csv),
as shown in the code below:

```{r}
library(data.table)
(best.dt <- fread("../assets/data_Classif_batchtools_best_valid.csv"))
```

## Easy dot plot visualization

A visualization method which is simple to code is shown below:

```{r dot}
library(ggplot2)
ggplot()+
  geom_point(aes(
    auc, loss),
    data=best.dt)+
  facet_grid(. ~ data.name)
```

The plot above shows four dots for each loss and data set. Already we
can see that loss=AUM tends to have the largest `auc` values, in each
data set. Rather than showing the four data sets using the same X axis
scale, we can show more subtle differences, by allowing each data set
to have its own X axis scale, as below.

```{r dot-scale-free}
ggplot()+
  geom_point(aes(
    auc, loss),
    data=best.dt)+
  facet_grid(. ~ data.name, scales="free", labeller=label_both)
```

Above we see each data set has its own scale, but some of the tick
marks are not readable. This can be fixed by specifying non-default
panel spacing values in the theme, as below.

```{r dot-panel-space}
ggplot()+
  geom_point(aes(
    auc, loss),
    data=best.dt)+
  facet_grid(. ~ data.name, scales="free", labeller=label_both)+
  theme(
    panel.spacing=grid::unit(1.5, "lines"))
```

In the plot above there is now another issue: the last X tick mark
goes off the right edge of the plot. To fix that we need to adjust the
plot margin, as below.

```{r dot-plot-margin}
ggplot()+
  geom_point(aes(
    auc, loss),
    data=best.dt)+
  facet_grid(. ~ data.name, scales="free", labeller=label_both)+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))
```

The plot above looks like a reasonable summary of the results, but the labels could be improved.

* We could explain more details about each algorithm in the Y axis labels.
* We could simplify the panel/facet variable names, `data.name` above, to simply `Data`, and add `N` for each.
* We could use the more common capital `AUC` rather than lower `auc`, and explain that it is the max on the validation set.

```{r dot-labels}
loss2show <- rev(c(
  Logistic="Logistic/Cross-entropy\n(classic baseline)",
  SquaredHinge="All Pairs Squared Hinge\n(recent alternative)",
  AUM="AUM=Area Under Min(FP,FN)\n(proposed complex loss)",
  NULL))
Loss_factor <- function(L){
  factor(L, names(loss2show), loss2show)
}
best.dt[, `:=`(
  Loss = Loss_factor(loss),
  Data = data.name
)]
ggplot()+
  geom_point(aes(
    auc, Loss),
    data=best.dt)+
  facet_grid(. ~ N + Data, scales="free", labeller=label_both)+
  scale_x_continuous(
    "Max validation AUC (4 random initializations)")+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))
```

Note the `Loss` names in code above is arranged to be consistent with
their display in the plot above: the `Loss` column factor levels come
from `loss2show`, and are used to determine the order of display of
the tick marks in the Y axis.

Similarly, the facets/panels are ordered by the first facet variable,
`N` (smallest N for STL10 on the left, largest N for MNIST on the
right). This order is different than previous plots, which had facets
in alphabetical order (CIFAR10 left, STL10 right). To display an
alternative facet/panel order, you would have to create a factor
variable with the levels in the desired order, similar to what we did
with `Loss` values for the Y axis above. (exercise for the reader)

## Display mean and standard deviation

Whereas in the previous section we displayed each random seed as a
different dot, below we compute and plot the mean and SD over random seeds.
And while we are at it, we can also compute the range (min and max), for the AUC as well as for the number of gradient descent epochs (which is the same as the number of steps here, since we used full gradient method, batch size = N).

```{r}
(best.wide <- dcast(
  best.dt,
  N + Data + Loss ~ .,
  list(mean, sd, length, min, max),
  value.var=c("auc","step_number")))
```

In the result table above, we also compute the `length` to double
check that the mean/etc was indeed taken over the four random seeds.
The code/plot below only uses the mean.

```{r mean-only}
ggplot()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_x_continuous(
    "Max validation AUC (Mean over 4 random initializations)")
```

The plot above is not very useful for comparing the different Loss
functions, because it only shows the mean, without showing any measure
of the variance. So we can not say if any Loss is significantly more
or less accurate than any other (we would need error bars or
confidence intervals to do that). We fix that in the code/plot below,
by computing `lo` and `hi` limits to display based on the SD.

```{r mean-sd}
best.wide[, `:=`(
  lo = auc_mean-auc_sd,
  hi = auc_mean+auc_sd
)]
ggplot()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    lo, Loss,
    xend=hi, yend=Loss),
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_x_continuous(
    "Max validation AUC (Mean ± SD over 4 random initializations)")
```

The plot above is much better, because it shows the SD as well as the mean.
We can see that AUM is significantly more accurate than the others in all of the data sets, except perhaps MNIST, in which the All Pairs Squared Hinge looks only slightly worse.
We could additionally write the values of mean and SD, as below.

```{r mean-sd-text-mid}
ggplot()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    lo, Loss,
    xend=hi, yend=Loss),
    data=best.wide)+
  geom_text(aes(
    auc_mean, Loss,
    label=sprintf(
      "%.4f±%.4f", auc_mean, auc_sd)),
    size=3,
    vjust=1.5,
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_x_continuous(
    "Max validation AUC (Mean ± SD over 4 random initializations)")
```

Above, only some of the text is readable, and others go outside of the panels.
To fix this, we can use `aes(hjust)`:

* The default `hjust=0.5`, used in the plot above, draws the text centered around the mean value.
* if the mean is less than the mid point of the panel X axis, then we
  can use `hjust=0` which means text will be left justified with the
  mean value as the limit. In other words, the text will start writing
  from the mean value, and go to the right of the mean value, but is
  guaranteed to not go left of the mean value, so it will not go off
  the panel to the left.
* otherwise, we can use `hjust=1` which means text will be right
  justified to the mean value.
  
To get this scheme to work, we need to compute the mid-point on the X
axis (auc) of each panel, which we do in the code below.

```{r mean-sd-aes-hjust}
best.wide[
, mid := (min(lo)+max(hi))/2
, by=Data]
ggplot()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    lo, Loss,
    xend=hi, yend=Loss),
    data=best.wide)+
  geom_text(aes(
    auc_mean, Loss,
    hjust=ifelse(auc_mean<mid, 0, 1),
    label=sprintf(
      "%.4f±%.4f", auc_mean, auc_sd)),
    size=3,
    vjust=1.5,
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_x_continuous(
    "Max validation AUC (Mean ± SD over 4 random initializations)")
```

The plot above has text that can be read to determine the mean and SD values of each loss, in each data set.

## P-value plot

To conclusively answer the question about whether AUM results in
larger Max validation AUC than the next best loss, we would need to
use a statistical significance test.
First we compute the best two loss functions for each dataset, as below.

```{r}
(best.two <- best.wide[
  order(N,-auc_mean)
][
, rank := rank(-auc_mean)
, by=.(N,Data)
][rank <= 2, .(N,Data,Loss,auc_mean,rank)])
```

Below we join with the original data.

```{r}
(best.two.join <- best.dt[best.two, .(N,Data,Loss,rank,seed,auc), on=.(N,Data,Loss)])
```

Below we reshape, which is required before doing the T-test in R.

```{r}
(best.two.wide <- dcast(best.two.join, N+Data+seed~rank, value.var="auc"))
```

Below we run T-tests to see if the top ranked AUC is significantly
greater than the next ranked AUC, for each data set.

```{r}
(test.dt <- best.two.wide[, {
  paired <- t.test(`1`, `2`, alternative="greater", paired=TRUE)
  unpaired <- t.test(`1`, `2`, alternative="greater", paired=FALSE)
  data.table(
    mean.of.diff=paired$estimate, p.paired=paired$p.value,
    m1=unpaired$estimate[1], m2=unpaired$estimate[2], p.unpaired=unpaired$p.value)
}, by=.(N,Data)])
```

The table above summarizes the results of the T-tests.

* The paired T-test is more powerful (gives you smaller P-values), but
  only works when you actually have paired observations, as we do here
  (AUC was computed for each loss and each random seed). Its
  `estimate` is the mean of the differences between each pair of AUC
  values.
* The unpaired T-test can be seen to have larger (less significant)
  P-values, but it may be useful to run as well, because `estimate`
  contains mean values for each of the two samples (here the two
  different loss functions).

To display the test result below we use a rectangle.

```{r p-value}
p.color <- "red"
text.size <- 3
ggplot()+
  theme_bw()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_rect(aes(
    xmin=m2, xmax=m1,
    ymin=-Inf, ymax=Inf),
    fill=p.color,
    alpha=0.5,
    data=test.dt)+
  geom_text(aes(
    m1, Inf, label=sprintf("Diff=%.4f P=%.4f ", mean.of.diff, p.paired)),
    data=test.dt,
    size=text.size,
    vjust=1.2,
    hjust=1)+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    lo, Loss,
    xend=hi, yend=Loss),
    data=best.wide)+
  geom_text(aes(
    auc_mean, Loss,
    hjust=ifelse(auc_mean<mid, 0, 1),
    label=sprintf(
      "%.4f±%.4f", auc_mean, auc_sd)),
    size=text.size,
    vjust=1.5,
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_y_discrete(
    "Loss")+
  scale_x_continuous(
    "Max validation AUC (Mean ± SD over 4 random initializations)")
```

In the plot above, we show the p-value, which is typically intepreted by comparing with the traditional significance threshold of 0.05, which corresponds to a 5% false positive rate. Seeing a p-value of 0.05 means that you have observed a difference that you would see about 5% of the time (simply due to random variation/noise), if there is really no difference between methods. So if we are trying to argue that one algorithm is better, then we want to see small p-values, which mean that we have observed differences that are so large, that it would be extremely unlikely to see such a difference by random chance.
* in STL10 there is a highly significant difference (p=0.002, order of magnitude less than 0.05).
* in CIFAR10 there is a significant difference (p=0.02 is less than 0.05),
* in FashionMNIST there is a slight difference (but we do not say significant because p=0.07 is still larger than 0.05),
* the difference in MNIST is not statistically significant (p=0.17 much larger than 0.05), 

Above we compared the best to the next best. An alternative is to compare the proposed to others, which we code below.
First we reshape wider, as below.

```{r}
(best.loss.wide <- dcast(best.dt, N + Data + seed ~ loss, value.var="auc"))
```

The table above has one column for each method/loss.
Then we define the proposed method column, and reshape the other columns taller, as below.

```{r}
proposed.loss <- "AUM"
(other.loss.vec <- best.dt[loss!=proposed.loss, unique(loss)])
(best.loss.tall <- melt(
  best.loss.wide,
  measure.vars=other.loss.vec,
  variable.name="other.loss",
  value.name="other.auc"))
```

The table above has a column for the Max Validation AUC of the proposed method (AUM), and has the Max Validation AUC of the other methods in the `other.auc` column. We can then run the T-test for each value of `other.loss`, using the code below.

```{r}
(test.proposed <- best.loss.tall[, {
  paired <- t.test(AUM, other.auc, alternative="greater", paired=TRUE)
  unpaired <- t.test(AUM, other.auc, alternative="greater", paired=FALSE)
  data.table(
    mean.of.diff=paired$estimate, p.paired=paired$p.value,
    mean.proposed=unpaired$estimate[1], mean.other=unpaired$estimate[2], p.unpaired=unpaired$p.value)
}, by=.(N,Data,other.loss)])
```

The table above has a row for each T-test, one for each data set and other loss function (other than the proposed AUM).
The final step is to visualize these data on the plot, as in the code below.

```{r p-others}
test.proposed[
, other.Loss := Loss_factor(other.loss)
]
ggplot()+
  theme_bw()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_segment(aes(
    mean.proposed, other.Loss,
    xend=mean.other, yend=other.Loss),
    color=p.color,
    alpha=0.5,
    linewidth=3,
    data=test.proposed)+
  geom_text(aes(
    mean.proposed, other.Loss,
    label=sprintf("Diff=%.4f P=%.4f", mean.of.diff, p.paired)),
    color=p.color,
    size=text.size,
    vjust=-0.5,
    hjust=1,
    data=test.proposed)+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    lo, Loss,
    xend=hi, yend=Loss),
    data=best.wide)+
  geom_text(aes(
    auc_mean, Loss,
    hjust=ifelse(auc_mean<mid, 0, 1),
    label=sprintf(
      "%.4f±%.4f", auc_mean, auc_sd)),
    size=text.size,
    vjust=1.5,
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_y_discrete(
    "Loss")+
  scale_x_continuous(
    "Max validation AUC (Mean ± SD over 4 random initializations)")
```

We can see in the plot above that there is red text and segments drawn to emphasize the p-value, and how it was computed, for each method other than the proposed AUM. There are a couple of issues though

* The Y axis tick mark ordering is no longer as expected, because ggplot2 drops factor levels by default, if some are not present in a given data layer. To avoid that we can use `scale_y_discrete(drop=FALSE)`.
* Some p-values are smaller than the limit of 4 decimal places, so we need a different method to display them, for example writing `P<0.0001` when that is true.

```{r p-others-no-drop}
ggplot()+
  theme_bw()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_segment(aes(
    mean.proposed, other.Loss,
    xend=mean.other, yend=other.Loss),
    color=p.color,
    alpha=0.5,
    linewidth=3,
    data=test.proposed)+
  geom_text(aes(
    mean.proposed, other.Loss,
    label=paste(
      sprintf("Diff=%.4f", mean.of.diff),
      ifelse(
        p.paired<0.0001, "P<0.0001",
        sprintf("P=%.4f", p.paired)))),
    color=p.color,
    size=text.size,
    vjust=-0.5,
    hjust=1,
    data=test.proposed)+
  geom_point(aes(
    auc_mean, Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    lo, Loss,
    xend=hi, yend=Loss),
    data=best.wide)+
  geom_text(aes(
    auc_mean, Loss,
    hjust=ifelse(auc_mean<mid, 0, 1),
    label=sprintf(
      "%.4f±%.4f", auc_mean, auc_sd)),
    size=text.size,
    vjust=1.5,
    data=best.wide)+
  facet_grid(. ~ N + Data, labeller=label_both, scales="free")+
  scale_y_discrete(
    "Loss",
    drop=FALSE)+
  scale_x_continuous(
    "Max validation AUC (Mean ± SD over 4 random initializations)")
```

## Display accuracy and computation time in scatter plot

In the plots above, we only examined prediction accuracy. Below we
additionally examine the number of iterations/epochs of gradient
descent, in order to determine which loss function results in fastest learning.

```{r scatter-grid}
ggplot()+
  theme_bw()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    legend.key.spacing.y=grid::unit(1, "lines"),
    axis.text.x=element_text(angle=30, hjust=1),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_point(aes(
    auc_mean, step_number_mean,
    color=Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    auc_min, step_number_mean,
    color=Loss,
    xend=auc_max, yend=step_number_mean),
    data=best.wide)+
  geom_segment(aes(
    auc_mean, step_number_min,
    color=Loss,
    xend=auc_mean, yend=step_number_max),
    data=best.wide)+
  facet_grid(~N+Data, labeller=label_both, scales="free")+
  scale_y_log10(
    "Gradient descent epochs\n(using best learning rate)")+
  scale_x_continuous(
    "Best validation AUC (dot=mean, segments=range over 4 random initializations)")
```

In the plot above, we again see Best validation AUC on the X axis, and
we see number of epochs on the Y axis. So we can see that the AUM loss
has largest Best validation AUC (so AUM can be more accurate), as well
as comparable/smaller number of epochs (so AUM can be faster). 

The plot above uses `facet_grid` which forces the Y axis to be the
same in each plot, even though we specified `scales="free"` (which
actually only affects the X axis in this case). Below we use
`facet_wrap` instead, in order to zoom in on the details of each
panel:

```{r scatter}
ggplot()+
  theme_bw()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    legend.key.spacing.y=grid::unit(1, "lines"),
    axis.text.x=element_text(angle=30, hjust=1),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_point(aes(
    auc_mean, step_number_mean,
    color=Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    auc_min, step_number_mean,
    color=Loss,
    xend=auc_max, yend=step_number_mean),
    data=best.wide)+
  geom_segment(aes(
    auc_mean, step_number_min,
    color=Loss,
    xend=auc_mean, yend=step_number_max),
    data=best.wide)+
  facet_wrap(~N+Data, nrow=1, labeller=label_both, scales="free")+
  scale_y_log10(
    "Gradient descent epochs\n(using best learning rate)")+
  scale_x_continuous(
    "Best validation AUC (dot=mean, segments=range over 4 random initializations)")
```

The plot above has a different Y axis for each panel/facet, due to `facet_wrap(scales="free")`.
This allows us to zoom in to see more detailed comparisons in each panel/facet.
However there are a couple of details worth fixing, for improved clarity:

* The default color scale in ggplot2 results in a blue and green which
  can be difficult to distinguish, so I recommend using Cynthia
  Brewer's color palettes, such as
  [Dark2](https://colorbrewer2.org/#type=qualitative&scheme=Dark2&n=3).
* Some data/segments go outside the range of the Y axis ticks, which
  can make the Y axis difficult to read, so we can use `geom_blank` to
  increase the Y axis range.

```{r scatter-improved}
dput(RColorBrewer::brewer.pal(3,"Dark2"))
loss.colors <- c("black", "#D95F02", "#7570B3")
names(loss.colors) <- loss2show
p <- function(Data,x,y)data.table(Data,x,y)
(blank.Data <- rbind(
  p("CIFAR10",0.8,100),
  p("MNIST",0.99,c(10,100000)),
  p("STL10",0.8,30)))
(blank.Data.N <- best.wide[, .(N,Data)][blank.Data, on=.(Data), mult="first"])
ggplot()+
  theme_bw()+
  theme(
    plot.margin=grid::unit(c(0,1,0,0), "lines"),
    legend.key.spacing.y=grid::unit(1, "lines"),
    axis.text.x=element_text(angle=30, hjust=1),
    panel.spacing=grid::unit(1.5, "lines"))+
  geom_blank(aes(x, y), data=blank.Data.N)+
  geom_point(aes(
    auc_mean, step_number_mean,
    color=Loss),
    shape=1,
    data=best.wide)+
  geom_segment(aes(
    auc_min, step_number_mean,
    color=Loss,
    xend=auc_max, yend=step_number_mean),
    data=best.wide)+
  geom_segment(aes(
    auc_mean, step_number_min,
    color=Loss,
    xend=auc_mean, yend=step_number_max),
    data=best.wide)+
  facet_wrap(~N+Data, nrow=1, labeller=label_both, scales="free")+
  scale_color_manual(
    values=loss.colors)+
  scale_y_log10(
    "Gradient descent epochs\n(using best learning rate)")+
  scale_x_continuous(
    "Best validation AUC (dot=mean, segments=range over 4 random initializations)")
```

Note above how the Y axes have expanded, and now there are tick marks
above the range of the data/segments, which makes the Y axis easier to
read. Also the color legend has changed: I use black for the proposed
method, and two other colors from Cynthia Brewer's Dark2 palette.

## Conclusions

Our goal was to explore how machine learning error/accuracy rates can be visualized, in order to compare different algorithms. 
We discussed various techniques for creating visualizations that make it easy for the reader to compare different algorithms.

## Session info

```{r}
sessionInfo()
```
