---
layout: post
title: AUC and AUM in torch
description: Demonstration of auto-grad
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2024-10-10-torch-roc-aum"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
conda.env <- "2023-08-deep-learning"
Sys.setenv(RETICULATE_PYTHON=sprintf(if(.Platform$OS.type=="unix")
  ##"/home/tdhock/.local/share/r-miniconda/envs/%s/bin/python"
  "/home/tdhock/miniconda3/envs/%s/bin/python"
  else "~/Miniconda3/envs/%s/python.exe", conda.env))
reticulate::use_condaenv(conda.env, required=TRUE)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to show how to use torch to compute ROC-AUC
(classic evaluation metric for binary classification) and AUM (Area
Under Min of False Positive and False Negative rates, our newly
proposed surrogate loss for ROC curve optimization).

```{python echo=FALSE}
repo_dir = r["repo.dir"]
fig_path = r["fig.path"]
import warnings
def p9_save(g, name):
    out_png = fig_path+name+".png"
    with warnings.catch_warnings():
        warnings.simplefilter("ignore")
        g.save(out_png)
    web_link = out_png.replace(repo_dir, "")
    print('![plot of %s](%s)'%(name, web_link))
# work-around for rendering plots under windows, which hangs within
# emacs python shell: instead write a PNG file and view in browser.
import os
import webbrowser
on_windows = os.name == "nt"
rendering = r.rendering if 'r' in dir() else False
using_agg = on_windows and not rendering
if using_agg:
    import matplotlib
    matplotlib.use("agg")
def show(g, name):
    if not using_agg:
        return p9_save(g, name)
    g.save("tmp.png")
    webbrowser.open('tmp.png')
```

## Introduction: binary classification and zero-one loss

In supervised binary classification, our goal is to learn a function
`f` using training inputs/features `x`, and outputs/labels `y`, such
that `f(x)=y` (on new/test data). To illustrate, the code below
defines a data set with four samples:

```{python}
import torch
four_labels = torch.tensor([-1,-1,1,1])
four_pred = torch.tensor([2.0, -3.5, -1.0, 1.5])
```

Note that `four_pred` in the code above is a vector of four `f(x)`
values (real numbers=predicted scores), and `four_labels` is a vector
of four `y` values (-1 for negative class, 1 for positive class). How
do we compute if these are good predictions for these labels?
To answer that question, we use an objective function, which computes a number that determines how good/bad our predictions are, relative to the labels in our data set.
The classic objective function for
evaluation is the zero-one loss, which computes the number/proportion
of labels which are mis-classified (in the test set).
The zero-one loss first computes the real `f(x)` scores to a class by thresholding at zero, and then computes the proportion of predicted classes do not match the label classes.

```{python}
def score_to_class(pred_score):
    return torch.where(pred_score < 0, -1, 1)
def pred_is_correct(pred_score, label_vec):
    pred_class = score_to_class(pred_score)
    return pred_class == label_vec
def zero_one_loss(pred_score, label_vec):
    return torch.where(pred_is_correct(pred_score, label_vec), 0, 1)
import pandas as pd
zero_one_df = pd.DataFrame({
    "label":four_labels,
    "score":four_pred,
    "zero_one_loss":zero_one_loss(four_pred, four_labels)
})
zero_one_df
```

The output above is a table with one row for each sample. The `zero_one_loss` column shows that if the sign of the score matches the label, then the zero-one loss is 0 (correct), otherwise it is 1 (incorrect). Below we take the mean of that vector, to get the proportion of incorrectly predicted labels:

```{python}
zero_one_df.zero_one_loss.mean()
```

## Imbalanced data and confusion matrices

In imbalanced data, for example with 99% negative and 1% positive
labels, other objective functions like AUC (Area Under ROC Curve) are
used for evaluation. The [Receiver Operating Characteristic (ROC)
curve](https://en.wikipedia.org/wiki/Receiver_operating_characteristic)
is a plot of True Positive Rate, as a function of False Positive
Rate. What are those?

* True Positive is when `f(x) > 0` (predict positive), and `y=0`
  (label positive), this is good! The opposite/negative prediction,
  `f(x) < 0`, with the same positive label, is called a False Negative
  (bad).
* False Positive is when `f(x) > 0` (predict positive), but `y=0`
  (label negative), this is bad! The opposite/positive prediction,
  `f(x) < 0`, with the same negative label, is called a True Negative
  (good).

These terms are also names for the entries in the confusion matrix, 

|          | prediction     |                |
| label    | `f(x) < 0`     | `f(x) > 0`     |
|----------|----------------|----------------|
| `y = -1` | True Negative  | False Positive |
| `y = 1`  | False Negative | True Positive  |

Below we add a confusion column with these names to our previous table:

```{python}
import numpy as np
def TF_pos_neg(pred_score, label_vec):
    pred_is_positive = score_to_class(pred_score) == 1
    pred_name = np.where(pred_is_positive, "Positive", "Negative")
    T_or_F = pred_is_correct(pred_score, label_vec).numpy()
    T_or_F_space = np.char.add(T_or_F.astype(str), " ")
    return np.char.add(T_or_F_space, pred_name)
zero_one_df["confusion"]=TF_pos_neg(four_pred, four_labels)
zero_one_df
```

The `confusion` column shows where each sample would appear in the
confusion matrix.  When we compute a ROC curve, we compute two
quantities. First, the True Positive Rate (TPR) is the number of true
positives, divided by the number of possible true positives (number of
positive labels).

```{python}
def get_TPR(df):
    return (df.confusion=="True Positive").sum()/(df.label==1).sum()
get_TPR(zero_one_df)
```

Next, the False Positive Rate (FPR) is the number of false positives,
divided by the number of possible false positives (number of negative
labels).

```{python}
def get_FPR(df):
    return (df.confusion=="False Positive").sum()/(df.label== -1).sum()
get_FPR(zero_one_df)
```

Both True Positive Rate and False Positive Rate are 0.5 in this
example, because there are two of each label; one of each label is
mis-classified.

## ROC curve computation

When we compute a ROC curve, we need to consider all possible
thresholds of the predicted score `f(x)`, not just the default
threshold of zero. In other words, the ROC curve is a 2D parametric
function: `[FPR(c),TPR(c)]`, where the parameter is `c`, a constant
that could be added to predicted scores, before using the threshold of
zero to determine whether we should predict if the class is positive
(`f(x)+c > 0`) or negative (`f(x)+c < 0`).
If `c` is very large, then `f(x_i)+c > 0` for all data `i`, so
  `FPR(c)=1` and `TPR(c)=1` (all data classified as positive as in the
  code below).


```{python}
def error_one_constant(constant):
    pred_const = four_pred+constant
    return pd.DataFrame({
        "label":four_labels,
        "score":four_pred,
        "zero_one_loss":zero_one_loss(pred_const, four_labels),
        "confusion":TF_pos_neg(pred_const, four_labels)
    })
error_one_constant(5)
```

If `c` is very small, then `f(x_i)+c < 0` for all data `i`, so
 `FPR(c)=0` and `TPR(c)=0` (all data classified as negative as in the code below).

```{python}
error_one_constant(-10)
```

Note that there are infinitely many different constants which we could add to the predicted values, which result in the same `confusion` values, and therefore the same point on the ROC curve. For example, below is another small constant which results in all negative predictions:

```{python}
error_one_constant(-20)
```

We can compute a ROC curve (inefficiently, quadratic time in the
number of samples) by looping over constants, and repeating the
computations, as in the code below.

```{python}
four_roc_df_list = []
constant_vec = list(-four_pred)+[-torch.inf]
constant_vec.sort()
def one_roc_point(constant):
    one_df = error_one_constant(constant)
    return pd.DataFrame({
        "constant":[float(constant)],
        "TPR":get_TPR(one_df),
        "FPR":get_FPR(one_df)
    })
roc_inefficient_df = pd.concat([
    one_roc_point(constant) for constant in constant_vec
])
roc_inefficient_df
```

The table above has one row for each point on the ROC curve, which is visualized using the code below.

```{python}
import plotnine as p9
p9.ggplot()+\
    p9.coord_equal()+\
    p9.geom_line(
        p9.aes(
            x="FPR",
            y="TPR",
        ),
        data=roc_inefficient_df
    )+\
    p9.geom_point(
        p9.aes(
            x="FPR",
            y="TPR",
        ),
        data=roc_inefficient_df
    )
```

The figure above shows a ROC curve with 5 points (the maximum number
of points for 4 data; there could be fewer if there are ties in the
predicted scores vector). We mentioned above that it was computed
inefficiently, which is caused by the for loop over constants. To
avoid that loop (quadratic time overall), we can instead sort the
predicted scores (log-linear time overall), as in the code below:

```{python}
def ROC_curve(pred_tensor, label_tensor):
    """Receiver Operating Characteristic curve.
    """
    is_positive = label_tensor == 1
    is_negative = label_tensor != 1
    fn_diff = torch.where(is_positive, -1, 0)
    fp_diff = torch.where(is_positive, 0, 1)
    thresh_tensor = -pred_tensor.flatten()
    sorted_indices = torch.argsort(thresh_tensor)
    fp_denom = torch.sum(is_negative) #or 1 for AUM based on count instead of rate
    fn_denom = torch.sum(is_positive) #or 1 for AUM based on count instead of rate
    sorted_fp_cum = fp_diff[
        sorted_indices].cumsum(axis=0)/fp_denom
    sorted_fn_cum = -fn_diff[
        sorted_indices].flip(0).cumsum(axis=0).flip(0)/fn_denom
    sorted_thresh = thresh_tensor[sorted_indices]
    sorted_is_diff = sorted_thresh.diff() != 0
    sorted_fp_end = torch.cat([sorted_is_diff, torch.tensor([True])])
    sorted_fn_end = torch.cat([torch.tensor([True]), sorted_is_diff])
    uniq_thresh = sorted_thresh[sorted_fp_end]
    uniq_fp_after = sorted_fp_cum[sorted_fp_end]
    uniq_fn_before = sorted_fn_cum[sorted_fn_end]
    FPR = torch.cat([torch.tensor([0.0]), uniq_fp_after])
    FNR = torch.cat([uniq_fn_before, torch.tensor([0.0])])
    return {
        "FPR":FPR,
        "FNR":FNR,
        "TPR":1 - FNR,
        "min(FPR,FNR)":torch.minimum(FPR, FNR),
        "min_constant":torch.cat([torch.tensor([-torch.inf]), uniq_thresh]),
        "max_constant":torch.cat([uniq_thresh, torch.tensor([torch.inf])])
    }
roc_efficient_df = pd.DataFrame(ROC_curve(four_pred, four_labels))
roc_efficient_df
```

The table above also has one row for each point on the ROC curve (same as the previous table), and it has additional columns which we will use later:

* `FNR=1-TPR` is the False Negative Rate,
* `min` is the minimum of `FPR` and `FNR`,
* and `min_constant`, `max_constant` give the range of constants which result in the corresponding error values (`min_constant` is actually the same as `roc_inefficient_df.constant`). For example, the second row means that adding any constant between -2 and -1.5 results in predicted classes that give FPR=0.5 and TPR=0, as we can verify using our previous function in the code below:

```{python}
one_roc_point(-1.7)
```

Exercise for the reader: try `one_roc_point` with some other constants, and check to make sure the results are consistent with `roc_efficient_df` above. 

## ROC curve interpretation and examples

How do we interpret the ROC curve? An ideal ROC curve would

* start at the bottom left (FPR=TPR=0, every sample predicted negative), 
* and then go straight to the upper left (FPR=0,TPR=1, every sample
  predicted correctly),
* and then go straight to the upper right (FPR=TPR=1, every sample
  predicted positive),
* so it would have an Area Under the Curve of 1.

So when we do ROC analysis, we can look at the curves, to see how close they get to the upper left, or we can just compute the Area Under the Curve (larger is better). To compute the Area Under the Curve, we use the trapezoidal area formula, which amounts to summing the rectangle and triangle under each segment of the curve, as in the code below.

```{python}
def ROC_AUC(pred_tensor, label_tensor):
    roc = ROC_curve(pred_tensor, label_tensor)
    FPR_diff = roc["FPR"][1:]-roc["FPR"][:-1]
    TPR_sum = roc["TPR"][1:]+roc["TPR"][:-1]
    return torch.sum(FPR_diff*TPR_sum/2.0)
ROC_AUC(four_pred, four_labels)
```

How do we get an ideal ROC curve, with AUC=1? We need to have predicted scores that are smaller for negative labels, and larger for positive labels. If any score for a negative label is greater than or equal to a score for a positive label, then that will result in a sub-optimal ROC curve. Note that positive scores for negative labels still can result in an ideal ROC curve, as long as those scores are less than the scores for the positive labels. Three example predicted score vectors are defined below:

```{python}
pred_dict = {
    "ideal":[1.0, 2, 3, 4],
    "constant":[9.0, 9, 9, 9],
    "anti-learning":[4.0, 3, 2, 1],
}
example_pred_df = pd.DataFrame(pred_dict)
example_pred_df["label"] = four_labels
example_pred_df
```

The constant predictions result in the worst ROC curve, which jumps
from FPR=TPR=0 to FPR=TPR=1. The anti-learning predictions have large scores for negative labels, and small scores for positive predictions (called anti-learning because we could just invert the score to obtain an ideal score). 
The code below visualizes the ROC curve for each of these three predictions:

```{python}
example_roc_df_list = []
for model, pred_list in pred_dict.items():
    pred_tensor = torch.tensor(pred_list)
    one_roc = ROC_curve(pred_tensor, four_labels)
    one_roc['model']=model
    one_roc["AUC"]=ROC_AUC(pred_tensor, four_labels).numpy()
    example_roc_df_list.append(pd.DataFrame(one_roc))
example_roc_df = pd.concat(example_roc_df_list)
p9.ggplot()+\
    p9.facet_grid(". ~ AUC + model", labeller="label_both")+\
    p9.coord_equal()+\
    p9.geom_line(
        p9.aes(
            x="FPR",
            y="TPR",
        ),
        data=example_roc_df
    )+\
    p9.geom_point(
        p9.aes(
            x="FPR",
            y="TPR",
        ),
        data=example_roc_df
    )
```

We can see in the figure above three ROC curves, and their
corresponding AUC values. Two of the curves have five points, which is
the max possible for four samples (no tied scores); the curve for
model=constant has two points, which is the min possible (all tied
scores).

## Derivative of log loss

Many algorithms for learning `f` are based on gradient descent, which
attempts to minimize a loss function that measures how well `f` works
for predicting the train data. But gradient descent does not work with
the zero-one loss, because its derivative is zero almost everywhere
(no information to learn how to update parameters to get better
predictions). So instead people use the logistic (cross-entropy)
loss, or the hinge loss, which have linear tails (gradient is -1 when
prediction is bad, which tells the learning algorithm to increase
those predictions). It is computed as in the code below, for each observation:

```{python}
def log_loss(pred_tensor, label_tensor):
    return torch.log(1+torch.exp(-label_tensor*pred_tensor))
log_loss(four_pred, four_labels)
```

The code below computes the mean log loss over all samples:

```{python}
def mean_log_loss(pred_tensor, label_tensor):
    return log_loss(pred_tensor, label_tensor).mean()
mean_log_loss(four_pred, four_labels)
```

And the code below computes the proportion incorrectly classified samples (error rate):

```{python}
def prop_incorrect(pred_tensor, label_tensor):
    return zero_one_loss(pred_tensor, label_tensor).float().mean()
prop_incorrect(four_pred, four_labels)
```

These functions and their gradients can be visualized using the code
below.

```{python}
obj_grad_df_list = []
pred_grid = np.arange(-4.0, 4, 0.5)
for objective in "prop_incorrect", "mean_log_loss":
    ofun = eval(objective)
    for pred_val in pred_grid:
        pred_tensor=torch.tensor(pred_val)
        pred_tensor.requires_grad = True
        label_tensor = torch.tensor([1])
        loss = ofun(pred_tensor, label_tensor)
        try:
            loss.backward()
            g_vec = pred_tensor.grad
        except:
            g_val = 0.0 if pred_val != 0 else torch.nan
            g_vec = torch.tensor([g_val])
        obj_grad_df_list.append(pd.DataFrame({
            "predicted_score":pred_val,
            "objective":objective,
            "function":["output","derivative"],
            "value":torch.cat([
                loss.reshape(1),
                g_vec.reshape(1)
            ]).detach().numpy(),
        }))
obj_grad_df = pd.concat(obj_grad_df_list)
p9.ggplot()+\
    p9.theme_bw()+\
    p9.geom_hline(
        p9.aes(
            yintercept="value"
        ),
        data=pd.DataFrame({"value":[0]}),
        color="grey"
    )+\
    p9.geom_point(
        p9.aes(
            x="predicted_score",
            y="value",
        ),
        data=obj_grad_df
    )+\
    p9.facet_grid("function ~ objective", labeller="label_both")
```

The figure above shows that the log loss (for a positive label)
outputs linear tails for negative predicted scores, which means a
derivative of -1 that can be used to find model parameters that result
in better predictions, in the context of gradient descent learning
algorithms. In contrast, the proportion incorrect is a step function,
with derivative 0 almost everywhere (except when predicted score is 0,
where it is undefined).

## Proposed AUM loss for ROC optimization

The log loss can be intepreted as a differentiable surrogate for the
zero-one loss. There is no derivative for the zero-one loss, so the
next best thing is to learn using the log loss, which is a convex/L1
relaxation (L1 meaning the log loss has linear tails and constant
gradient). Is it possible to do something similar with the ROC AUC?
Yes! Recently [in JMLR23](https://jmlr.org/papers/v24/21-0751.html)
we proposed a new loss function called the AUM, Area Under Min of
False Positive and False Negative rates. We showed that is can be
interpreted as a L1 relaxation of the sum of min of False Positive and
False Negative rates, over all points on the ROC curve. We
additionally showed that AUM is piecewise linear, and differentiable
almost everywhere, so can be used in gradient descent learning
algorithms. Finally, we showed that minimizing AUM encourages points
on the ROC curve to move toward the upper left, thereby encouraging
large AUC. Computation of the AUM loss requires first
computing ROC curves (same as above), which we visualize below using a
letter next to each point on the curve:

```{python}
roc_efficient_df["letter"]=["A","B","C","D","E"]
p9.ggplot()+\
    p9.coord_equal()+\
    p9.geom_line(
        p9.aes(
            x="FPR",
            y="TPR",
        ),
        data=roc_efficient_df
    )+\
    p9.geom_text(
        p9.aes(
            x="FPR+0.05",
            y="TPR+0.05",
            label="letter",
        ),
        data=roc_efficient_df
    )+\
    p9.geom_point(
        p9.aes(
            x="FPR",
            y="TPR",
        ),
        data=roc_efficient_df
    )
```

The AUM loss is defined as the area under the minimum of False Positive and False Negative rates, which are the same data we used to compute the ROC curve, and can be visualized using the code below:

```{python}
roc_long = pd.melt(
    roc_efficient_df,
    value_vars=["FPR","FNR","min(FPR,FNR)"],
    id_vars=["min_constant","max_constant"])
roc_efficient_df["xtext"]=np.where(
    roc_efficient_df["min_constant"] == -np.inf,
    roc_efficient_df["max_constant"]-1,
    np.where(
        roc_efficient_df["max_constant"] == np.inf,
        roc_efficient_df["min_constant"]+1,
        (roc_efficient_df["min_constant"]+roc_efficient_df["max_constant"])/2))
gg = p9.ggplot()+\
    p9.theme_bw()+\
    p9.geom_rect(
        p9.aes(
            xmin="min_constant",
            xmax="max_constant",
            ymin="0",
            ymax="min(FPR,FNR)"
        ),
        data=roc_efficient_df)+\
    p9.geom_vline(
        p9.aes(
            xintercept="min_constant",
        ),
        color="grey",
        data=roc_efficient_df)+\
    p9.geom_text(
        p9.aes(
            x="xtext",
            y="0.75",
            label="letter"
        ),
        data=roc_efficient_df)+\
    p9.geom_segment(
        p9.aes(
            x="min_constant",
            xend="max_constant",
            y="value",
            yend="value",
            color="variable",
            size="variable"
        ),
        data=roc_long)+\
    p9.scale_color_manual(
        values={
            "FPR":"red",
            "FNR":"deepskyblue",
            "min(FPR,FNR)":"black"
        })+\
    p9.scale_size_manual(
        values={
            "FPR":3,
            "FNR":5,
            "min(FPR,FNR)":1,
        })+\
    p9.scale_x_continuous("Constant added to predicted values")+\
    p9.scale_y_continuous("value")
gg
show(gg, "err_funs")
```


ROC curve optimizationBelow I wrote an implementation of
our AUM loss function,

```{python}
def Proposed_AUM(pred_tensor, label_tensor):
    """Area Under Min(FP,FN)

    Differentiable loss function for imbalanced binary classification
    problems. Minimizing AUM empirically results in maximizing Area
    Under the ROC Curve (AUC). Arguments: pred_tensor and label_tensor
    should both be 1d tensors, vectors of real-valued predictions and
    integer labels for each observation in the set/batch. Labels
    should be either 1 for positive class or any other value for
    negative class.

    """
    roc = ROC_curve(pred_tensor, label_tensor)
    return torch.sum(roc["min"][1:-1] * roc["min_thresh"][1:].diff())

label_vec = torch.tensor([0,0,1,1])
pred_vec = torch.tensor([1.0, -1, 1, 0])
pred_vec = torch.tensor([1.0, 2, 3, 4])
Proposed_AUM(pred_vec, label_vec)
ROC_AUC(pred_vec, label_vec)

```

The implementation above consists of vectorized torch operations, all
of which should be differentiable almost everywhere. To check the
automatically computed gradient from torch with the directional
derivatives described in our paper, we can use the backward method,

```{python}
label_vec = [0, 1]
pred_diff_vec = torch.arange(-2, 2, 0.5)
obj_grad_df_list = []
for pred_diff in pred_diff_vec:
    pred_vec = [0, pred_diff]
    for objective in "Proposed_AUM","ROC_AUC":
        pred_tensor = torch.tensor(pred_vec)
        pred_tensor.requires_grad = True
        label_tensor = torch.tensor(label_vec)
        ofun = eval(objective)
        loss = ofun(pred_tensor, label_tensor)
        try:
            loss.backward()
            g_vec = pred_tensor.grad
        except:
            g_vec = torch.tensor([0.0, 0])
        obj_grad_df_list.append(pd.DataFrame({
            "pred_diff":pred_diff.numpy(),
            "objective":objective,
            "function":["output","deriv0","deriv1"],
            "value":torch.cat([loss.reshape(1),g_vec]).detach().numpy(),
        }))
obj_grad_df = pd.concat(obj_grad_df_list)

p9.ggplot()+\
    p9.theme_bw()+\
    p9.geom_hline(
        p9.aes(
            yintercept="value"
        ),
        data=pd.DataFrame({"value":[0]}),
        color="grey"
    )+\
    p9.geom_point(
        p9.aes(
            x="pred_diff",
            y="value",
        ),
        data=obj_grad_df
    )+\
    p9.facet_grid("function ~ objective", labeller="label_both")
    
```

## Simple differentiable points

Let us consider an example from the aum R package,

```{r}
bin.diffs <- aum::aum_diffs_binary(c(0,1))
aum::aum(bin.diffs, c(-10,10))
```

The R code and output above shows that for one negative label with
predicted value -10, and one positive label with predicted value 10,
we have AUM=0 and directional derivatives are also zero. That is also
seen in the python code below:

```{python}
y_list = [0, 1]
def loss_grad(pred_vec, label_vec):
    pred_tensor = torch.tensor(pred_vec)
    pred_tensor.requires_grad = True
    label_tensor = torch.tensor(label_vec)
    loss = AUM(pred_tensor, label_tensor)
    loss.backward()
    return loss, pred_tensor.grad
loss_grad([-10.0, 10.0], y_list)
```

Another example is for the same labels but the opposite predicted
values, for which the R code is below,

```{r}
aum::aum(bin.diffs, c(10,-10))
```

The R code and output above shows that we have AUM=20 and derivative 1
for the first/negative example, and derivative -1 for the
second/positive example. Those derivatives indicate that the AUM can
be decreased by decreasing the predicted value for the first/negative
example and/or increasing the predicted vale for the second/positive
example. Consistent results can be observed from the python code
below,

```{python}
loss_grad([10.0, -10.0], y_list)
```

## Non-differentiable points

As discussed in our paper, the AUM is not differentiable everywhere.
So what happens when you use auto-grad on a non-differentiable loss?
Apparently the backward method returns a subgradient, as explained in
[an issue discussing autograd of L1 loss
function](https://github.com/pytorch/pytorch/issues/7172).

One example, again taken from the aum R package, is when both
predicted values are zero,

```{r}
aum::aum(bin.diffs, c(0,0))
```

The output above indicates AUM=0 with directional derivatives which
are not equal on the left and right. The first row of the directional
derivative matrix says that decreasing the first/negative predicted
value will result in no change to AUM, whereas increasing will result
in increasing AUM. The second row says that decreasing the
second/positive predicted value results in increasing AUM, whereas
increasing the predicted value results in no change to AUM. Is the
python torch auto-grad code consistent?

```{python}
loss_grad([0.0, 0.0], y_list)
```

The output above says that the gradient is zero, which is OK (these
predicted values are a minimum) but it is missing some information
about what happens nearby.

## Non-convex points

The implementation above of AUM is only for binary classification, but
in our paper we also discuss an application to changepoint detection
problems. In the case of changepoint detection problems with
non-monotonic error functions, the AUM may be non-convex, and there
may be points at which there are no subgradients. Some are used as
[test cases in our R package](https://github.com/tdhock/aum/blob/main/tests/testthat/test-CRAN.R).
What does auto-grad return in that case? (exercise for the reader)

Update Aug 2024. The `AUM` function above uses the minimum of total FP and FN, whereas there now is a more complete python AUM implementation which can use rates (FPR/FNR) rather than totals.

* [rate only](https://github.com/tdhock/max-generalized-auc/blob/master/data_Classif.py#L41)
* [count total or rate](https://github.com/tdhock/max-generalized-auc/blob/master/figure-aum-neural-networks-data.py#L83)


## Conclusions

TODO

## Session info

```{r}
sessionInfo()
```
