---
layout: post
title: New code for various kinds of cross-validation
description: Cross-validation in R with mlr3
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2024-04-15-cv-all-same-new"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=8,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
options(width=80)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this blog post is to explore machine learning in the
context of relatively small data, which may not be large enough
(relative to the strength of the signal) to learn the pattern. In that
case, we may wish to combine our small data set with another data set,
to see if the learning algorithm works better.  For example, in a
collaboration with Bruce Hungate and Jeff Propster of NAU ECOSS, we
have several qSIP data sets (quantitative Stable Isotope Probing),
from different experiments/sites/etc, and we would like to know if we
can combine them, and learn a more accurate model, than if we learned
separate models for each subset. Here we explore what kinds of data
could result in better learning, when we combine data sets.

## Simulation

In [a previous
blog](https://tdhock.github.io/blog/2023/variable-size-train/) we
explained how to use `mlr3resampling::ResamplingVariableSizeTrain`,
with simulated regression problems. Here we adapt one of those
simulations, of data with one feature and and output/label with a sin
pattern.

```{r simulationScatter}
N <- 300
abs.x <- 20
set.seed(1)
x.vec <- runif(N, -abs.x, abs.x)
str(x.vec)
library(data.table)
(task.dt <- data.table(
  x=x.vec,
  y = sin(x.vec)+rnorm(N,sd=0.5)))
reg.task <- mlr3::TaskRegr$new(
  "sin", task.dt, target="y"
)
library(animint2)
ggplot()+
  geom_point(aes(
    x, y),
    shape=1,
    data=task.dt)
```

Above we see a scatterplot of the simulated data. The goal of the
learning algorithm will be to predict y from x.

### Visualizing test error as a function of train set size

In the code below, we define a K-fold cross-validation experiment,
with 

* K=2 `folds`,
* a `ratio` of 0.9 between the train set sizes,
* and 19 `sizes` of train sets smaller than the full size.

```{r VariableSizeTrainCV}
reg_size_cv <- mlr3resampling::ResamplingSameOtherSizesCV$new()
reg_size_cv$param_set$values$folds <- 2
reg_size_cv$param_set$values$ratio <- 0.9
reg_size_cv$param_set$values$sizes <- 19
reg_size_cv$param_set$values$seeds <- 3
reg_size_cv$instantiate(reg.task)
reg_size_cv$instance$iteration.dt
```

The table above shows that there is a CV iteration/split to compute
for each train set size (20 to 150), random seeds (1 to 3), and
cross-validation fold (1 or 2).

```{r}
(reg.learner.list <- list(
  if(requireNamespace("rpart"))mlr3::LearnerRegrRpart$new(),
  mlr3::LearnerRegrFeatureless$new()))
(reg.bench.grid <- mlr3::benchmark_grid(
  reg.task,
  reg.learner.list,
  reg_size_cv))
if(require(future))plan("multisession")
lgr::get_logger("mlr3")$set_threshold("warn")
(reg.bench.result <- mlr3::benchmark(
  reg.bench.grid, store_models = TRUE))
reg.bench.score <- mlr3resampling::score(reg.bench.result)
reg.bench.score[1]
```

The output above includes one row of the resulting scores.
The plot below shows test error for several train sizes.

```{r train_size_err}
(train_size_vec <- unique(reg.bench.score$n.train.groups))
ggplot()+
  scale_x_log10(
    breaks=train_size_vec)+
  scale_y_log10(
    "Mean squared error on test set")+
  geom_line(aes(
    n.train.groups, regr.mse,
    subset=paste(algorithm, seed),
    color=algorithm),
    shape=1,
    data=reg.bench.score)+
  geom_point(aes(
    n.train.groups, regr.mse, color=algorithm),
    shape=1,
    data=reg.bench.score)+
  facet_grid(
    test.fold~.,
    labeller=label_both,
    scales="free")
```

Above we plot the test error for each fold and train set size. We can
see that if there are less than about 80 samples, the rpart
learning algorithm has about the same error as featureless
(representing the error rate of no relationship learned between
input/output). So if we have subsets of data that really are the same,
but each is smaller than this critical threshold for this problem,
then rpart will not be able to learn anything unless we put the
subsets together. Let's try it!

## Simulating and combining identical subsets

The code below assigns three test subsets to the randomly simulated
data.

```{r}
task.dt[, random_subset := rep(1:3, l=.N)][]
table(subset.tab <- task.dt$random_subset)
```

The output above shows that 
there are an equal number (`r subset.tab[1]`) 
of each subset. 
Below we define a task using that subset,

```{r}
subset.task <- mlr3::TaskRegr$new(
  "sin", task.dt, target="y")
subset.task$col_roles$stratum <- "random_subset"
subset.task$col_roles$subset <- "random_subset"
subset.task$col_roles$feature <- "x"
str(subset.task$col_roles)
```

Below we define cross-validation using two folds, as above.

```{r}
same_other_cv <- mlr3resampling::ResamplingSameOtherCV$new()
same_other_cv$param_set$values$folds <- 2
```

So using the 2-fold cross-validation, half of the data in each subset
will be assigned to train, and the other half to test. So each "same"
train set will have only 50 samples (for the same subset), each "other"
train set will have 100 samples, and each "all" train set will have
150 samples. In the previous section, we saw that rpart needs at least
100 samples to learn, so we should expect that "same" is just as
inaccurate as featureless, "other" is a bit better, and "all" is
better still.
We compute and plot the results using the code below,

```{r SameOtherCV}
(same.other.grid <- mlr3::benchmark_grid(
  subset.task,
  reg.learner.list,
  same_other_cv))
(same.other.result <- mlr3::benchmark(
  same.other.grid, store_models = TRUE))
same.other.score <- mlr3resampling::score(same.other.result)
same.other.score[1]

ggplot()+
  geom_point(aes(
    regr.mse, train.subsets, color=algorithm),
    data=same.other.score)+
  facet_grid(. ~ test.subset, labeller=label_both, scales="free")+
  scale_x_log10(
    "Mean squared prediction error (test set)")

same.other.wide <- dcast(
  same.other.score,
  test.subset + train.subsets + algorithm ~ .,
  list(mean,sd),
  value.var="regr.mse")
ggplot()+
  geom_segment(aes(
    regr.mse_mean+regr.mse_sd, train.subsets,
    xend=regr.mse_mean-regr.mse_sd, yend=train.subsets,
    color=algorithm),
    data=same.other.wide)+
  geom_point(aes(
    regr.mse_mean, train.subsets, color=algorithm),
    data=same.other.wide)+
  facet_grid(. ~ test.subset, labeller=label_both, scales="free")+
  scale_x_log10(
    "Mean squared prediction error (test set)")
```

The figure above shows a test subset in each panel, the train subsets on
the y axis, the test error on the x axis, the two different algorithms
are shown in two different colors. We can clearly see that 

* For `train.subsets=same`, rpart has about the same test error as
  featureless, which indicates that nothing has been learned (not
  enough data).
* For `train.subsets=other`, rpart test error is significantly smaller
  than featureless, indicating that some non-trivial relationship
  between inputs and outputs has been learned.
* For `train.subsets=all`, rpart test error is smaller still, which
  indicates that combining all of the subsets is beneficial in this
  case (when the pattern is exactly the same in the different
  subsets).
  
## Conclusions

We have shown how to use `mlr3resampling` to determine the number of
train samples which is required to get non-trivial prediction
accuracy. In the simulation above, that number was about 100 train
samples. We then defined subsets and cross-validation such that there
were only 50 train samples per subset, so training on just samples from
the same subset is not enough (at least for the rpart learning
algorithm). We observed smaller test error rates when training on
other/all subsets (larger train sets, all with the same
distribution). Overall this is a convincing demonstration that it is
possible for "other" and "all" to be more accurate than "same" --- for
the case where the subsets really do have an identical pattern. If in
real data we see the opposite (same is the best), then this implies
that there is a different pattern to learn in each subset.

## Session info

```{r}
sessionInfo()
```

## UPDATE 15 Apr 2024

Code above uses the new version 2024.4.15 of mlr3resampling, in which
we now use `subset` instead of `group`, for consistency with the usage
of `group` in other `mlr3` packages. See [old blog
post](https://tdhock.github.io/blog/2024/when-is-all-better-than-same/)
for a discussion about how to do the analyses above using the old
classes (not recommended).
