---
layout: post
title: Are pipe operations linear or quadratic?
description: A demonstration of atime on mlr3torch
---

```{r Ropts, echo=FALSE, results='hide'}
repo.dir <- normalizePath("..")
post.id <- "2025-02-19-linear-vs-quadratic-pipe-ops"
fig.path <- paste0(file.path(repo.dir, "assets", "img", post.id), "/")
dir.create(fig.path, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=10, ## TODO python figures wider? look at prev issue.
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
conda.env <- "2023-08-deep-learning"
conda.env <- "torch-aum"
RETICULATE_PYTHON <- sprintf(if(.Platform$OS.type=="unix")
  ##"/home/tdhock/.local/share/r-miniconda/envs/%s/bin/python"
  "/home/tdhock/miniconda3/envs/%s/bin/python"
  else "~/AppData/Local/Miniconda3/envs/%s/python.exe", conda.env)
Sys.setenv(RETICULATE_PYTHON=RETICULATE_PYTHON)
##reticulate::use_condaenv(dirname(RETICULATE_PYTHON), required=TRUE)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to explain how to use
[atime](https://github.com/tdhock/atime), my R package for asymptotic
benchmarking, to determine the asymptotic complexity of pipe
operations in mlr3.

## Background: mlr3pipelines man page

mlr3 is a framework for machine learning in R.  One related package is
mlr3pipelines, which makes it easy to create "pipelines" of operations
related to machine learning model training (feature selection,
etc). To create a pipeline we use the `%>>%` operator; its documentation on
`help("%>>%",package="mlr3pipelines")` says

```
%>>% always creates deep copies of its input arguments, so they
cannot be modified by reference afterwards. To access individual
‘PipeOp’s after composition, use the resulting ‘Graph’'s $pipeops
list. %>>!%, on the other hand, tries to avoid cloning its first
argument: If it is a ‘Graph’, then this ‘Graph’ will be modified
in-place.

When %>>!% fails, then it leaves ‘g1’ in an incompletely modified
state. It is therefore usually recommended to use %>>%, since the
very marginal gain of performance from using %>>!% often does not
outweigh the risk of either modifying objects by-reference that
should not be modified or getting graphs that are in an
incompletely modified state. However, when creating long ‘Graph’s,
chaining with %>>!% instead of %>>% can give noticeable
performance benefits because %>>% makes a number of
‘clone()’-calls that is quadratic in chain length, %>>!% only
linear.
```

The man page above indicates that there are two operators which can be
used to create a pipeline:

* `%>>%` copies its arguments and is supposed to be quadratic time,
* `%>>!%` avoids copies and is supposed to be linear time.

The goal of this post is to use atime to verify these claims.

## Test

First we list available pipe ops:

```{r}
mlr3pipelines::po()
```

Below, we combine a few instances of the first operation shown above:

```{r}
po_list <- list(
  mlr3pipelines::po("adas_1"),
  mlr3pipelines::po("adas_2"))
Reduce(mlr3pipelines::`%>>%`, po_list)
Reduce(mlr3pipelines::`%>>!%`, po_list)
```

We see above that the two results are consistent.  Note that we use
`Reduce` with a list of pipe operations, to avoid attaching,
`library(mlr3pipelines)` (easier to see which objects are defined in
which packages).

## Verification

One way to test would be to use binary operators, as below:

```{r}
FUN_names <- paste0("%>>",c("","!"),"%")
FUN_list <- lapply(FUN_names, getFromNamespace, "mlr3pipelines")
names(FUN_list) <- FUN_names
(expr.list.binary <- atime::atime_grid(
  list(FUN=FUN_names),
  binary=Reduce(FUN_list[[FUN]], po_list)))
```

Another way would be to use the `concat_graphs` function, as below:

```{r}
(expr.list.concat <- atime::atime_grid(
  list(in_place=c(TRUE,FALSE)),
  concat=Reduce(
    function(x,y)mlr3pipelines::concat_graphs(x,y,in_place=in_place),
    po_list)))
```

We can test both by combining the lists in the code below:

```{r plot-atime}
atime_list <- atime::atime(
  setup={
    po_list <- lapply(
      paste0("adas_", 1:N),
      mlr3pipelines::po)
  },
  expr.list=c(expr.list.binary, expr.list.concat),
  seconds.limit=1)
plot(atime_list)+
  ggplot2::scale_x_log10(
    breaks=2^seq(1,5),
    limits=c(2,100))
```

From the figure above, we see that there are two different slopes on the log-log plot:

* `binary FUN=%>>%` and `concat in_place=FALSE` have larger slope.
* `binary FUN=%>>!%` and `concat in_place=TRUE` have smaller slope.

We estimate the asymptotic complexity class via the code below,


```{r plot-refs}
atime_refs <- atime::references_best(atime_list)
plot(atime_refs)
```

The figure above suggests that

* `binary FUN=%>>%` and `concat in_place=FALSE` have quadratic `N^2` time complexity
* `binary FUN=%>>!%` and `concat in_place=TRUE` have linear `N` time complexity,

where `N` is the number of operations in the pipeline.

## Conclusions

We have used atime to verify the linear/quadratic time complexity
claims in the mlr3pipelines package.

## Session info

```{r}
sessionInfo()
```
