---
layout: post
title: Worst case for change-point pruning methods
description: Pathological data examples
---

```{r Ropts, echo=FALSE, results='hide'}
repo.dir <- normalizePath("..")
post.id <- "2025-04-16-PELT-vs-FPOP-poisson-worst"
fig.path <- paste0(file.path(repo.dir, "assets", "img", post.id), "/")
dir.create(fig.path, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=10, ## TODO python figures wider? look at prev issue.
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
conda.env <- "2023-08-deep-learning"
conda.env <- "torch-aum"
RETICULATE_PYTHON <- sprintf(if(.Platform$OS.type=="unix")
  ##"/home/tdhock/.local/share/r-miniconda/envs/%s/bin/python"
  "/home/tdhock/miniconda3/envs/%s/bin/python"
  else "~/AppData/Local/Miniconda3/envs/%s/python.exe", conda.env)
Sys.setenv(RETICULATE_PYTHON=RETICULATE_PYTHON)
##reticulate::use_condaenv(dirname(RETICULATE_PYTHON), required=TRUE)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to compare two pruning methods for speeding
up the optimal partitioning algorithm, similar to my previous posts:

* [Poisson loss, average case](https://tdhock.github.io/blog/2025/PELT-vs-FPOP/)
* [Square loss, average case](https://tdhock.github.io/blog/2025/PELT-vs-fpopw/)
* [Square loss, worst case](https://tdhock.github.io/blog/2025/PELT-vs-fpopw-worst/)

* We assume that we want to find change-points in a sequence of `N` data.
* All algorithms discussed are instances of dynamic programming, which
  has a for loop over all `N` data. In each iteration, the algorithm
  considers a certain number of candidate change-points `C` (on
  average). The algorithm is overall `O(NC)` time.
* Optimal partitioning (OPART) computes the best segmentation for a
  given loss function, data sequence, and non-negative penalty
  value. Each iteration considers the loss for each previous candidate
  change-point, `C=O(N)` linear time per iteration, which implies
  quadratic `O(N^2)` time overall.
* Pruned Exact Linear Time (PELT) reduces the number of candidate
  change-points that must be considered at each iteration, from the
  whole data sequence size `N`, to the size of a segment `S`, so
  `C=O(S)`. This gives an algorithm which is `O(NS)` time overall. The
  worst case of this algorithm is a data set with no changes, for
  which it has the same quadratic `O(N^2)` time complexity as the
  classic OPART algorithm.
* Functional Pruning Optimal Partitioning (FPOP) reduces the number of
  candidate change-points to a number which is always less than the
  number considered by PELT (see Maidstone 2017 paper for detailed
  proof). The goal of this post is to show that it prunes more than
  PELT, even in the worst case.
  
## PELT

Below we define a function which implements PELT for the Poisson loss.

```{r}
PELT <- function(sim.mat, penalty, prune=TRUE, verbose=FALSE){
  if(!is.matrix(sim.mat))sim.mat <- cbind(sim.mat)
  cum.data <- rbind(0, apply(sim.mat, 2, cumsum))
  sum_trick <- function(m, start, end)m[end+1,,drop=FALSE]-m[start,,drop=FALSE]
  cost_trick <- function(start, end){
    sum_vec <- sum_trick(cum.data, start, end)
    N <- end+1-start
    mean_vec <- sum_vec/N
    ifelse(sum_vec==0, 0, sum_vec*(1-log(mean_vec)))
  }
  N.data <- nrow(sim.mat)
  pelt.change.vec <- rep(NA_integer_, N.data)
  pelt.cost.vec <- rep(NA_real_, N.data+1)
  pelt.cost.vec[1] <- -penalty
  pelt.candidates.vec <- rep(NA_integer_, N.data)
  candidate.vec <- 1L
  if(verbose)index_dt_list <- list()
  for(up.to in 1:N.data){
    N.cand <- length(candidate.vec)
    pelt.candidates.vec[up.to] <- N.cand
    last.seg.cost <- rowSums(cost_trick(candidate.vec, rep(up.to, N.cand)))
    prev.cost <- pelt.cost.vec[candidate.vec]
    cost.no.penalty <- prev.cost+last.seg.cost
    total.cost <- cost.no.penalty+penalty
    if(verbose)index_dt_list[[up.to]] <- data.table(
      data_i=up.to, change_i=candidate.vec, cost=total.cost/up.to)
    best.i <- which.min(total.cost)
    pelt.change.vec[up.to] <- candidate.vec[best.i]
    total.cost.best <- total.cost[best.i]
    pelt.cost.vec[up.to+1] <- total.cost.best
    keep <- if(isTRUE(prune))cost.no.penalty < total.cost.best else TRUE
    candidate.vec <- c(candidate.vec[keep], up.to+1L)
  }
  list(
    change=pelt.change.vec,
    cost=pelt.cost.vec,
    candidates=pelt.candidates.vec,
    index_dt=if(verbose)rbindlist(index_dt_list))
}
decode <- function(best.change){
  seg.dt.list <- list()
  last.i <- length(best.change)
  while(last.i>0){
    first.i <- best.change[last.i]
    seg.dt.list[[paste(last.i)]] <- data.table(
      first.i, last.i)
    last.i <- first.i-1L
  }
  rbindlist(seg.dt.list)[seq(.N,1)]
}
```

# Previous worst case demonstration

The PeakSegDisk R package contains a demonstration of the worst case
time complexity in the [Worst case
vignette](https://cloud.r-project.org/web/packages/PeakSegDisk/vignettes/Worst_case.html). Here we adapt that code to atime:

```{r}
cache.rds <- "2025-04-17-PELT-vs-FPOP-poisson-worst.rds"
if(file.exists(cache.rds)){
  atime_list <- readRDS(cache.rds)
}else{
  data(Mono27ac, package="PeakSegDisk", envir=environment())
  library(data.table)
  make_data <- function(some.cov)list(
    real=some.cov,
    constant=data.table(some.cov)[, count := rpois(.N, 100)],
    decreasing=data.table(some.cov)[, count := .N:1],
    increasing=data.table(some.cov)[, count := 1:.N])
  expr.list <- atime::atime_grid(list(
    penalty=c(0.01, 1e6),
    data_type=names(make_data(Mono27ac$coverage))),
    FPOP={
      df <- data.list[[data_type]]
      ufit <- PeakSegOptimal::UnconstrainedFPOP(df$count, penalty=penalty)
      with(ufit, data.table(
        segments=sum(is.finite(mean.vec)),
        mean.intervals=mean(intervals.vec)))
    },
    PELT={
      df <- data.list[[data_type]]
      pfit <- PELT(df$count, penalty)
      seg_dt <- decode(pfit$change)
      data.table(
        segments=nrow(seg_dt),
        mean.intervals=mean(pfit$candidates))
    },
    PeakSeg={
      df <- data.list[[data_type]]
      fit <- PeakSegDisk::PeakSegFPOP_df(df, penalty)
      fit$loss[, .(segments, mean.intervals)]
    })
  set.seed(1)
  atime_list <- atime::atime(
    N=as.integer(10^seq(1, log10(nrow(Mono27ac$coverage)), by=0.2)),
    result=TRUE,
    expr.list=expr.list,
    seconds.limit = 1,
    setup={
      data.list <- make_data(Mono27ac$coverage[1:N])
    })
  saveRDS(atime_list, cache.rds)
}
```

```{r}
atime_refs <- atime::references_best(atime_list)
algo.colors <- c(
  OPART="grey50",
  PeakSeg="orange",
  PELT="red",
  FPOP="blue")
library(ggplot2)
ggplot()+
  geom_line(aes(
    N, empirical,
    color=expr.grid,
    linetype=factor(penalty)),
    data=atime_refs$measurements)+
  facet_grid(unit ~ data_type, labeller=label_both, scales="free")+
  scale_x_log10()+
  scale_y_log10()+
  scale_color_manual(values=algo.colors)
```

# Simulate data sequences

We simulate data in two scenarios: no changes, and always decreasing.
```{r}
sim_fun_list <- list(
  no_changes=function(N){
    rep(5000, N)
  },
  increasing_linear=function(N){
    seq(1, N)*100
  },
  increasing_quadratic=function(N){
    seq(1, N)^2
  })
lapply(sim_fun_list, function(f)f(10))
```

As can be seen above, both functions return a vector of values that
represent the true segment mean.
Below we use both functions with a three different data sizes.

```{r}
library(data.table)
N_data_vec <- c(10, 100)
sim_data_list <- list()
sim_changes_list <- list()
for(N_data in N_data_vec){
  for(simulation in names(sim_fun_list)){
    sim_fun <- sim_fun_list[[simulation]]
    data_mean_vec <- sim_fun(N_data)
    end <- which(diff(data_mean_vec) != 0)
    set.seed(1)
    data_value <- rpois(N_data, data_mean_vec)
    sim_data_list[[paste(N_data, simulation)]] <- data.table(
      N_data, simulation, data_i=seq_along(data_value), data_value)
    sim_changes_list[[paste(N_data, simulation)]] <- data.table(
      N_data, simulation, end)
  }
}
addSim <- function(DT)DT[, Simulation := paste0("\n", simulation)][]
(sim_changes <- addSim(rbindlist(sim_changes_list)))
```

Above we see the table of simulated change-points. 

* For `constant_changes` simulation, there are always 3 change-points.
* For `linear_changes` simulation, there are more change-points when
  there are more data.
  
Below we visualize the simulated data.

```{r sim-data}
(sim_data <- addSim(rbindlist(sim_data_list)))
library(ggplot2)
ggplot()+
  theme_bw()+
  theme(text=element_text(size=14))+
  geom_point(aes(
    data_i, data_value),
    color="grey50",
    data=sim_data)+
  facet_grid(Simulation ~ N_data, labeller=label_both, scales="free_x", space="free")+
  scale_x_continuous(
    breaks=seq(0,max(N_data_vec),by=5))
```

We see in the figure above that the data are quite different.

```{r}
pelt_info_list <- list()
pelt_segs_list <- list()
both_index_list <- list()
penalty <- 100
algo.prune <- c(
  OPART=FALSE,
  PELT=TRUE)
for(N_data in N_data_vec){
  for(simulation in names(sim_fun_list)){
    N_sim <- paste(N_data, simulation)
    data_value <- sim_data_list[[N_sim]]$data_value
    for(algo in names(algo.prune)){
      prune <- algo.prune[[algo]]
      fit <- PELT(data_value, penalty=penalty, prune=prune, verbose = TRUE)
      both_index_list[[paste(N_data, simulation, algo)]] <- data.table(
        N_data, simulation, algo, fit$index_dt)
      fit_seg_dt <- decode(fit$change)
      pelt_info_list[[paste(N_data,simulation,algo)]] <- data.table(
        N_data,simulation,algo,
        candidates=fit$candidates,
        cost=fit$cost[-1],
        data_i=seq_along(data_value))
      pelt_segs_list[[paste(N_data,simulation,algo)]] <- data.table(
        N_data,simulation,algo,
        fit_seg_dt)
    }
  }
}
(pelt_info <- addSim(rbindlist(pelt_info_list)))
(pelt_segs <- addSim(rbindlist(pelt_segs_list)))
```

We see in the result tables above that the segmentations are the same,
using pruning and no pruning. Below we visualize the number of
candidates considered.

```{r pelt-prune}
algo.colors <- c(
  OPART="grey50",
  PELT="red",
  FPOP="blue")
ggplot()+
  theme_bw()+
  theme(
    panel.spacing=grid::unit(1,"lines"),
    text=element_text(size=15))+
  scale_color_manual(
    breaks=names(algo.colors),
    values=algo.colors)+
  geom_point(aes(
    data_i, candidates, color=algo),
    data=pelt_info)+
  facet_grid(Simulation ~ N_data, labeller=label_both, scales="free_x", space="free")+
  scale_x_continuous(
    breaks=seq(0,max(N_data_vec),by=5))+
  scale_y_continuous(
    breaks=c(10, 100, 400))+
  theme(panel.grid.minor=element_blank())+
  coord_cartesian(expand=FALSE)
```

We can see in the figure above that PELT considers a much smaller
number of change-points, that tends to reset near zero, for each
change-point detected. 

* In the bottom simulation with linear changes, the number of
  change-point candidates considered by PELT is constant (does not
  depend on number of data), so PELT is linear time, whereas OPART is
  quadratic, in the number of data.
* In the top simulation with constant changes, the number of
  change-point candidates considered by PELT is linear in the number
  of data, so both OPART and PELT are quadratic time in the number of
  data.

## FPOP

Now we run FPOP, which is another pruning method, which is more
complex to implement efficiently, so we use C++ code
in the R package fpopw.

```{r fpop-prune}
if(FALSE){
  remotes::install_github("TODO")
}
fpop_info_list <- list()
fpop_segs_list <- list()
for(N_data in N_data_vec){
  for(simulation in names(sim_fun_list)){
    N_sim <- paste(N_data, simulation)
    data_value <- sim_data_list[[N_sim]]$data_value
    pfit <- PeakSegOptimal::UnconstrainedFPOP(
      data_value, penalty=penalty, verbose_file=tempfile())
    both_index_list[[paste(N_data, simulation, "FPOP")]] <- data.table(
      N_data, simulation, algo="FPOP", pfit$index_dt
    )[, let(
      data_i=data_index+1L,
      change_i=change+2L
    )][]
    end <- pfit$t.est
    start <- c(1, end[-length(end)]+1)
    fpop_info_list[[paste(N_data,simulation)]] <- data.table(
      N_data,simulation,
      algo="FPOP",
      candidates=pfit$intervals,
      cost=pfit$cost,
      data_i=seq_along(data_value))
    fpop_segs_list[[paste(N_data,simulation)]] <- data.table(
      N_data,simulation,start,end)
  }
}
(fpop_info <- addSim(rbindlist(fpop_info_list)))
(fpop_segs <- rbindlist(fpop_segs_list))
both_info <- rbind(pelt_info, fpop_info)
ggplot()+
  theme_bw()+
  theme(
    panel.spacing=grid::unit(1,"lines"),
    text=element_text(size=15))+
  scale_color_manual(
    breaks=names(algo.colors),
    values=algo.colors)+
  geom_point(aes(
    data_i, candidates, color=algo),
    data=both_info)+
  facet_grid(Simulation ~ N_data, labeller=label_both, scales="free_x", space="free")+
  scale_x_continuous(
    breaks=seq(0,max(N_data_vec),by=5))+
  scale_y_continuous(
    breaks=c(10, 100, 400))+
  theme(panel.grid.minor=element_blank())+
  coord_cartesian(expand=FALSE)
```

In the figure above, we can see the advantage of FPOP: the number of
change-points considered does not increase with the number of data,
even with constant number of changes (larger segments).

* In the top simulation with constant changes, the number of
  change-points candidates considered is linear in the number of data,
  so both OPART and PELT are quadratic time in the number of data,
  whereas FPOP is sub-quadratic (empirically linear or log-linear).
  
The table below compares the cost and number of candidates for the last data point.

```{r}
both_info[data_i==max(data_i)][order(simulation, algo)]
```

We see above that in both simulations, all three algos compute the same cost. 
We also see that the number of candidates is smallest for FPOP in both simulations.
PELT has substantial pruning (10 candidates) for the case of linear changes, 
but not much pruning (100 candidates) for the case of constant changes.

## Vary penalty, fix data size

```{r}
fpop_summary_list <- list()
N <- 100
simulation <- "increasing_linear"
for(penalty in 10^seq(-3,10)){
  N_sim <- paste(N_data, simulation)
  data_value <- sim_data_list[[N_sim]]$data_value
  pfit <- PeakSegOptimal::UnconstrainedFPOP(
    data_value, penalty=penalty, verbose_file=tempfile())
  both_index_list[[paste(N_data, simulation, "FPOP")]] <- data.table(
    N_data, simulation, algo="FPOP", pfit$index_dt
  )[, let(
    data_i=data_index+1L,
    change_i=change+2L
  )][]
  fpop_summary_list[[paste(penalty)]] <- data.table(
    penalty,
    segments=sum(is.finite(pfit$mean.vec)),
    mean.candidates=mean(pfit$intervals))
}
(fpop_summary <- rbindlist(fpop_summary_list))
```

## Heat maps
  
Another way to view this is by looking at the cost of each candidate
change-point considered, as in the heat map below.

```{r cost-heat}
algo.levs <- c("OPART","PELT","FPOP")
(both_index <- addSim(rbindlist(both_index_list, use.names=TRUE))[, let(
  Algorithm = factor(algo, algo.levs)
)][])
for(a in algo.levs){
  gg <- ggplot()+
    ggtitle(a)+
    theme_bw()+
    theme(text=element_text(size=20))+
    coord_equal()+
    geom_tile(aes(
      data_i, change_i, fill=cost),
      data=both_index[Algorithm==a])+
    scale_fill_gradient(low="black", high="red")+
    facet_grid(Simulation ~ N_data, label=label_both)
  print(gg)
}
```

Another way to visualize it is in the plot below, which super-imposes
the three algos.

```{r candidates-compare}
ggplot()+
  theme_bw()+
  theme(text=element_text(size=15))+
  geom_tile(aes(
    data_i, change_i, fill=Algorithm),
    data=both_index)+
  scale_fill_manual(values=algo.colors)+
  scale_x_continuous(
    breaks=seq(0,max(N_data_vec),by=20))+
  facet_grid(Simulation ~ N_data, label=label_both, scales="free", space="free")
```

It is clear that FPOP prunes much more than PELT, especially in the
case of constant changes (segments that get larger with the overall
number of data).

## Double check FPOP for constant changes

Before running a simulation on varying data sizes, here we check that
the penalty value results in the right number of change-points, for a
larger data set.

```{r}
N_data <- 1e5
sim_fun <- sim_fun_list$constant_changes
data_mean_vec <- sim_fun(N_data)
set.seed(1)
data_value <- rnorm(N_data, data_mean_vec, 2)
pfit <- fpopw::Fpop(data_value, penalty)
pfit$t.est
```

The result above shows that there are four segments detected (three
change-points), which is the expected number in our "constant changes"
simulation.
  
## atime comparison

The `atime()` function can be used to perform asymptotic time/memory/etc comparisons. 
This means that we will increase N, and monitor how fast certain quantities grow with N.
We begin by defining the data sizes N of interest:

```{r}
base_N <- c(100,200,400,800)
(all_N <- unlist(lapply(10^seq(0,3), function(x)x*base_N)))
```

The data sizes above are on a log scale between 10 and 1,000,000.
Next, we define a list that enumerates the different combinations in the experiment.

```{r}
grid_args <- list(
  list(simulation=names(sim_fun_list)),
  FPOP=quote({
    pfit <- fpopw::Fpop(
      data_list[[simulation]], penalty, verbose_file=tempfile())
    count_dt <- pfit$model[, .(intervals=.N), by=data_i]
    data.frame(
      mean_candidates=mean(count_dt$intervals),
      segments=length(pfit$t.est),
      max_seg_size=max(diff(c(0,pfit$t.est))))
  }))
for(algo in names(algo.prune)){
  prune <- algo.prune[[algo]]
  grid_args[[algo]] <- substitute({
    data_value <- data_list[[simulation]]
    fit <- PELT(data_value, penalty, prune=prune)
    dec <- decode(fit$change)
    data.frame(
      mean_candidates=mean(fit$candidates),
      segments=nrow(dec),
      max_seg_size=max(diff(c(0,dec$last.i))))
  }, list(prune=prune))
}
(expr.list <- do.call(atime::atime_grid, grid_args))
```

Above we see a list with `r length(expr.list)` expressions to run, and
a data table with the corresponding number of rows. Note that each expression 

* returns a data frame with one row and three columns that will be used as units to analyze as a function of N.
* should depend on data size N, which does not appear in the
  expressions above, but it is used to define `data_list` in the `setup`
  argument below:

```{r}
cache.rds <- "2025-04-15-PELT-vs-fpopw.rds"
if(file.exists(cache.rds)){
  atime_list <- readRDS(cache.rds)
}else{
  atime_list <- atime::atime(
    N=all_N,
    setup={
      data_list <- list()
      for(simulation in names(sim_fun_list)){
        sim_fun <- sim_fun_list[[simulation]]
        data_mean_vec <- sim_fun(N)
        set.seed(1)
        data_list[[simulation]] <- rnorm(N_data, data_mean_vec, 2)
      }
    },
    expr.list=expr.list,
    seconds.limit=1,
    result=TRUE)
  saveRDS(atime_list, cache.rds)
}
```

Note in the code above that we run the timing inside `if(file.exists`, which is a caching mechanism.
The result is time-consuming to compute, so the first time it is computed, it is saved to disk as an RDS file, and then subsequently read from disk, to save time.
Finally, we plot the different units as a function of data size N, in the figure below.

```{r atime, fig.height=7}
refs_list <- atime::references_best(atime_list)
ggplot()+
  theme(text=element_text(size=12))+
  directlabels::geom_dl(aes(
    N, empirical, color=expr.grid, label=expr.grid),
    data=refs_list$measurements,
    method="right.polygons")+
  geom_line(aes(
    N, empirical, color=expr.grid),
    data=refs_list$measurements)+
  scale_color_manual(
    "algorithm",
    guide="none",
    values=algo.colors)+
  facet_grid(unit ~ simulation, scales="free")+
  scale_x_log10(
    "N = number of data in sequence",
    breaks=10^seq(2,6),
    limits=c(NA, 1e7))+
  scale_y_log10("")
```

Above the figure shows a different curve for each algorithm, and a
different panel for each simulation and measurement unit. We can observe that

* PELT memory usage in kilobytes has a larger slope for constant
  changes, than for linear changes. This implies a larger asymptotic complexity class.
* The maximum segment size, `max_seg_size`, is the same for all algos,
  indicating that the penalty was chosen appropriately, and the
  algorithms are computing the same optimal solution.
* Similarly, the mean number of candidates considered by the PELT
  algorithm has a larger slope for constant changes (same as
  OPART/linear in N), than for linear changes (relatively constant).
* For PELT seconds, we see a slightly larger slope for seconds in
  constant changes, compared to linear changes.
* As expected, number of `segments` is constant for a constant number
  of changes (left), and increasing for a linear number of changes
  (right).
  
Below we plot the same data in a different way, that facilitates
comparisons across the type of simulation:

```{r compare-sims, fig.height=7}
refs_list$meas[, let(
  Simulation = sub("_","\n",simulation),
  Algorithm = factor(expr.grid, algo.levs)
)][]
ggplot()+
  theme(
    legend.position="none",
    text=element_text(size=12))+
  directlabels::geom_dl(aes(
    N, empirical, color=Simulation, label=Simulation),
    data=refs_list$measurements,
    method="right.polygons")+
  geom_line(aes(
    N, empirical, color=Simulation),
    data=refs_list$measurements)+
  facet_grid(unit ~ Algorithm, scales="free")+
  scale_x_log10(
    "N = number of data in sequence",
    breaks=10^seq(2,6),
    limits=c(NA, 1e7))+
  scale_y_log10("")
```

The plot above makes it easier to notice some interesting trends in
the mean number of candidates:

* For PELT and FPOP the mean number of candidates is increases for a
  constant number of changes, but at different rates (FPOP much slower
  than PELT).

```{r plot-refs, fig.width=12, fig.height=7}
expr.levs <- CJ(
  algo=algo.levs,
  simulation=names(sim_fun_list)
)[algo.levs, paste0(algo,"\n",simulation), on="algo"]
edit_expr <- function(DT)DT[
, expr.name := factor(sub(" simulation=", "\n", expr.name), expr.levs)]
edit_expr(refs_list$meas)
edit_expr(refs_list$plot.ref)
plot(refs_list)+
  theme(panel.spacing=grid::unit(1.5,"lines"))
```

The plot above is an empirical verification of our earlier complexity claims.

* OPART `mean_candidates` is linear, `O(N)`, and time and memory are
  quadratic, `O(N^2)`.
* PELT with constant changes has linear `mean_candidates`, `O(N)`, and
  quadratic time and memory, `O(N^2)` (same as OPART, no asymptotic speedup).
* PELT with linear changes has sub-linear `mean_candidates` (constant
  or log), and linear or log-linear time/memory.
* FPOP always has sub-linear `mean_candidates` (constant or log), and
  linear or log-linear time/memory.
  
The code/plot below shows the speedups in this case.

```{r pred-seconds}
pred_list <- predict(refs_list)
ggplot()+
  theme_bw()+
  theme(text=element_text(size=20))+
  geom_line(aes(
    N, empirical, color=expr.grid),
    data=pred_list$measurements[unit=="seconds"])+
  directlabels::geom_dl(aes(
    N, unit.value, color=expr.grid, label=sprintf(
      "%s\n%s\nN=%s",
      expr.grid,
      ifelse(expr.grid=="FPOP", "C++", "R"),
      format(round(N), big.mark=",", scientific=FALSE, trim=TRUE))),
    data=pred_list$prediction,
    method=list(cex=1.5,"top.polygons"))+
  scale_color_manual(
    "algorithm",
    guide="none",
    values=algo.colors)+
  facet_grid(. ~ simulation, scales="free", labeller=label_both)+
  scale_x_log10(
    "N = number of data in sequence",
    breaks=10^seq(2,5),
    limits=c(NA, 1e6))+
  scale_y_log10(
    "Computation time (seconds)",
    breaks=10^seq(-3,0),
    limits=10^c(-4,1))
```

The figure above shows the throughput (data size N) which is possible
to compute in 1 second using each algorithm. We see that FPOP is
10-100x faster than OPART/PELT. Part of the difference is that
OPART/PELT were coded in R, whereas FPOP was coded in C++
(faster). Another difference is that FPOP is asymptotically faster
with constant changes, as can be seen by a smaller slope for FPOP,
compared to OPART/PELT.

Below we zoom in on the number of candidate change-points considered
by each algorithm.

```{r pred-candidates}
cand_dt <- pred_list$measurements[unit=="mean_candidates"]
ggplot()+
  theme_bw()+
  theme(text=element_text(size=20))+
  geom_line(aes(
    N, empirical, color=expr.grid),
    data=cand_dt)+
  directlabels::geom_dl(aes(
    N, empirical, color=expr.grid, label=expr.grid),
    data=cand_dt,
    method=list(cex=2, "right.polygons"))+
  scale_color_manual(
    "algorithm",
    guide="none",
    values=algo.colors)+
  facet_grid(. ~ simulation, scales="free", labeller=label_both)+
  scale_x_log10(
    "N = number of data in sequence",
    breaks=10^seq(2,6),
    limits=c(NA, 1e7))+
  scale_y_log10("Mean number of candidate change-points")
```

The figure above shows that 

* OPART always considers a linear number of candidates (slow).
* PELT also considers a linear number of candidates (slow), when the number
  of changes is constant.
* PELT considers a sub-linear number of candidates (fast), when the
  number of changes is linear.
* FPOP always considers sub-linear number of candidates (fast), which
  is either constant for a linear number of changes, or logarithmic for
  a constant number of changes.

## Conclusions

We have explored three algorithms for optimal change-point detection.

* The classic OPART algorithm is always quadratic time in the number of data to segment.
* The Pruned Exact Linear Time (PELT) algorithm can indeed be linear
  time in the number of data to segment, but only when the number of
  change-points grows with the number of data points. When the number
  of change-points is constant, there are ever-increasing segments,
  and the PELT algorithm must compute a cost for each candidate on
  these large segments; in this situation, PELT suffers from the same
  quadratic time complexity as OPART.
* The FPOP algorithm is fast (linear or log-linear) in both of the
  scenarios we examined.
  
## Session info

```{r}
sessionInfo()
```
