---
layout: post
title: Creating large imbalanced data benchmarks
description: Tutorial with OpenML
---

```{r Ropts, echo=FALSE, results='hide'}
repo.dir <- normalizePath("..")
post.id <- "2025-07-14-imbalance-openml"
fig.path <- paste0(file.path(repo.dir, "assets", "img", post.id), "/")
dir.create(fig.path, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=10, ## TODO python figures wider? look at prev issue.
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to show how to create imbalanced classification data
sets for use with our recently proposed [SOAK
algorithm](https://arxiv.org/abs/2410.08643), which will be able to
tell us if we can train on imbalanced data, and get accurate
predictions on balanced data (and vice versa).

## Read and order OpenML data

OpenML is a web site and repository where we can download machine learning benchmark data sets.
We begin by reading the list of meta-data,

```{r}
omlds <- OpenML::listOMLDataSets(limit=1e6)
library(data.table)
omldt <- data.table(omlds)[order(-number.of.instances)]
class_dt <- omldt[is.finite(number.of.classes)][number.of.classes>1]
unique(class_dt[, .(
  name, rows=number.of.instances, Nclass=number.of.classes,
  Nmajor=majority.class.size, Nminor=minority.class.size
)])[1:100]
```

The table above shows the 100 largest data sets, in terms of number of instances/rows.
The code below reads the largest data set, higgs.

```{r}
higgs.id <- class_dt[1, data.id]
if(FALSE){
  higgs.data <- OpenML::getOMLDataSet(higgs.id)
}
higgs_dt <- fread("higgs.csv")
higgs_dt[1]
```

We see that these data have a `target` column (output to predict). The class distribution
is as follows:

```{r}
(higgs_counts <- higgs_dt[, .(
  count=.N,
  prop=.N/nrow(higgs_dt)
), by=target][order(-prop)])
```

In the table above, we see the overall `count` (number of rows) for each target.
We see in the `prop` column above that there are about an equal amount of data
for each class, from 47% to about 53%.

## Proportional versus random ordering

We would like to preserve these
proportions when we take subsamples of the data. To do that, we first
take a random order of the data (in case the original file had some
special structure), and then we assign a new proportional ordering
based on the label value.

```{r}
my.seed <- 1
set.seed(my.seed)
rand_ord <- higgs_dt[, sample(.N)]
prop_ord <- data.table(y=higgs_dt$target[rand_ord])[
, prop_y := seq(0,1,l=.N), by=y
][, order(prop_y)]
ord_list <- list(
  random=rand_ord,
  proportional=rand_ord[prop_ord])
ord_prop_dt_list <- list()
for(ord_name in names(ord_list)){
  ord_vec <- ord_list[[ord_name]]
  y_ord <- higgs_dt$target[ord_vec]
  for(prop_data in c(0.01, 0.1, 1)){
    N <- nrow(higgs_dt)*prop_data
    N_props <- data.table(y=y_ord[1:N])[, .(
      count=.N,
      prop_y=.N/N
    ), by=y][order(-prop_y)]
    ord_prop_dt_list[[paste(ord_name, prop_data)]] <- data.table(
      ord_name, prop_data, N_props)
  }
}
ord_prop_dt <- rbindlist(ord_prop_dt_list)
dcast(ord_prop_dt, ord_name + prop_data ~ y, value.var="prop_y")
```

The output above shows that the proportional ordering is somewhat more
stable than the random ordering, which has some variations in the `0` and `1`
columns, between the different rows (values of prop_data).

## Creating class imbalance, one imbalance proportion

We would like to split the rows into two subsets, one balanced, and one imbalanced.
The balanced subset will have an equal number of positive and negative labels.
The imbalanced subset should be the same size as the balanced subset, but have imbalance in the labels, with a certain proportion of negative labels, `Pneg`, defined below.

```{r}
(Tpos <- higgs_counts$count[1])
(Tneg <- higgs_counts$count[2])
Pneg <- 0.001
```

Next, we compute the number of rows in each subset. The formulas below can be derived from assuming:

* `N` is the number of positive (and negative) labels in the balanced subset.
* `n_pos` and `n_neg` are the numbers of postive/negative labels in the imbalanced subset: `Pneg(n_neg+n_pos)=n_neg`.
* The subsets are the same size: `n_neg + n_pos= 2*N`.
* All of the positive labels are used: `Tpos = n_pos + N`.

```{r}
(N <- as.integer(Tpos/(3-2*Pneg)))
(n_pos <- Tpos-N)
(n_neg <- as.integer(Pneg*n_pos/(1-Pneg)))
```

The code below can be used to check the results.

```{r}
rbind(n_pos+N, Tpos) # all positive labels are used.
rbind(n_neg+N, Tneg) # negative labels used is less than total negative labels.
rbind(2*N, n_pos+n_neg) # imbalanced and balanced subsets are same size.
```

## Creating class imbalance, several imbalance proportions

We will create different versions of `higgs`, each version having two subsets:

* one which is balanced: 50% positive labels, 50% negative labels.
* one which is imbalanced: a smaller proportion of negative labels
  (from 10% to 0.1%).

To do that, we use the following code to compute the number of samples
we would need:

```{r}
target_prop <- 10^seq(-1, -5)
(smaller_dt <- data.table(
  target_prop,
  N_pos_neg = as.integer(Tpos/(3-2*target_prop))
)[
, n_pos := Tpos-N_pos_neg
][
, n_neg := as.integer(target_prop*n_pos/(1-target_prop))
][
, check_N_im := n_pos+n_neg
][, let(
  check_N_bal = N_pos_neg*2,
  check_prop = n_neg/check_N_im
)][])
```

Above we see 

* `target_prop`, the desired proportion of negative labels in the imbalanced subset,
* `N_pos_neg`, the number of negative (and positive) labels in the balanced subset,
* `n_pos` and `n_neg`, the number of positive/negative labels in the imbalanced subset,
* `check_prop`, the empirical proportion of negative labels in the imbalanced subset, 
* `check_N_bal` and `check_N_im`, the number of rows in the balanced/imbalanced subsets.

It is clear from the table above that the empirical proportions are
consistent with the desired proportions. To create the different imbalanced data
sets, we loop over the rows of this table, to assign rows to each
subset of each imbalanced variant (each corresponding to a column of
`subset_mat`).

```{r}
subset_mat <- matrix(
  NA, nrow(higgs_dt), nrow(smaller_dt),
  dimnames=list(
    NULL,
    target_prop=paste0(
      "seed",
      my.seed,
      "_prop",
      smaller_dt$target_prop)))
emp_y_list <- list()
emp_props_list <- list()
higgs_ord <- higgs_dt[, .(target, .I)][ord_list$proportional]
for(im_i in 1:nrow(smaller_dt)){
  im_row <- smaller_dt[im_i]
  im_count_dt <- im_row[, rbind(
    data.table(subset="b", target=c(0,1), count=N_pos_neg),
    data.table(subset="i", target=c(0,1), count=c(n_neg, n_pos))
  )]
  higgs_ord[, subset := NA_character_]
  for(target_value in c(1,0)){
    tval_dt <- im_count_dt[target==target_value]
    sub_vals <- tval_dt[, rep(subset, count)]
    tval_idx <- which(higgs_ord$target==target_value)
    some_idx <- tval_idx[1:length(sub_vals)]
    higgs_ord[some_idx, subset := sub_vals]
  }
  subset_mat[higgs_ord$I, im_i] <- higgs_ord$subset
  (im_higgs <- data.table(
    target_prop=im_row$target_prop,
    subset=subset_mat[,im_i],
    target=higgs_dt$target
  )[, idx := .I][!is.na(subset)])
  emp_y_list[[im_i]] <- im_higgs[, .(
    count=.N
  ), by=.(target_prop,subset,target)]
  emp_props_list[[im_i]] <- im_higgs[, .(
    count=.N,
    first=idx[1], 
    last=idx[.N]
  ), by=.(target_prop,subset,target)
  ][
  , prop_in_subset := count/sum(count)
  , by=subset
  ][]
}
emp_y <- rbindlist(emp_y_list)
(emp_props <- rbindlist(emp_props_list))
```

The table above can be used to verify that the subset assignments are
consistent with the target label proportions. It has one row for each
unique combination of target proportion, subset, and label (target). For
each value of `target_prop`, we have `prop_in_subset=0.5` when `subset=b` (balanced).

The subset assignment above was based on the idea of using all of the positive targets in each subset.
There are other constraints we could consider:

* Each smaller target proportion is a strict subset: going to 1% from
  10% just involves taking away some samples. For example, with 2000
  samples overall, 10% would look like 500/500 in balanced and 900/100
  imbalanced. Going to 1% would mean 455/455 in balanced (for a total
  of 910), and 900/10 imbalanced (also a total of 910).
* Same data size across imbalance proportions. For example, keeping
  500/500 for each balanced set, and then adjusting the size of the
  imbalanced set, so it always has a total of 1000 samples. For 10% we
  have 900/100, for 1% we have 10/990, etc.

Note that some of the constraints discussed above are mutually
exclusive (can not be done at the same time).

## Write the subset columns to a new CSV file

Finally, we create new columns for each subset.

```{r}
fwrite(subset_mat, "2025-07-14-imbalance-openml.csv")
mysystem <- function(x)cat(system(x, intern=TRUE), sep="\n")
mysystem("head 2025-07-14-imbalance-openml.csv")
mysystem("wc -l 2025-07-14-imbalance-openml.csv")
mysystem("du -k 2025-07-14-imbalance-openml.csv")
```

## Conclusions

This tutorial showed how to divide a data set into two subsets with the same number of samples (balanced and imbalanced). Each
imbalanced data set is represented as a new column (with the same
number of rows as the original data file), with two values: `b` for balanced
and `i` for imbalanced, that can be efficiently saved to a new CSV file
(without having to copy or modify the original data CSV). 
Each column in the resulting CSV files can be used to create a
different mlr3 Task (each with a different definition of subset), so
our recently proposed [SOAK
algorithm](https://arxiv.org/abs/2410.08643) can be used to determine
if a learning algorithm is able to generalize between data subsets
with different proportions of labels (50% minority class versus 1%,
etc).

## Session info

```{r}
sessionInfo()
```
