---
layout: post
title: Comparing neural network architectures using mlr3torch
description: Convolutional network versus linear model
---

```{r Ropts, echo=FALSE, results='hide'}
repo.dir <- normalizePath("..")
post.id <- "2025-03-20-mlr3torch-conv"
fig.path <- paste0(file.path(repo.dir, "assets", "img", post.id), "/")
dir.create(fig.path, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=10, ## TODO python figures wider? look at prev issue.
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this blog is to demonstrate improvements in prediction
accuracy using neural networks, compared to linear models, for image
classification tasks, using mlr3torch in R.

## Introduction

A major advantage of the mlr3torch framework in R, is that it is
relatively simple to compare different learning algorithms, on
different data sets, in terms of prediction accuracy in
cross-validation. For example, [in a previous
post](https://tdhock.github.io/blog/2024/mlr3torch/), we showed that a
torch linear model with early stopping regularization is slightly but
significantly more accurate than an L1-regularized linear model, for
image classification.

The goal here is to reproduce the results similar to [figures in my
Two New Algos
slides](https://github.com/tdhock/cv-same-other-paper?tab=readme-ov-file#27-mar-2025),
which show that convolutional neural networks are more accurate than
linear models in image classification.

## Read MNIST data

We begin by reading the MNIST data,

```{r}
library(data.table)
MNIST_dt <- fread("~/projects/cv-same-other-paper/data_Classif/MNIST.csv")
dim(MNIST_dt)
data.table(
  name=names(MNIST_dt),
  first_row=unlist(MNIST_dt[1]),
  last_row=unlist(MNIST_dt[.N]))
```

We then create a `label` factor column, and a task:

```{r}
MNIST_dt[, label := factor(y)]
mtask <- mlr3::TaskClassif$new(
  "MNIST", MNIST_dt, target="label")
mtask$col_roles$stratum <- "label"
mtask$col_roles$feature <- grep("^[0-9]+$", names(MNIST_dt), value=TRUE)
mtask
```

The Task above is the mlr3 representation of the MNIST data set with
the meta-data about which columns should be used for
input/output/stratum/etc.

## Defining neural network architecture

We can use the function below to define neural networks with the same
learning rate, step size, etc, but different architecture (number of
hidden layers/units).

```{r}
measure_list <- mlr3::msrs(c("classif.logloss", "classif.ce"))
n.epochs <- 200
make_torch_learner <- function(id,...){
  po_list <- c(
    list(
      mlr3pipelines::po(
        "select",
        selector = mlr3pipelines::selector_type(c("numeric", "integer"))),
      mlr3torch::PipeOpTorchIngressNumeric$new()),
    list(...),
    list(
      mlr3pipelines::po("nn_head"),
      mlr3pipelines::po(
        "torch_loss",
        mlr3torch::t_loss("cross_entropy")),
      mlr3pipelines::po(
        "torch_optimizer",
        mlr3torch::t_opt("sgd", lr=0.1)),
      mlr3pipelines::po(
        "torch_callbacks",
        mlr3torch::t_clbk("history")),
      mlr3pipelines::po(
        "torch_model_classif",
        batch_size = 100,
        patience=n.epochs,
        measures_valid=measure_list,
        measures_train=measure_list,
        predict_type="prob",
        epochs = paradox::to_tune(upper = n.epochs, internal = TRUE)))
    )
    graph <- Reduce(mlr3pipelines::concat_graphs, po_list)
    glearner <- mlr3::as_learner(graph)
    mlr3::set_validate(glearner, validate = 0.5)
    mlr3tuning::auto_tuner(
      learner = glearner,
      tuner = mlr3tuning::tnr("internal"),
      resampling = mlr3::rsmp("insample"),
      measure = mlr3::msr("internal_valid_score", minimize = TRUE),
      term_evals = 1,
      id=id,
      store_models = TRUE)
}
```

The function above has a variable number of inputs:

* the first argument is a `learner_id` to display in the results.
* the other arguments are used to define the neural network
  architecture, after torch ingress numeric, and before `nn_head`,
  which is mlr3torch terminology for the last layer of the neural
  network (output size determined by the task).
  
The code below defines a list of learners:

```{r}
n.pixels <- 28
learner.list <- list(
  make_torch_learner("torch_linear"),
  make_torch_learner(
    "torch_dense_50",
    mlr3pipelines::po(
      "nn_linear",
      out_features = 50),
    mlr3pipelines::po("nn_relu_1", inplace = TRUE)
  ),
  make_torch_learner(
    "torch_conv",
    mlr3pipelines::po(
      "nn_reshape",
      shape=c(-1,1,n.pixels,n.pixels)),
    mlr3pipelines::po(
      "nn_conv2d_1",
      out_channels = 20,
      kernel_size = 6),
    mlr3pipelines::po("nn_relu_1", inplace = TRUE),
    mlr3pipelines::po(
      "nn_max_pool2d_1",
      kernel_size = 4),
    mlr3pipelines::po("nn_flatten"),
    mlr3pipelines::po(
      "nn_linear",
      out_features = 50),
    mlr3pipelines::po("nn_relu_2", inplace = TRUE)
  ),
  mlr3::LearnerClassifFeatureless$new()$configure(id="featureless"),
  mlr3learners::LearnerClassifCVGlmnet$new()$configure(id="cv_glmnet")
)
```

The code above uses `make_torch_learner` three times:

* `torch_linear` creates a linear model (one layer of parameters to
  learn). Even with no architecture arguments, the `nn_head` generates
  a linear layer which is compatible with the inputs and outputs.
* `torch_dense_50` creates a neural network with one hidden layer, of
  50 hidden units, followed by ReLU activation (two layers of
  parameters to learn).
* `torch_conv` creates a neural network with a 2D convolution,
  followed by ReLU activation, followed by 2D max pooling, followed by
  a flatten operation and a linear layer, then a final ReLU (three
  layers of parameters to learn).
  
## Defining and computing a benchmark grid

The code below combines the task, learners, and a resampling method
(3-fold cross-validation), in a benchmark grid:

```{r}
kfoldcv <- mlr3::rsmp("cv")
kfoldcv$param_set$values$folds <- 3
(bench.grid <- mlr3::benchmark_grid(
  mtask,
  learner.list,
  kfoldcv))
```

The code below does the computation.

```{r}
reg.dir <- "2025-03-20-mlr3torch-conv"
cache.RData <- paste0(reg.dir,".RData")
if(file.exists(cache.RData)){
  load(cache.RData)
}else{
  if(FALSE){#code below only works on the cluster.
    unlink(reg.dir, recursive=TRUE)
    reg = batchtools::makeExperimentRegistry(
      file.dir = reg.dir,
      seed = 1,
      packages = "mlr3verse"
    )
    mlr3batchmark::batchmark(
      bench.grid, store_models = TRUE, reg=reg)
    job.table <- batchtools::getJobTable(reg=reg)
    chunks <- data.frame(job.table, chunk=1)
    batchtools::submitJobs(chunks, resources=list(
      walltime = 60*60,#seconds
      memory = 2000,#megabytes per cpu
      ncpus=1,  #>1 for multicore/parallel jobs.
      ntasks=1, #>1 for MPI jobs.
      chunks.as.arrayjobs=TRUE), reg=reg)
    batchtools::getStatus(reg=reg)
    jobs.after <- batchtools::getJobTable(reg=reg)
    table(jobs.after$error)
    ids <- jobs.after[is.na(error), job.id]
    bench.result <- mlr3batchmark::reduceResultsBatchmark(ids, reg = reg)
  }else{
    ## In the code below, we declare a multisession future plan to
    ## compute each benchmark iteration in parallel on this computer
    ## (data set, learning algorithm, cross-validation fold). For a
    ## few dozen iterations, using the multisession backend is
    ## probably sufficient (I have 12 CPUs on my work PC).
    if(require(future))plan("multisession")
    bench.result <- mlr3::benchmark(bench.grid, store_models = TRUE)
  }
  save(bench.result, file=cache.RData)
}
```

The code above took several hours on my 14-core laptop.

## Test error results

First, we use `score()` to compute test error rates:

```{r}
score_dt <- bench.result$score()
(score_some <- score_dt[order(classif.ce), .(
  learner_id=factor(learner_id, unique(learner_id)),
  iteration,
  percent_error=100*classif.ce)])
```

The table above has one row for each learning algorithm, and each
train/test split in 3-fold CV (iteration). It is sorted by error rate,
so it is easy to see that the convolutional network is most accurate,
and all algorithms have smaller test error rates than featureless.
Below we visualize these test error rates in a figure:

```{r percent-err-all, fig.height=2}
library(ggplot2)
ggplot()+
  geom_point(aes(
    percent_error, learner_id),
    shape=1,
    data=score_some)+
  scale_x_continuous(
    breaks=seq(0,100,by=10),
    limits=c(0,90))
```

The figure above shows a dot for each learning algorithm and
train/test split.  The algorithms on the Y axis are sorted by error
rate, so it is easy to see that `torch_conv` has the smallest error
rate, the `torch_dense_50` has slightly larger error rate,
etc. Interestingly, the `torch_linear` model (early stopping
regularization) has a slightly smaller error rate than `cv_glmnet`
(linear model with L1 regularization).

Below we compute mean and standard deviation for each learning
algorithm, across the three cross-validation splits:

```{r}
(score_stats <- dcast(
  score_some,
  learner_id ~ .,
  list(mean, sd),
  value.var="percent_error"))
```

The table above has one row for each learning algorithm, and columns
for mean and standard deviation of the error rate. We visualize these
numbers in the figure below:

```{r percent-err-mean-sd, fig.height=2}
score_show <- score_stats[learner_id!="featureless"]
ggplot()+
  geom_point(aes(
    percent_error_mean, learner_id),
    shape=1,
    data=score_show)+
  geom_segment(aes(
    percent_error_mean+percent_error_sd, learner_id,
    xend=percent_error_mean-percent_error_sd, yend=learner_id),
    data=score_show)+
  geom_text(aes(
    percent_error_mean, learner_id,
    label=sprintf("%.2f±%.2f", percent_error_mean, percent_error_sd)),
    vjust=-0.5,
    data=score_show)+
  coord_cartesian(xlim=c(0,10))+
  scale_x_continuous(
    "Percent error on test set (mean ± SD over 3 folds in CV)",
    breaks=seq(0,10,by=2))
```

The figure above omits featureless, to emphasize the subtle
differences between the non-trivial learning algorithms. To compute
p-values for differences, we can do the following.

```{r}
(levs <- levels(score_some$learner_id))
(pval_dt <- data.table(comparison_i=1:3)[, {
  two_levs <- levs[comparison_i+c(0,1)]
  lev2rank <- structure(c("lo","hi"), names=two_levs)
  i_long <- score_some[
    learner_id %in% two_levs
  ][
  , rank := lev2rank[paste(learner_id)]
  ][]
  i_wide <- dcast(i_long, iteration ~ rank, value.var="percent_error")
  paired <- with(i_wide, t.test(lo, hi, alternative = "l", paired=TRUE))
  unpaired <- with(i_wide, t.test(lo, hi, alternative = "l", paired=FALSE))
  data.table(
    learner.lo=factor(two_levs[1],levs),
    learner.hi=factor(two_levs[2],levs),
    p.paired=paired$p.value,
    p.unpaired=unpaired$p.value,
    mean.diff=paired$est,
    mean.lo=unpaired$est[1],
    mean.hi=unpaired$est[2])
}, by=comparison_i])
```

The table above has one row for each comparison, with columns for mean
error rates and P-values (in paired and unpaired T-tests).
These P-values are shown in the figure below.

```{r pval, fig.height=2}
ggplot()+
  geom_segment(aes(
    mean.lo, learner.lo,
    xend=mean.hi, yend=learner.lo),
    data=pval_dt,
    size=2,
    color="red")+
  geom_text(aes(
    x=(mean.lo+mean.hi)/2, learner.lo,
    label=sprintf("P=%.4f", p.paired)),
    data=pval_dt,
    vjust=1.5,
    color="red")+
  geom_point(aes(
    percent_error_mean, learner_id),
    shape=1,
    data=score_show)+
  geom_segment(aes(
    percent_error_mean+percent_error_sd, learner_id,
    xend=percent_error_mean-percent_error_sd, yend=learner_id),
    size=1,
    data=score_show)+
  geom_text(aes(
    percent_error_mean, learner_id,
    label=sprintf("%.2f±%.2f", percent_error_mean, percent_error_sd)),
    vjust=-0.5,
    data=score_show)+
  coord_cartesian(xlim=c(0,10))+
  scale_y_discrete("algorithm")+
  scale_x_continuous(
    "Percent error on test set (mean ± SD over 3 folds in CV)",
    breaks=seq(0,10,by=2))
```

The plot above contains P-values in red, which show that 

* the dense neural network has a significantly larger error rate than the convolutional neural network,
* the torch linear model has a significantly larger error rate than the dense neural network,
* the `cv_glmnet` linear model has slightly larger error rate than the
  torch linear model, but this difference is not statistically
  significant (at least using the classic threshold of P=0.05).

## Checking subtrain/validation loss curves

Next, we can check if we have used enough epochs for the early
stopping regularization in torch. If we have used enough, we should
see the min validation loss at an intermediate value, not at the
smallest number of epochs (underfitting), nor at the largest number of
epochs (overfitting). First, we extract the number of epochs that was
used by mlr3 auto-tuner, for the torch learners:

```{r subtrain-valid-logloss}
(score_torch <- score_dt[
  grepl("torch",learner_id)
][
, best_epoch := sapply(
  learner, function(L)unlist(L$tuning_result$internal_tuned_values))
][])
```

The table above has a `best_epoch` column which shows the number of epochs that was chosen by the mlr3 auto-tuner. 
Next, we extract the history that was used to select the best epoch:

```{r}
(history_torch <- score_torch[, {
  L <- learner[[1]]
  M <- L$archive$learners(1)[[1]]$model
  M$torch_model_classif$model$callbacks$history
}, by=.(learner_id, iteration)])
```

The table above has one row for each learner, iteration, and epoch of training. 
There are columns named train/valid logloss/ce, which we convert from wide to long format below:

```{r}
(history_long <- nc::capture_melt_single(
  history_torch,
  set=nc::alevels(valid="validation", train="subtrain"),
  ".classif.",
  measure=nc::alevels("logloss", ce="prop_error")))
```

The table above has a `set` column that is consistent with the Deep Learning book of Goodfellow:

* The full data set is split into train and test sets,
* and then the train set is split into subtrain and validation sets.

The code below can be used to visualize the logistic loss for each epoch:

```{r}
ggplot()+
  theme_bw()+
  theme(legend.position=c(0.9, 0.15))+
  geom_vline(aes(
    xintercept=best_epoch),
    data=score_torch)+
  geom_text(aes(
    best_epoch, Inf, label=paste0(" best epoch=", best_epoch)),
    vjust=1.5, hjust=0,
    data=score_torch)+
  geom_line(aes(
    epoch, value, color=set),
    data=history_long[measure=="logloss"])+
  facet_grid(iteration ~ learner_id, labeller=label_both)+
  scale_y_log10("logistic loss")+
  scale_x_continuous("epoch")
```

The figure above shows 

* panels from left to right for different learning algorithms (neural network architectures),
* panels from top to bottom for different train/test splits in cross-validation (iteration in mlr3 terms),
* Y axis on the log scale to emphasize subtle differences. For example
  it is clear that the subtrain loss gets much smaller for the
  convolutional model, than for the dense/linear models.
* a vertical black line in each panel to indicate the best epoch used by the mlr3 auto-tuner. It is clear that this is the min of the validation loss.

The code below considers only the first train/test split in cross-validation, and shows the classification error in addition to the log loss:

```{r subtrain-valid-both}
get_fold <- function(DT,it=1)DT[iteration==it]
history_fold <- get_fold(history_long)
score_fold <- get_fold(score_torch)
min_fold <- history_fold[
, .SD[value==min(value)]
, by=.(learner_id, measure, set)
][, point := "min"]
ggplot()+
  theme_bw()+
  theme(legend.position=c(0.9, 0.2))+
  geom_vline(aes(
    xintercept=best_epoch),
    data=score_fold)+
  geom_text(aes(
    best_epoch, Inf, label=paste0(" best epoch=", best_epoch)),
    vjust=1.5, hjust=0,
    data=score_fold)+
  geom_line(aes(
    epoch, value, color=set),
    data=history_fold)+
  geom_point(aes(
    epoch, value, color=set, fill=point),
    shape=21,
    data=min_fold)+
  scale_fill_manual(values=c(min="black"))+
  facet_grid(measure ~ learner_id, labeller=label_both, scales="free")+
  scale_x_continuous("epoch")+
  scale_y_log10("")
```

The figure above additionally has black dots to emphasize the minima of each curve.
It is clear that the vertical black line (best number of epochs from mlr3 auto-tuner) 
agrees with the min validation loss (black dot computed using min of each history curve).
Finally, we see that the neural networks overfit with perfect classification accuracy on the subtrain set (black dots at bottom of panel), whereas the linear model subtrain error rate never goes below 1%. 

## Conclusions

We have explained how to use mlr3torch to compare various neural
network architectures for image classification, on the MNIST data set.
We showed how to code figures that compare test error rates, and allow
to check if the max number of epochs was selected appropriately.  We
saw that a convolutional neural network with three layers of weights
was more accurate than a dense neural network with two layers of
weights, which was in turn more accurate than a linear model with only
one layer of weights.  Using a learning rate of 0.1 and a batch size
of 100, we saw that 200 epochs was sufficient to observe overfitting,
and also to avoid underfitting.

## Session info

```{r}
sessionInfo()
```
