---
layout: post
title: Bike ride map and time series data viz
description: A demonstration of animint2 and sf
---

```{r Ropts, echo=FALSE, results='hide'}
repo.dir <- normalizePath("..")
post.id <- "2025-01-19-bike-ride-data-viz"
fig.path <- paste0(file.path(repo.dir, "assets", "img", post.id), "/")
dir.create(fig.path, showWarnings = FALSE, recursive = TRUE)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=fig.path,
  fig.width=10, ## TODO python figures wider? look at prev issue.
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
conda.env <- "2023-08-deep-learning"
conda.env <- "torch-aum"
RETICULATE_PYTHON <- sprintf(if(.Platform$OS.type=="unix")
  ##"/home/tdhock/.local/share/r-miniconda/envs/%s/bin/python"
  "/home/tdhock/miniconda3/envs/%s/bin/python"
  else "~/AppData/Local/Miniconda3/envs/%s/python.exe", conda.env)
Sys.setenv(RETICULATE_PYTHON=RETICULATE_PYTHON)
##reticulate::use_condaenv(dirname(RETICULATE_PYTHON), required=TRUE)
in_render <- !is.null(knitr::opts_knit$get('rmarkdown.pandoc.to'))
in_knit <- isTRUE(getOption('knitr.in.progress'))
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The goal of this post is to explain how to use
[animint2](https://github.com/animint/animint2/), my R package for
animated interactive data visualization, to create a linked map and
time series of bike rides I did in 2009.

## Background: master internship motivation for interactive graphics

Since my master internship at INRA Jouy-en-Josas in 2009, I had some
ideas about using interactive data visualization to better understand
complex data and algorithms. At that time, I was studying an iterative
sampling algorithm for inferring the parameters of a hierarchical
Bayesian model. The best visualization tools at the time were
lattice/ggplot2 and animation packages in R. Using that approach, the
only interactions possible are moving forward and backward in time,
which resulted in data visualizations such as [this
one](https://nicholsonppp.r-forge.r-project.org/2009-08-19/index.htm),
which remarkably still works after 15 years! You can click play and
watch the animation like a video, but clicking on the plots does not
accomplish any interaction.

But it could be useful to be able to click the plot, for example to
update the selected locus or time point! That is called "direct
manipulation" in the interactive graphics literature. How could I make
graphics with those kinds of interactions? They are a very specific
kind of interaction, "click on some visual element, to change the
highlighted value of some variable." (locus or generation in that
example) At the end of my master internship, I decided to shift gears,
and focus on studying the math behind machine learning, so these
interactive graphics ideas sat on the backburner of my mind for a few
years.

## First version of animint

In 2012, I defended my PHD in Paris, and then flew out to Tokyo to
spend 2013 in Masashi Sugiyama's machine learning lab. During that
time, I got to working on my interactive graphics ideas again, this
time with a focus on visualizing machine learning algorithms. I coded
a first version of the [animint](https://github.com/tdhock/animint) R
package, and Susan VanderPlas worked in GSOC'13 to implement a bunch
of key features. I then created a first
[gallery](https://rcdata.nau.edu/genomic-ml/public_html/animint/) of
examples, including a [re-make of my master internship data
viz](https://rcdata.nau.edu/genomic-ml/public_html/animint/evolution/viz.html),
but with the interactive features that I had envisioned.
On that web page, 

* the left plot shows a time series for the selected locus, and you
  can click to select a different time point (generation).
* the right plot shows an overview of all loci for the selected time
  point (generation), and you can click to select a different locus.

How do we define that data visualization? We use R/ggplot2 code, with
two new keywords that can be defined for any geom:

* `showSelected` means to only show data rows that correspond to the values of the current selection.
* `clickSelects` means that the geom can be clicked to change the current selection.

The code on that web page uses the old animint package, which is no longer recommended. Below we port the code to the new animint2 package, which is recommended:

```{r}
library(animint2)
data(generation.loci)
##generation.loci <- subset(generation.loci, generation %% 10 == 1)
colormap <- c(blue="blue",red="red",ancestral="black",neutral="grey30")
ancestral <- subset(generation.loci,population==1 & generation==1)
ancestral$color <- "ancestral"
viz <- animint(
  title="Evolution simulation with drift and selection",
  source="https://tdhock.github.io/blog/2025/bike-ride-data-viz/",
  ts=ggplot()+
    ggtitle("Click to select generation")+
    theme(legend.position="none")+
    geom_line(aes(
      generation, frequency, group=paste(population,color),
      colour=color),
      showSelected=c("locus","color"),
      data=generation.loci)+
    make_text(generation.loci, 50, 1.05, "locus")+
    make_tallrect(generation.loci, "generation")+
    scale_colour_manual("population", values=colormap)+
    geom_point(aes(
      generation, frequency, color=color),
      showSelected=c("locus","color"),
      data=ancestral),
  loci=ggplot()+
    ggtitle("Click to select locus")+
    geom_point(aes(
      locus, frequency, colour=color,
      key=paste(locus, population)),
      showSelected="generation",
      chunk_vars=character(),
      data=generation.loci)+
    geom_point(aes(
      locus, frequency, colour=color),
      data=ancestral)+
    scale_colour_manual("population", values=colormap)+
    make_tallrect(generation.loci, "locus")+
    make_text(generation.loci, 35, 1, "generation"),
  duration=list(generation=1000),
  time=list(variable="generation", ms=2000))
if(FALSE){
  animint2pages(viz, "2025-01-19-evolution-drift-selection")
}
```

The interactive data viz created using the code above [can be viewed
here](https://tdhock.github.io/2025-01-19-evolution-drift-selection).
Using this method, it is remarkably simple to create a wide variety of
different data visualizations. The point of the [original
gallery](https://rcdata.nau.edu/genomic-ml/public_html/animint/) was
to highlight the variety of different kinds of visualizations which
are possible to create. The [modern
gallery](https://animint.github.io/gallery/) contains a much larger
array of example data visualizations, but remarkably does not contain
the interaction instructions that were present in the original gallery
(click X to change Y). That kind of information can be very useful for
somebody who does not know how to interact with the data
visualization. Similar "help the user understand what interactions are
possible" features are planned for the upcoming GSOC'25, which will
hopefully implement a [guided
tour](https://github.com/animint/animint2/issues/150) and [tutorial
videos](https://github.com/animint/animint2/issues/151).

## Back to the master

Right now I am working in Paris again, and I thought it would be fun
to re-visit another old project, [gpsdb, GPS data website
generator](https://gpsdb.sourceforge.net/). For my birthday in 2009,
my classmates gifted me a GPS, which I used to create data sets for my
bike rides in and around Paris. The gpsdb software was a couple of
python scripts that could download data from the GPS, and then convert
it to KML files, which were required for visualizing the data on
google maps, back in the day. Remarkably, the web site that I created,
and those data files, are still online, 15 years later. For example
[all.kml](https://gpsdb.sourceforge.net/out/all.kml) contains all of
the GPS traces I created in 2009. However google maps no longer
accepts arbitrary KML files (these are too large), so how can we
visualize them? Below we explain how to do it using animint2.

## Clone gpsdb and read GPS data

First step: clone the repository to get the GPS data.

```{r}
if(!file.exists("gpsdb-code")){
  system("hg clone http://hg.code.sf.net/p/gpsdb/code gpsdb-code")
}
gpx.glob <- "gpsdb-code/gpx/*"
library(data.table)
(gpx.dt <- data.table(gpx=Sys.glob(gpx.glob)))
```

The table above shows one row for each GPX data file.
Each one of them has contents as below:

```{r}
one.file <- gpx.dt$gpx[1]
gpx.lines <- readLines(one.file)
cat(gpx.lines[1:20],sep="\n")
```

The output above includes lat and lon values that we would like to
extract and visualize on a map.  But oddly, these files have lat and
lon reversed (Paris is actually near longitude=2 degrees, not
latitude=2 degrees as is shown in the output above). To read these data, we can
use the function below, which switches lat and lon:

```{r}
name_pat <- function(name, pattern)list(
  pattern, '="', 
  nc::group(name, "[0-9.]+", as.numeric))
read_lat_lon <- function(path)nc::capture_all_str(
  path,
  name_pat("longitude","lat"),
  '" ',
  name_pat("latitude","lon"))
read_lat_lon(one.file)
```

The code above uses regular expressions to read the GPS data. The
`name_pat` function is useful for defining a pattern that occurs twice
(lat and lon). The `read_lat_lon` function returns a data table with
`longitude` and `latitude` columns, and one row for each GPS ping in
the file, as can be seen in the output above.

To read all of the files into a data table, we use the code below,

```{r}
(lat.lon.dt <- nc::capture_first_glob(
  gpx.glob,
  "/", timestamp="[^/]+", "[.]gpx$",
  READ=read_lat_lon))
```

The output above includes a `timestamp` column which comes from the
file name, and the other columns which come from the result of reading
the file contents. The `gpx.glob` defines a set of file names, which
are matched to the provided pattern, with named arguments/groups
included in the output (`timestamp`). The `READ` argument specifies
how to read each file, via the `read_lat_lon` function.

Already we can visualize these data on a map:

```{r pathAll}
library(animint2)
ggplot()+
  geom_path(aes(
    longitude, latitude, group=timestamp),
    data=lat.lon.dt)
```

In the plot above, we see that there is a large set of data in the
upper left, and a small amount of data in the lower right.
We can zoom to the upper left via the code below:

```{r pathZoom}
path.show <- lat.lon.dt[longitude < 4]
(gg.path <- ggplot()+
  geom_path(aes(
    longitude, latitude, group=timestamp),
    data=path.show))
```

The figure above kind of looks like a spider web, with the center
being Paris. 

We can add details about the start and end of each trip via the code below:

```{r pathEnds}
start.end.dt <- path.show[
, .SD[c(1,.N), .(
  latitude,longitude,where=c("start","end"),
  direction=diff(longitude))]
, by=timestamp]
gg.path+
  geom_point(aes(
    longitude, latitude, fill=where),
    shape=21,
    data=start.end.dt)+
  scale_fill_manual(values=c(start="white",end="black"))
```

The plot above shows a dot at the start and end of each trip.

## Context: nearby cities

The current map is not super useful currently, because we do not see
names of the start/end cities (nor a familiar unit of distance like
kilometers). To get the context of nearby cities, we can download a
data set which has lat/lon coordinates of each city in France.

```{r allCities}
if(!file.exists("gps-villes-de-france.csv")){
  download.file("https://www.data.gouv.fr/fr/datasets/r/51606633-fb13-4820-b795-9a2a575a72f1", "gps-villes-de-france.csv")
}
(villes.dt <- fread("gps-villes-de-france.csv"))
ggplot()+
  geom_point(aes(
    longitude, latitude),
    data=villes.dt)
```

Looking at the plot above, we see a hexagon in the upper right, which
corresponds to "France métropolitaine" (mainland European France). The
other points correspond to outlying islands and other territories. Our
map only needs context of the Paris region, so we don't want to
display all of these cities. Let's just display the nearest cities to
the start and end. To do that we can use
[`sf::st_nearest_feature`](https://r-spatial.github.io/sf/reference/st_nearest_feature.html),
but first we have to convert our data to `sf` structures, as in the
code below.

* `sf::st_point` creates a single geometry representing a point on a map.
* `sf::st_sfc` creates a list column of geometries, 
* `sf::st_sf` creates a data frame with a simple feature list column
  (so you can associate various other data to each geometry, such as
  timestamp, etc).

```{r}
(start.end.sf <- sf::st_sf(start.end.dt[, .(
  timestamp, where,
  geometry=sf::st_sfc(apply(
    cbind(longitude, latitude),
    1,
    sf::st_point,
    simplify=FALSE
  ))
)]))
(villes.sf <- sf::st_sf(villes.dt[, .(
  label,
  geometry=sf::st_sfc(apply(
    cbind(longitude, latitude),
    1,
    sf::st_point,
    simplify=FALSE
  ))
)]))
nearest.index.vec <- sf::st_nearest_feature(start.end.sf, villes.sf)
(villes.start.end <- data.table(
  start.end.dt[,.(timestamp,direction,where)],
  villes.dt[nearest.index.vec]))
```

The table above contains one row for each start/end position, and
columns that describe the closest nearby city, which can be displayed via:

```{r nearbyCities}
ggplot()+
  geom_path(aes(
    longitude, latitude, group=timestamp, color=what),
    data=data.table(path.show, what="ride"))+
  geom_point(aes(
    longitude, latitude, fill=where, color=what),
    shape=21,
    data=data.table(start.end.dt, what="ride"))+
  scale_fill_manual(values=c(start="white",end="black"))+
  scale_color_manual(values=c(
    ride="black",
    "nearby cities"="red"))+
  geom_point(aes(
    longitude, latitude, fill=where,
    color=what),
    shape=21,
    data=data.table(villes.start.end, what="nearby cities"))+
  geom_text(aes(
    longitude, latitude, label=label,
    color=what,
    hjust=ifelse(
      direction<0,
      ifelse(where=="start", 0, 1),
      ifelse(where=="start", 1, 0))),
    showSelected=c("timestamp","where"),
    data=data.table(villes.start.end, what="nearby cities"))
```

The map above is overplotted, so interactivity is useful to focus on
some details. We would like to be able to click on one trip, and show
its details only. How to implement that? Using animint2 we would specify

* `clickSelects="timestamp"` to specify that we want to click on an
  element to specify the selected ride (file timestamp).
* `showSelected="timestamp"` to specify that we want to display only
  the subset of data for the currently selected ride (file timestamp).
  
The end result looks like
[this](https://tdhock.github.io/2025-01-19-bike-rides-around-paris-2009/). Exercise
for the reader: starting from the code above, can you create the
interactive map and time series yourself?

## Conclusions

We have explored how to read GPS bike ride data into R, along with
city locations for context, and then display them together on a
map. We explained various features of `animint2` for interactive data
visualization, and `sf` for spatial data manipulation (finding the
closest city to the start/end coordinates of each ride).

## Session info

```{r}
sessionInfo()
```
