---
layout: post
title: More precise imbalanced data generation
description: No more rounding issues
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-08-04-cran-dates"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
options(width=120)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to automatically get publication dates off CRAN Archive web pages.

To construct a related work table for a paper, we want publication dates from CRAN packages.

As mentioned in [my previous post](https://data-table-raft.netlify.app/posts/2026-04-15-happy_birthday-toby_hocking/) about `data.table`’s 20th birthday, we can download the CRAN archive meta-data via:

```{r}
adb <- tools::CRAN_archive_db()
head(adb)
```

Above we see a list of matrices, one per package.

Below we define the packages of interest by parsing an org-mode readme file using a regular expression.

```{r}
(pdt <- nc::capture_all_str(
  "~/projects/sparse-label-positions/README.org",
  " ",
  pkg="[^ ]+?",
  "::"))
```

Above we see a table with one row per package of interest.

Below we query the current CRAN web pages to determine if each package is currently published.

```{r}
pdt[, available := {
  tfile <- file.path(tempdir(), pkg)
  if(!file.exists(tfile)){
    u <- sprintf("https://cran.r-project.org/web/packages/%s", pkg)
    download.file(u, tfile)
  }
  adt <- nc::capture_all_str(tfile, dl="tar.gz")
  nrow(adt)>0
}, by=pkg][]
```

Above we see that all but two packages are currently available.

Below we create a data table for packages of interest.

```{r}
library(data.table)
(time.dt <- pdt[
, as.data.table(adb[[paste(pkg)]])[, .(mtime)], by=.(pkg=factor(pkg, pkg), available)
][
, year := strftime(mtime, "%Y")
][])
```

Above we see a table with one row per publication time, per package of interest.

Below we reshape to one row per package:

```{r}
time.wide <- dcast(
  time.dt,
  pkg + available ~ .,
  list(min, max),
  value.var=c("mtime", "year")
)[
, years := sprintf("%s–%s", year_min, year_max)
][]
time.wide[, .(pkg, available, years)]
```

Above we see the packages presented in the same order as they were defined.

```{r}
time.wide[order(mtime_min), .(pkg, available, years)]
```

Above packages are sorted by first publication date.

Below packages are sorted by last publication date.

```{r}
time.wide[order(mtime_max), .(pkg, available, years)]
```

## Conclusions

We have shown how to get CRAN publication dates for a list of R packages of interest.

## session info

```{r}
sessionInfo()
```
