---
layout: post
title: More precise imbalanced data generation
description: No more rounding issues
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-06-11-exact-downsampling"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
options(width=120)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to continue the exploration of [Creating large imbalanced data benchmarks](https://tdhock.github.io/blog/2025/imbalance-openml/).

## Problem

We are creating machine learning algorithms for imbalanced classification problems.
We would like to know if these algorithms work well for training on imbalanced data, and predicting on balanced data, and vice versa.
We therefore would like to create different subsets for testing.
Assume we start with the `higgs` data, binary classification with this many data in each class:

```{r}
Tpos=5829123
Tneg=5170877
```

Below are the target proportions of the minority negative class:

```{r}
lseq <- 10^seq(-1, -5)
(Target_prop <- sort(c(4.5*lseq, lseq)))
```

Below we repeat the calculations in the previous post:

```{r}
library(data.table)
(orig_dt <- data.table(
  Target_prop,
  N_pos_neg = as.integer(Tpos/(3-2*Target_prop))
)[
, n_pos := Tpos-N_pos_neg
][
, n_neg := as.integer(Target_prop*n_pos/(1-Target_prop))
][
, n_imb := n_pos+n_neg
][, let(
  pos=n_pos+N_pos_neg,
  neg=n_neg+N_pos_neg,
  check_prop = n_neg/n_imb
)][])
```

* `Target_prop` is the desired proportion of negative samples in the imbalanced subset.
* `N_pos_neg` is the number of positive and negative samples in the balanced subset.
* `n_pos` and `n_neg` are the numbers of samples in the imbalanced subset, and `n_imb` is their sum.
* `pos` and `neg` are the total numbers of samples, across both subsets. We see that `pos=Tpos` in every row, and that `neg<Tneg` is true in every row except the last.
* `check_prop` is the actual proportion of negative samples in the imbalanced subset.

The numbers above seem reasonable, but

* `Target_prop=4.5e-01` (almost balanced) has `neg>Tneg=5170877` which is not possible (not enough negative samples).
* the `Target_prop` (desired imbalance) is often a bit larger than `check_prop` (actual imbalance).

## Proposed fix

To fix the issues above, we start with the idea that for any target proportion, we want `n_neg` to be the largest integer such that the number of positive samples used is less than or equal the total number of samples: `N + n_pos <= Tpos` (and same for negative).
That gives us the code below, which also works for negative proportions greater than 0.5:

```{r}
compute_target_counts <- function(p_neg, Tpos, Tneg){
  p_small <- ifelse(p_neg<0.5, p_neg, 1-p_neg)
  n_small = as.integer(pmin(
    2*Tpos*p_small/(3*(1-p_neg)+p_neg),
    2*Tneg*p_small/(1+2*p_neg)
  ))
  data.table(p_neg)[, let(
    n_pos = ifelse(
      p_neg<0.5,
      n_small*(1-p_neg)/p_neg,
      n_small),
    n_neg = ifelse(
      p_neg<0.5,
      n_small,
      n_small*p_neg/(1-p_neg))
  )][
  , n_imb := n_pos+n_neg
  ][
  , N_pos := n_imb/2
  ][
  , N_neg := n_imb-N_pos
  ][, let(
    pos=n_pos+N_pos,
    neg=n_neg+N_neg
  )][
  , check_prop := n_neg/n_imb
  ][]
}
p_neg <- sort(c(Target_prop, 1-Target_prop), decreasing = TRUE)
compute_target_counts(p_neg, Tpos, Tneg)
```

Above we see the issues are fixed:

* Every row has a valid `neg <= Tneg = 5170877` (no more `neg` values that are too large to create), even for large values of the target proportion for the negative class, `p_neg >= 0.45`.
* `check_prop` is equal to `p_neg` in every row (no more small differences).

Below we do the computation for some other target proportions:

```{r}
compute_target_counts(
  sort(seq(0.25, 0.75, by=0.05), decreasing = TRUE),
  Tpos, Tneg)
compute_target_counts(
  sort(seq(0.45, 0.55, by=0.01), decreasing = TRUE),
  Tpos, Tneg)
compute_target_counts(
  sort(seq(0.99, 0.999, by=0.001), decreasing = TRUE),
  Tpos, Tneg)
compute_target_counts(
  sort(seq(0.001, 0.01, by=0.001), decreasing = TRUE),
  Tpos, Tneg)
```

Above we can see that this method is accurate even for these arbitrary target proportions.

## Visualization

```{r higgs}
br <- c(10^seq(-5, -1), 0.5)
breaks=unique(c(br, 1-br))
p_lo <- c(
  seq(0.01, 0.5, by=0.01),
  10^seq(-5, -1, by=0.1))
p_grid <- unique(sort(c(p_lo, 1-p_lo)))
(grid_dt <- compute_target_counts(p_grid, Tpos, Tneg))
grid_dt[, diff := p_neg-check_prop][order(abs(diff))]
library(ggplot2)
(grid_long <- melt(grid_dt, measure.vars=measure(prefix, class, pattern="^(|n_|N_)(pos|neg)$")))

n_max <- grid_long[, .SD[value==max(value)], by=.(prefix, class)]

(gg <- ggplot()+
   ggtitle("total counts")+
   theme_bw()+
   scale_size_manual(values=c(
     neg=3, pos=1))+
   scale_linetype_manual(values=c(
     "solid",
     "dotted",
     "dashed"))+
   geom_line(aes(
     p_neg, value, color=class, size=class, linetype=prefix),
     data=grid_long)+
   geom_hline(aes(
     yintercept=value),
     data=data.frame(value=c(Tpos, Tneg)))+
   scale_fill_manual(values=c(max="black"))+
   geom_point(aes(
     p_neg, value, color=class, fill=point),
     shape=21,
     data=n_max[, point := "max"]))
```

```{r higgs-log}
gg+
  scale_x_continuous(transform="logit", breaks=breaks)+
  scale_y_log10()
```

## other data

```{r}
Laribi_dt <- fread("~/projects/stratified-group-cv/data/Laribi2024.csv")
```

```{r Laribi}
ltab <- table(Laribi_dt$target)
(Laribi_grid_dt <- compute_target_counts(p_grid, ltab[1], ltab[2])[is.finite(check_prop)])
Laribi_grid_dt[, diff := p_neg-check_prop][order(abs(diff))]
(grid_long <- melt(Laribi_grid_dt, measure.vars=measure(prefix, class, pattern="^(|n_|N_)(pos|neg)$")))

n_max <- grid_long[, .SD[value==max(value)], by=.(prefix, class)]

(gg <- ggplot()+
   ggtitle("total counts")+
   theme_bw()+
   scale_size_manual(values=c(
     neg=3, pos=1))+
   scale_linetype_manual(values=c(
     "solid",
     "dotted",
     "dashed"))+
   geom_line(aes(
     p_neg, value, color=class, size=class, linetype=prefix),
     data=grid_long)+
   geom_hline(aes(
     yintercept=value),
     data=data.frame(value=as.numeric(ltab)))+
   scale_fill_manual(values=c(max="black"))+
   geom_point(aes(
     p_neg, value, color=class, fill=point),
     shape=21,
     data=n_max[, point := "max"]))
```

```{r Laribi-log}
gg+
  scale_x_continuous(transform="logit", breaks=breaks)+
  scale_y_log10()
```

## session info

```{r}
sessionInfo()
```
