---
layout: post
title: Finding issues in bibtex
description: Parsing bibtex using regular expressions
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-09-22-find-bib-issues"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=4)
options(width=120)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

Since writing
[Generate publications page](https://tdhock.github.io/blog/2024/auto-pubs-page/)
in 2024,
I have been updating my
[publications](https://tdhock.github.io/publications/) page by editing 
[TDH-refs.bib](/assets/TDH-refs.bib).
Now my university wants me to update [my entry in the Répertoire des Spécialistes](https://www.usherbrooke.ca/recherche/fr/specialistes/details/Toby.Dylan.Hocking), which pulls publications from [my ORCID works](https://orcid.org/0000-0002-3146-0865).
They have a bibtex import feature, but it gives error messages that are difficult to interpret.
For example, using [TDH-refs-old-sept-2026.bib](/assets/TDH-refs-old-sept-2026.bib) I get the error below:

```
This file cannot be read. Please check the BibTeX formatting and try again.
Value expected: single_value2023 } @article{Hillman2023jmlr, author = {Jonathan Hillman and Toby Dylan Hocking}, title = {Optimizing {ROC} Curves with a Sort-Based Surrogate Loss for Binary Classification and Changepoint Detection}, journal = {Journal of Machine Learning Research}, year = {2023}, volume = {24}, number = {70}, pages = {1--24}, links={[Publisher](https://jmlr.org/papers/v24/21-0751.html), [Preprint](https://arxiv.org/abs/2107.01285), [Software](https://github.com/tdhock/aum), [Reproducible](https://github.com/tdhock/max-generalized-auc), [Video](https://www.youtube.com/watch?v=GX6aqsi96IQ)}, url = {http://jmlr.org/papers/v24/21-0751.html} } @INPROCEEDINGS{Sweeney2023insect, author={Sweeney, Nathaniel and Xu, Caroline and Shaw, Joseph A. and Hocking, Toby D. and Whitaker, Bradley M.}, booktitle={2023 Intermountain Engineering, Technology and Computing (IETC)}, title={Insect Identification in Pulsed Lidar Images Using Changepoint Detection Algorithms}, year={2023}, pages={93-97}, links={[Publisher](https://ieeexplore.ieee.org/abstract/document/10152205)}, 
```

After deleting the links and trailing commas I get [TDH-refs-simplified-sept-2026.bib](/assets/TDH-refs-simplified-sept-2026.bib) with the error below:

```
This file cannot be read. Please check the BibTeX formatting and try again.
TypeError: Token mismatch: match
```

Neither error is helpful enough for me to figure out how to fix my bibtex input.
Below we use R to check the bib files.

## Parse bib into R

Parsing bibtex files is easy using regex. In fact, that is one of the
examples on `?nc::capture_all_str`:

```{r}
refs.bib <- "~/tdhock.github.io/assets/TDH-refs-old-sept-2026.bib"
refs.vec <- readLines(refs.bib)
at.lines <- grep("^@", refs.vec, value=TRUE)
str(at.lines)
```

The output above shows that there are `r length(at.lines)` lines that start
with `@` in the bib file. Below we use a regex to convert each item
into one row of a data table:

```{r}
options(datatable.prettyprint.char=15)
(refs.dt <- nc::capture_all_str(
  refs.vec,
  nc::before_match(
    "@",
    type="[^{]+", tolower,
    "[{]",
    ref="[^,]+",
    ",\n",
    fields="(?s).*?", function(x)gsub("\\{\\{", "\\{ \\{", x),
    "[}](?:$|\n *\n)")))
refs.dt[c(1,.N), fields]
```

The output above shows that the bib file was converted to a table with `r nrow(refs.dt)` rows.

## Check for fields too ling

```{r}
refs.dt[grep("@", fields)]
```

## Check for trailing commas

```{r}
refs.dt[grep(",\\s*$", fields)]
```

## Parsing fields 

First we look at the number of lines with an equals sign, each of which is probably a field.

```{r}
eq.lines <- grep("=", refs.vec, value=TRUE)
str(eq.lines)
```

Above we see `r length(eq.lines)` fields.

Below we parse the `fields` column:

```{r}
strip <- function(x)gsub("^\\s*|,\\s*$", "", gsub('[{}"]', "", x))
field.pattern <- list(
  "\\s*",
  variable="[^= ]+", tolower,
  "\\s*=",
  value=".*", strip)
(refs.fields.before.match <- refs.dt[, nc::capture_all_str(
  fields, nc::before_match(field.pattern)),
  by=.(type, ref)])
```

The table above has extra rows with empty match, which we exclude below:

```{r}
(refs.fields <- refs.fields.before.match[match!=""])
```

Above we see `r nrow(refs.fields)` fields, consistent with the simpler `grep` parsing above. 
If it is not consistent, we can use the code below to find out where:

```{r}
(eq.dt <- nc::capture_first_vec(eq.lines, field.pattern))
eq.dt[!refs.fields, on=.(variable,value)]
eq.counts <- eq.dt[, .(eq.count=.N), by=.(variable,value)]
refs.fields[, .(ref.count=.N), by=.(variable,value)][eq.counts,on=.(variable,value)][eq.count!=ref.count]
```

## Verify clean

Normally there should not be any quotes or curly braces in fields:

```{r}
cat(grep('[{}"]', refs.fields$value, value=TRUE), sep="\n\n")
```

## Conclusion

We have seen how to parse and check a bib file.

* `nc::before_match()` creates a pattern that parses an entire string into groups `before` and `match`.

## Session info

```{r}
sessionInfo()
```
