---
layout: post
title: Hierarchical imbalanced data generation
description: More controlled comparisons
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-09-14-hierachical-imbalanced-subset-generation"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
options(width=120)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to continue the exploration of [More precise imbalanced data generation](https://tdhock.github.io/blog/2026/exact-downsampling).
That post explained how to divide a binary classification data set into two subsets:

* one balanced subset with 50% positive, 50% negative labels.
* one imbalanced subset with an arbitrary proportion of negative labels, `p_neg`.
* such that the two subsets have the same number of samples.

That setup can be used for comparing imbalanced versus balanced training, in the context of our recently proposed [SOAK algorithm, Hocking et al, Comp Stat Data Anal 2026](https://doi.org/10.1002/sam.70055).
However that method does not allow for controlled comparison of training on different imbalance ratios, which would be interesting to explore how much class imbalance a learning algorithm can handle.
That would require a hierarchical definition of imbalanced subsets, which was discussed in [another previous blog](https://tdhock.github.io/blog/2025/imbalance-openml/) (but without control for sample size).
In this blog we explain a method which has both properties: 

* subsets have equal sample sizes,
* and different imbalance proportions are hierarchical.

Both properties are desirable, to ensure controlled comparison.
If we see a difference in test error rates, we want to be sure that the effect is due to the class imbalance, and not some other difference between train sets (like number of samples or which samples were used).

## Problem with previous method

We are creating machine learning algorithms for imbalanced classification problems.
We would like to know if these algorithms work well for training on imbalanced data, and predicting on balanced data, and vice versa.
We therefore would like to create different subsets for testing.
Assume we start with the `higgs` data, binary classification with this many data in each class:

```{r}
Tpos=5829123L
Tneg=5170877L
```

Below are the target proportions of the negative class:

```{r}
(Target_prop <- 10^seq(-1, -5))
```

Below we repeat the calculations in the previous post:

```{r}
library(data.table)
compute_target_counts <- function(p_neg, Tpos, Tneg){
  n_pos_max <- 2*Tpos*(1-p_neg)/(3-2*p_neg)
  n_pos_max_neg <- n_pos_max*p_neg/(1-p_neg)
  n_pos_max_N <- (n_pos_max_neg+n_pos_max)/2
  n_pos_max_neg_extra <- Tneg-n_pos_max_neg-n_pos_max_N
  n_neg_max <- 2*Tneg*p_neg/(1+2*p_neg)
  n_neg_max_pos <- n_neg_max*(1-p_neg)/p_neg
  n_neg_max_N <- (n_neg_max_pos+n_neg_max)/2
  ##n_neg_max_pos_extra <- Tpos-n_neg_max_pos-n_neg_max_N
  N_pos_neg <- as.integer(floor(
    ifelse(n_pos_max_neg_extra<0, n_neg_max_N, n_pos_max_N)))
  n_small <- as.integer(round(ifelse(
    p_neg<0.5,
    ifelse(n_pos_max_neg_extra<0, n_neg_max, n_pos_max_neg),
    ifelse(n_pos_max_neg_extra<0, n_neg_max_pos, n_pos_max))))
  n_large <- 2L*N_pos_neg-n_small
  data.table(
    p_neg,
    N_pos_neg,
    n_pos=ifelse(p_neg<0.5, n_large, n_small),
    n_neg=ifelse(p_neg<0.5, n_small, n_large)
  )[, let(
    pos = n_pos + N_pos_neg,
    neg = n_neg + N_pos_neg,
    check_prop = n_neg/(n_neg+n_pos)
  )][, let(
    prop_diff=check_prop-p_neg,
    extra_imb=n_pos+n_neg - 2L*N_pos_neg,
    unused_neg=Tneg-neg,
    unused_pos=Tpos-pos
  )][]
}
p_neg <- sort(c(Target_prop, 1-Target_prop), decreasing = TRUE)
(prev_dt <- compute_target_counts(p_neg, Tpos, Tneg))
```

Above we see the results from the previous post.

* it works for negative proportions greater than 0.5.
* There are always the same number of samples in the imbalanced and balanced subsets (`extra_imb=0` in every row).
* Every row has a valid `neg <= Tneg = 5170877` (no `neg` values that are too large to create), even for large values of the target proportion for the negative class, `p_neg >= 0.45`. This is verified above: all rows have non-negative numbers in the `unused` columns.

So each row defines a valid pair of subsets (one balanced, one imbalanced), but the rows are not consistent with each other.
Look at the `N_pos_neg` column. Values are not quite the same, so we can’t use this method to create hierarchical subsets.
We would like to have, for example:

* `p_neg=0.01` and `p_neg=0.1` having the same sized subsets, and
* `p_neg=0.01` having a subset of the negative examples of `p_neg=0.1`, and
* `p_neg=0.1` having a subset of the positive examples of `p_neg=0.01`.

## New hierarchical method

We assume there is a list of desired proportions of the negative class in the imbalanced subset:

```{r}
Target_prop
```

The most extreme is:

```{r}
(xp_neg <- min(Target_prop))
```

The main idea of the new method is to divide the whole data into three sets:

* two sets X and Y with equal numbers of positive and negative samples,
* one set E with some extra samples of either the positive or negative class.

For each target proportion of negative samples in the imbalanced subset,
these sets will be used to create subsets:

* Xi versus Y: an imbalanced version of X, created by removing the first few rows of the minority class, and by adding the first few rows of E.
* X versus Yi: an imbalanced version of Y, created by removing the first few rows of the minority class, and by adding the first few rows of E.
* X versus Y: both subsets balanced, for comparing prediction error rates with a baseline training method.

### Identifying active constraint

There are two inequality constraints to consider in the computation of `N`, the number of positive samples (and negative samples) in A and B (there are 2N samples in A, 2N samples in B, 4N samples total across A and B).

First constraint is the same as the previous method, `N + n_pos <= Tpos`.
If this is active, then the number of positive samples in the most imbalanced subset is the limiting factor.

```{r}
(n_pos_max <- 2*Tpos*(1-xp_neg)/(3-2*xp_neg))
(n_pos_max_neg <- n_pos_max*xp_neg/(1-xp_neg))
(n_pos_max_N <- as.integer(floor((n_pos_max_neg+n_pos_max)/2)))
(n_pos_max_bal_neg <- n_pos_max_N*2)
Tneg-n_pos_max_bal_neg
```

We see above that there are many negative samples left over, so this works.
Below we consider the other option, could N be limited by the number of negative samples in the balanced subset?

```{r}
(Tneg_N <- as.integer(floor(Tneg/2)))
(Tneg_n_pos <- Tneg*(1-xp_neg))
(Tneg_imb_pos <- Tneg_n_pos+Tneg_N)
Tpos-Tneg_imb_pos
```

We see above a negative value, indicating this is not feasible.
Maxing out the negative examples does not leave enough extra positive examples.
The other constraint must be active.

### Computing subset sizes

```{r}
extab <- function(Tprop, N, Tpos, Tneg){
  dt <- data.table(
    p_neg=sort(c(0.5, Tprop)),
    n_bal=2L*N
  )[, let(
    n_neg=as.integer(round(2*N*p_neg))
  )][, let(
    n_pos=n_bal-n_neg
  )][, let(
    n_imb=n_pos+n_neg,
    unused_pos=Tpos-n_pos-N,
    unused_neg=Tneg-n_neg-N
  )][]
  list(props=dt, params=dt[, data.table(
    extra_pos=max(n_pos-N),
    extra_neg=max(n_neg-N),
    N)])
}
extab(Target_prop, n_pos_max_N, Tpos, Tneg)
```

We see above that `n_bal == n_imb` for each row, which means the two subsets have the same number of samples, and we can create hierarchical subsets.

### Other constraint active

When would the other constraint become active?
If there are not enough negative samples, for example ten times fewer:

```{r}
Tpos=5829123L
Tneg=517087L

(n_pos_max <- 2*Tpos*(1-xp_neg)/(3-2*xp_neg))
(n_pos_max_neg <- n_pos_max*xp_neg/(1-xp_neg))
(n_pos_max_N <- as.integer(floor((n_pos_max_neg+n_pos_max)/2)))
(n_pos_max_bal_neg <- n_pos_max_N*2)
Tneg-n_pos_max_bal_neg
```

Above the negative number indicates that there are not enough negative samples for this constraint to be active.
Below we compute the feasibility for the other constraint.

```{r}
(Tneg_N <- as.integer(floor(Tneg/2)))
(Tneg_n_pos <- Tneg*(1-xp_neg))
(Tneg_imb_pos <- Tneg_n_pos+Tneg_N)
Tpos-Tneg_imb_pos
```

Above we see a positive number, which indicates that the number of positive samples is feasible.

```{r}
extab(Target_prop, Tneg_N, Tpos, Tneg)
```

Above we see `extra_neg` close to zero, indicating the number of negative samples was the limiting factor.

### Large target proportion of negative samples in imbalanced subset

How does the algorithm work for target proportion greater than one half?
In that case there are two inequalities to check:

* `N + n_neg <= Tneg`
* `2*N <= Tpos`

```{r}
Tpos=5829123L
Tneg=5170877L
(Target_prop <- 1-10^seq(-1, -5))
(xp_neg <- max(Target_prop))

(n_neg_max <- 2*Tneg*xp_neg/(1+2*xp_neg))
(n_neg_max_pos <- n_neg_max*(1-xp_neg)/xp_neg)
(n_neg_max_N <- (n_neg_max_pos+n_neg_max)/2)
(n_neg_max_bal_pos <- n_neg_max_N*2)
Tpos-n_neg_max_bal_pos
```

Above the positive number indicates that are enough positive samples for this constraint to be active.
Below we compute the feasibility for the other constraint.

```{r}
(Tpos_N <- as.integer(floor(Tpos/2)))
(Tpos_n_pos <- Tpos*(1-xp_neg))
(Tpos_n_neg <- Tpos_N*2-Tpos_n_pos)
(Tpos_imb_neg <- Tpos_n_neg+Tpos_N)
Tneg-Tpos_imb_neg
```

The negative number indicates this second constraint is not active.

### General code for both cases

The function below works for both cases.

```{r}
ntab <- function(Target_prop, Tpos, Tneg){
  if(all(Target_prop<0.5)){
    Tminor <- Tneg
    Tmajor <- Tpos
    p_minor <- Target_prop
  }else if(all(Target_prop>0.5)){
    Tminor <- Tpos
    Tmajor <- Tneg
    p_minor <- 1-Target_prop
  }else stop("Target_prop should be numeric, with each entry greater than 0.5, or each entry less than 0.5")
  xp_minor <- min(p_minor)
  (n_pos_max <- 2*Tmajor*(1-xp_minor)/(3-2*xp_minor))
  (n_pos_max_neg <- n_pos_max*xp_minor/(1-xp_minor))
  (n_pos_max_N <- as.integer(floor((n_pos_max_neg+n_pos_max)/2)))
  (n_pos_max_bal_neg <- n_pos_max_N*2)
  (Tminor_N <- as.integer(floor(Tminor/2)))
  N <- if(Tminor<n_pos_max_bal_neg)Tminor_N else n_pos_max_N
  extab(Target_prop, N, Tpos, Tneg)
}
ntab(1-10^seq(-1, -5), 5829123L, 5170877L)
ntab(1-10^seq(-1, -5), 582912L, 5170877L)
ntab(10^seq(-1, -5), 5829123L, 5170877L)
ntab(10^seq(-1, -5), 5829123L, 517087L)
ntab(10^seq(-1, -2), 3000L, 2000L)
```

We see in all the results above that `n_bal == n_imb`.
Either positive or negative numbers are the limiting factor.

## Creating CSV data based on counts

After having computed the target number of samples in each subset, we need to assign rows to A/B/E sets.
Here is a dummy data table,

```{r}
dummy.dt <- data.table(
  y=rep(0:1, c(2000,3000)),
  x1=NA, x2=NA) #example feature columns.
Tlist <- setNames(as.list(table(dummy.dt$y)), c("Tneg", "Tpos"))
Tlist$Target_prop <- 10^seq(-1, -2)
Tlist
(count.list <- do.call(ntab, Tlist))
```

Above we see the expected counts for a small problem with 2000 samples in A and B, with 980 extra positives.

```{r}
set.seed(1)
(ind.dt <- data.table(dummy.dt)[
, row := .I
][sample(.N)][
, set := NA_character_
][])
```

Above we see a random ordering of the data rows, set not yet assigned.

```{r}
label.list <- list(pos=1,neg=0)
N <- count.list$params$N
for(label.name in names(label.list)){
  label.value <- label.list[[label.name]]
  label.extra <- count.list$params[[paste0("extra_", label.name)]]
  set.values <- rep(c("X","Y","E"), c(N,N,label.extra))
  label.i <- which(ind.dt$y==label.value)[seq_along(set.values)]
  ind.dt[label.i, set := set.values]
}
ind.dt[, table(set, y, useNA="always")]
```

Above we see set has been assigned,

* 1000 positive samples in each of X and Y,
* 1000 negative samples in each of X and Y,
* 980 positive samples in E,
* 20 unused positive samples (missing set).

We assign fold below too.

```{r}
n.folds <- 5L
ind.dt[, fold := rep(1:n.folds, length.out=.N), by=.(set, y)][]
```

Now we create the output columns representing the different subsets.

```{r}
(out.unsort <- ind.dt[, data.table(
  fold,
  Xb_Yb=ifelse(set %in% c("X","Y"), set, NA))])
```

The `Xb_Yb` column represents the random assignment of two balanced subsets (a baseline).

```{r}
out.unsort[, table(fold, Xb_Yb)]
```

Above we can see the distribution of folds across subsets is uniform.
Below we create the other output columns, representing mapping of data rows to splits with an imbalanced subset.

```{r}
imb.counts <- count.list$props[p_neg != 0.5]
pos.part <- function(x)ifelse(x<0, 0, x)
for(cformat in c("Xineg%s_Yb", "Xb_Yineg%s")){
  for(imb.i in nrow(imb.counts):1){
    imb.row <- imb.counts[imb.i]
    j.name <- sprintf(cformat, imb.row$p_neg)
    set(
      out.unsort,
      j=j.name,
      value=ind.dt$set)
    imb.set <- ifelse(grepl("Xi", cformat), "X", "Y")
    for(label.name in names(label.list)){
      label.value <- label.list[[label.name]]
      label.n <- imb.row[[paste0("n_", label.name)]]
      find.rep.dt <- rowwiseDT(
        find.set=, sign=, rep.set=,
        imb.set, 1, NA, #rm
        "E", -1, imb.set)#add
      for(find.rep.i in 1:nrow(find.rep.dt)){
        find.rep.row <- find.rep.dt[find.rep.i]
        possible.indices <- ind.dt[, which(y==label.value & set==find.rep.row$find.set)]
        change.n <- pos.part((N-label.n)*find.rep.row$sign)
        change.indices <- possible.indices[seq_len(change.n)]
        set(
          out.unsort,
          i=change.indices,
          j=j.name,
          value=find.rep.row$rep.set)
      }
      is.E <- which(out.unsort[[j.name]]=="E")
      set(
        out.unsort,
        i=is.E,
        j=j.name,
        value=NA)
    }
  }
}
```

Finally we sort back to the original row order:

```{r}
orig.ord <- order(ind.dt$row)
(with.y.set <- data.table(ind.dt[, .(y, set)], out.unsort)[orig.ord])
```

The column names can be interpreted as follows:

* `b` suffix for balanced: `Xb` and `Yb`.
* `i` suffix for imbalanced, with a specified proportion of negative labels: `ineg0.1`, `ineg0.01`.

Below we count each unique row pattern.

```{r}
with.y.set[, .(rows=.N), keyby=names(with.y.set)]  
```

Above we can see the desired properties.

* In the first half of the table, with `y=0`, when a row has been removed in `ineg0.1`, then it is also removed in `ineg0.01` (hierarchical removal of negative samples as negative proportion is decreased).
* In the second half of the table, with `y=1`, when a row has been added in `ineg0.1`, then it is also added in `ineg0.01` (hierarchical addition of positive samples as negative proportion is decreased).
* The same E rows are used for addition in Xi and Yi.
* The distribution of folds is uniform: theres are equal numbers of rows with each pattern in each fold (stratification).

Below we create a table for output, with rows in the same order as the original table.

```{r}
(out.sort <- out.unsort[orig.ord])
```

The table above is a split info table that we can save alongside the original table.

```{r}
fwrite_head <- function(dt){
  fwrite(dt, data.file <- tempfile())
  cat(head(readLines(data.file)), sep="\n")
}
fwrite_head(dummy.dt)
fwrite_head(out.sort)
```

The output above shows the first few lines of the CSV file that describes the cross-validation experiment we have created.

## Verification

Now we verify that the counts are reasonable.

```{r}
check.dt.list <- list()
for(sub.col.i in 2:ncol(out.sort)){
  sub.col.name <- names(out.sort)[sub.col.i]
  two.cols <- out.sort[, c(1, sub.col.i), with=FALSE]
  setnames(two.cols, c("fold", "subset"))
  count.dt <- data.table(dummy.dt, two.cols)[, .(
    rows=.N
  ), keyby=.(y, fold, subset)]
  stats.dt <- dcast(
    count.dt,
    subset + y ~ .,
    list(sum, var, length),
    value.var="rows")
  check.dt.list[[sub.col.name]] <- data.table(sub.col.name, stats.dt)
}
(check.dt <- rbindlist(check.dt.list))
```

The table above has one row per combination of CSV column, subset, and label. We see that the results are reasonable.

* subset is either X or Y.
* number of rows per subset is always 2000.
* no variance between number of rows across folds, which means fold assignment respects stratification and subsets.

## Higgs data

Below we compute CSV subset files for two random seeds, and for downsampling both classes.

```{r}
(higgs.dt <- fread("higgs.csv", select="target"))

get_subsets <- function(y.vec, p_neg, n.folds=5L){
  Tlist <- setNames(as.list(table(y.vec)), c("Tneg", "Tpos"))
  Tlist$Target_prop <- p_neg
  Tlist
  (count.list <- do.call(ntab, Tlist))
  (ind.dt <- data.table(y=y.vec)[
  , row := .I
  ][sample(.N)][
  , set := NA_character_
  ][])
  label.list <- list(pos=1,neg=0)
  N <- count.list$params$N
  for(label.name in names(label.list)){
    label.value <- label.list[[label.name]]
    label.extra <- count.list$params[[paste0("extra_", label.name)]]
    set.values <- rep(c("X","Y","E"), c(N,N,label.extra))
    label.i <- which(ind.dt$y==label.value)[seq_along(set.values)]
    ind.dt[label.i, set := set.values]
  }
  ind.dt[, fold := rep(1:n.folds, length.out=.N), by=.(set, y)][]
  (out.unsort <- ind.dt[, data.table(
    fold,
    Xb_Yb=ifelse(set %in% c("X","Y"), set, NA))])
  imb.counts <- count.list$props[p_neg != 0.5][order(abs(p_neg-0.5))]
  pos.part <- function(x)ifelse(x<0, 0, x)
  for(cformat in c("Xineg%s_Yb", "Xb_Yineg%s")){
    for(imb.i in 1:nrow(imb.counts)){
      imb.row <- imb.counts[imb.i]
      j.name <- sprintf(cformat, imb.row$p_neg)
      set(
        out.unsort,
        j=j.name,
        value=ind.dt$set)
      imb.set <- ifelse(grepl("Xi", cformat), "X", "Y")
      for(label.name in names(label.list)){
        label.value <- label.list[[label.name]]
        label.n <- imb.row[[paste0("n_", label.name)]]
        find.rep.dt <- rowwiseDT(
          find.set=, sign=, rep.set=,
          imb.set, 1, NA, #rm
          "E", -1, imb.set)#add
        for(find.rep.i in 1:nrow(find.rep.dt)){
          find.rep.row <- find.rep.dt[find.rep.i]
          possible.indices <- ind.dt[, which(y==label.value & set==find.rep.row$find.set)]
          change.n <- pos.part((N-label.n)*find.rep.row$sign)
          change.indices <- possible.indices[seq_len(change.n)]
          set(
            out.unsort,
            i=change.indices,
            j=j.name,
            value=find.rep.row$rep.set)
        }
        is.E <- which(out.unsort[[j.name]]=="E")
        set(
          out.unsort,
          i=is.E,
          j=j.name,
          value=NA)
      }
    }
  }
  orig.ord <- order(ind.dt$row)
  out.unsort[orig.ord]
}

(higgs.p <- (10^seq(-3, -1))*0.5)
p.list <- list(neg=higgs.p, pos=1-higgs.p)
for(minor.class in names(p.list)){
  p.vec <- p.list[[minor.class]]
  for(seed in 1:2){
    set.seed(seed)
    (higgs.sub.dt <- get_subsets(higgs.dt$target, p.vec))
    print(out.csv <- sprintf(
      "higgs_subsets_seed=%d_pneg[%s,%s].csv",
      seed, min(p.vec), max(p.vec)))
    fwrite(higgs.sub.dt, out.csv)
  }
}
higgs.sub.dt
```

Above we see the last table created and saved to CSV.

```{r}
table(higgs.sub.dt$fold, paste(higgs.sub.dt[["Xb_Yineg0.9995"]], higgs.dt$target))
```

Above we see that the data counts across subsets and folds are as expected (X balanced, Y imbalanced mostly negative).

```{r}
cat(system("du -ms higgs*csv", intern=TRUE), sep="\n")
```

Above we see that the CSV files we created are 141 MB (for large `pneg`) or 147 MB (for small `pneg`).
This makes sense because there are fewer negative samples, as shown below.

```{r}
table(higgs.dt$target)
```

## Conclusions

We have shown how to split a binary classification data set into two subsets.

* Either the two subsets are both balanced (baseline),
* or one or the other subset is imbalanced to given proportions.
* each subset always has the same sample size, 
* and more imbalanced subsets are hierarchical: minor class samples are removed, major class samples are added (so there is a certain kind of continuity between imbalance proportions).

This code will be useful for creating imbalanced classification benchmarks, for comparing various machine learning algorithms.

## Session info

```{r}
sessionInfo()
```
