---
layout: post
title: Generating imbalanced subsets for SOAK benchmark
description: Creating meta-data for several data files
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-09-15-class-data-imb-subsets"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
options(width=120)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to continue the exploration of [Hierarchical imbalanced data generation](https://tdhock.github.io/blog/2026/hierachical-imbalanced-subset-generation).
That post explained how to divide a binary classification data set into hierarchical imbalanced subsets:

* Either the two subsets are both balanced (baseline),
* or one or the other subset is imbalanced to given proportions.
* each subset always has the same sample size, 
* and more imbalanced subsets are hierarchical: minor class samples are removed, major class samples are added (so there is a certain kind of continuity between imbalance proportions).

## Download benchmarks

Our goal is to apply this imbalance generation method to each data set used in the SOAK paper.
First, we download these data from zenodo.

```{r}
if(FALSE){
  unlink("SOAK_data_Classif.zip")
}
if(!file.exists("SOAK_data_Classif.zip")){
  options(timeout=9999)
  download.file(
    "https://zenodo.org/records/18273949/files/SOAK_data_Classif.zip?download=1",
    "SOAK_data_Classif.zip")
}
if(!file.exists("data_Classif")){
  unzip("SOAK_data_Classif.zip")
}
```

Then we read the target y values:

```{r}
(y.dt <- nc::capture_first_glob(
  Sys.glob("data_Classif/*"),
  "/",
  data.name=".*?",
  "[.]",
  READ=function(x)fread(x, select="y")))
```

Then we analyze them:

```{r}
(meta.dt <- y.dt[
, odd := as.integer(factor(y)) %% 2
, by=data.name
][, .(
  pos=sum(odd==1),
  neg=sum(odd==0),
  rows=.N
), by=data.name][
, min := pmin(pos, neg)
][order(min)])
```

The responses were binarized:  `odd` is 1 if factorized `y` was odd, otherwise 0.
We exclude small data sets:

```{r}
data.list <- split(y.dt[, .(y, odd)], y.dt$data.name)[
  meta.dt[min>10000, data.name]
]
str(data.list)
```

## Generate subsets

```{r}
extab <- function(Tprop, N, Tpos, Tneg){
  dt <- data.table(
    p_neg=sort(c(0.5, Tprop)),
    n_bal=2L*N
  )[, let(
    n_neg=as.integer(round(2*N*p_neg))
  )][, let(
    n_pos=n_bal-n_neg
  )][, let(
    n_imb=n_pos+n_neg,
    unused_pos=Tpos-n_pos-N,
    unused_neg=Tneg-n_neg-N
  )][]
  list(props=dt, params=dt[, data.table(
    extra_pos=max(n_pos-N),
    extra_neg=max(n_neg-N),
    N)])
}
ntab <- function(Target_prop, Tpos, Tneg){
  if(all(Target_prop<0.5)){
    Tminor <- Tneg
    Tmajor <- Tpos
    p_minor <- Target_prop
  }else if(all(Target_prop>0.5)){
    Tminor <- Tpos
    Tmajor <- Tneg
    p_minor <- 1-Target_prop
  }else stop("Target_prop should be numeric, with each entry greater than 0.5, or each entry less than 0.5")
  xp_minor <- min(p_minor)
  (n_pos_max <- 2*Tmajor*(1-xp_minor)/(3-2*xp_minor))
  (n_pos_max_neg <- n_pos_max*xp_minor/(1-xp_minor))
  (n_pos_max_N <- as.integer(floor((n_pos_max_neg+n_pos_max)/2)))
  (n_pos_max_bal_neg <- n_pos_max_N*2)
  (Tminor_N <- as.integer(floor(Tminor/2)))
  N <- if(Tminor<n_pos_max_bal_neg)Tminor_N else n_pos_max_N
  extab(Target_prop, N, Tpos, Tneg)
}
get_subsets <- function(y.vec, p_neg, n.folds=5L){
  Tlist <- setNames(as.list(table(y.vec)), c("Tneg", "Tpos"))
  Tlist$Target_prop <- p_neg
  Tlist
  (count.list <- do.call(ntab, Tlist))
  (ind.dt <- data.table(y=y.vec)[
  , row := .I
  ][sample(.N)][
  , set := NA_character_
  ][])
  label.list <- list(pos=1,neg=0)
  N <- count.list$params$N
  for(label.name in names(label.list)){
    label.value <- label.list[[label.name]]
    label.extra <- count.list$params[[paste0("extra_", label.name)]]
    set.values <- rep(c("X","Y","E"), c(N,N,label.extra))
    label.i <- which(ind.dt$y==label.value)[seq_along(set.values)]
    ind.dt[label.i, set := set.values]
  }
  ind.dt[, fold := rep(1:n.folds, length.out=.N), by=.(set, y)][]
  (out.unsort <- ind.dt[, data.table(
    fold,
    Xb_Yb=ifelse(set %in% c("X","Y"), set, NA))])
  imb.counts <- count.list$props[p_neg != 0.5][order(abs(p_neg-0.5))]
  pos.part <- function(x)ifelse(x<0, 0, x)
  for(cformat in c("Xineg%s_Yb", "Xb_Yineg%s")){
    for(imb.i in 1:nrow(imb.counts)){
      imb.row <- imb.counts[imb.i]
      j.name <- sprintf(cformat, imb.row$p_neg)
      set(
        out.unsort,
        j=j.name,
        value=ind.dt$set)
      imb.set <- ifelse(grepl("Xi", cformat), "X", "Y")
      for(label.name in names(label.list)){
        label.value <- label.list[[label.name]]
        label.n <- imb.row[[paste0("n_", label.name)]]
        find.rep.dt <- rowwiseDT(
          find.set=, sign=, rep.set=,
          imb.set, 1, NA, #rm
          "E", -1, imb.set)#add
        for(find.rep.i in 1:nrow(find.rep.dt)){
          find.rep.row <- find.rep.dt[find.rep.i]
          possible.indices <- ind.dt[, which(y==label.value & set==find.rep.row$find.set)]
          change.n <- pos.part((N-label.n)*find.rep.row$sign)
          change.indices <- possible.indices[seq_len(change.n)]
          set(
            out.unsort,
            i=change.indices,
            j=j.name,
            value=find.rep.row$rep.set)
        }
        is.E <- which(out.unsort[[j.name]]=="E")
        set(
          out.unsort,
          i=is.E,
          j=j.name,
          value=NA)
      }
    }
  }
  orig.ord <- order(ind.dt$row)
  out.unsort[orig.ord]
}

for(data.name in names(data.list)){
  data.dt <- data.list[[data.name]]
  floor.log10 <- floor(log10(min(table(data.dt$odd)))-1)
  (p.neg <- (10^seq(-floor.log10, -1))*0.5)
  p.list <- list(neg=p.neg, pos=1-p.neg)
  for(minor.class in names(p.list)){
    p.vec <- p.list[[minor.class]]
    for(seed in 1:2){
      set.seed(seed)
      (data.sub.dt <- get_subsets(data.dt$odd, p.vec))
      print(out.csv <- sprintf(
        "data_Classif_imb_subsets/%s/seed=%d_pneg[%s,%s].csv",
        data.name, seed, min(p.vec), max(p.vec)))
      out.dir <- dirname(out.csv)
      dir.create(out.dir, showWarnings = FALSE, recursive = TRUE)
      fwrite(data.sub.dt, out.csv)
    }
  }
}
table(data.sub.dt$fold, paste(data.sub.dt[["Xb_Yineg0.99995"]], data.dt$odd))
```

```{r}
cat(system("du -ms data_Classif_imb_subsets/*/*", intern=TRUE), sep="\n")
```

Above we can see the csv files created are from 1 to 42 MB.

## Session info

```{r}
sessionInfo()
```
