---
layout: post
title: More precise imbalanced data generation
description: No more rounding issues
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-06-11-exact-downsampling"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
options(width=120)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to continue the exploration of [Creating large imbalanced data benchmarks](https://tdhock.github.io/blog/2025/imbalance-openml/).

## Problem

We are creating machine learning algorithms for imbalanced classification problems.
We would like to know if these algorithms work well for training on imbalanced data, and predicting on balanced data, and vice versa.
We therefore would like to create different subsets for testing.
Assume we start with the `higgs` data, binary classification with this many data in each class:

```{r}
Tpos=5829123L
Tneg=5170877L
```

Below are the target proportions of the negative class:

```{r}
lseq <- 10^seq(-1, -5)
(Target_prop <- sort(c(4.5*lseq, lseq)))
```

Below we repeat the calculations in the previous post:

```{r}
library(data.table)
(orig_dt <- data.table(
  Target_prop,
  N_pos_neg = as.integer(Tpos/(3-2*Target_prop))
)[
, n_pos := Tpos-N_pos_neg
][
, n_neg := as.integer(Target_prop*n_pos/(1-Target_prop))
][
, n_imb := n_pos+n_neg
][, let(
  pos=n_pos+N_pos_neg,
  neg=n_neg+N_pos_neg,
  check_prop = n_neg/n_imb
)][])
```

* `Target_prop` is the desired proportion of negative samples in the imbalanced subset.
* `N_pos_neg` is the number of positive and negative samples in the balanced subset.
* `n_pos` and `n_neg` are the numbers of samples in the imbalanced subset, and `n_imb` is their sum.
* `pos` and `neg` are the total numbers of samples, across both subsets. We see that `pos=Tpos` in every row, and that `neg<Tneg` is true in every row except the last.
* `check_prop` is the actual proportion of negative samples in the imbalanced subset.

The numbers above seem reasonable, but

* There are almost the same number of samples in the imbalanced and balanced subsets, but not quite (difference shown in `extra_imb` column below).
* `Target_prop=4.5e-01` (almost balanced) has `neg>Tneg=5170877` which is not possible (not enough negative samples, shown as negative value in `unused_neg` column below).

```{r}
orig_dt[, .(
  Target_prop,
  prop_diff=check_prop-Target_prop,
  extra_imb=n_imb - N_pos_neg*2,
  unused_neg=Tneg-neg,
  unused_pos=Tpos-pos
)]
```

## Proposed fix

To fix the issues above, we start with the idea that for any target proportion, we want `n_neg` to be the largest integer such that the number of positive samples used is less than or equal the total number of samples: `N + n_pos <= Tpos` (and same for negative).
We use these facts

* `2*N_pos_neg = n_pos + n_neg`
* `p_neg = n_neg/(n_neg+n_pos)`
* `N_pos_neg + n_pos <= Tpos`
* `N_pos_neg + n_neg <= Tneg`

Starting with the inequality `N_pos_neg + n_neg <= Tneg`, we use the definition of `N_pos_neg` and `p_neg` to get `n_neg/(2*p_neg)*(2*p_neg+1) <= Tneg`, which we solve for `n_neg`. Same for the `Tpos` inequality.
If the variable were real numbers, it would be a linear program, and we could interpret the code below as solving by attempting to identify which constraint is active.

Actually the variables are integers, so we need to do some rounding.
The approach below first rounds `N_pos_neg` which is the number of negative and positive samples in the balanced subset, which therefore has `N_pos_neg*2` samples (always an even number).
To make sure the imbalanced subset has the same number of samples, we round the counts of the smaller class (`n_small` in the code below), and then define the counts of the larger class to be consistent.
That gives us the code below,

```{r}
compute_target_counts <- function(p_neg, Tpos, Tneg){
  n_pos_max <- 2*Tpos*(1-p_neg)/(3-2*p_neg)
  n_pos_max_neg <- n_pos_max*p_neg/(1-p_neg)
  n_pos_max_N <- (n_pos_max_neg+n_pos_max)/2
  n_pos_max_neg_extra <- Tneg-n_pos_max_neg-n_pos_max_N
  n_neg_max <- 2*Tneg*p_neg/(1+2*p_neg)
  n_neg_max_pos <- n_neg_max*(1-p_neg)/p_neg
  n_neg_max_N <- (n_neg_max_pos+n_neg_max)/2
  ##n_neg_max_pos_extra <- Tpos-n_neg_max_pos-n_neg_max_N
  N_pos_neg <- as.integer(floor(
    ifelse(n_pos_max_neg_extra<0, n_neg_max_N, n_pos_max_N)))
  n_small <- as.integer(round(ifelse(
    p_neg<0.5,
    ifelse(n_pos_max_neg_extra<0, n_neg_max, n_pos_max_neg),
    ifelse(n_pos_max_neg_extra<0, n_neg_max_pos, n_pos_max))))
  n_large <- 2L*N_pos_neg-n_small
  data.table(
    p_neg,
    N_pos_neg,
    n_pos=ifelse(p_neg<0.5, n_large, n_small),
    n_neg=ifelse(p_neg<0.5, n_small, n_large)
  )[, let(
    pos = n_pos + N_pos_neg,
    neg = n_neg + N_pos_neg,
    check_prop = n_neg/(n_neg+n_pos)
  )][, let(
    prop_diff=check_prop-p_neg,
    extra_imb=n_pos+n_neg - 2L*N_pos_neg,
    unused_neg=Tneg-neg,
    unused_pos=Tpos-pos
  )][]
}
p_neg <- sort(c(Target_prop, 1-Target_prop), decreasing = TRUE)
(fix_dt <- compute_target_counts(p_neg, Tpos, Tneg))
```

Above we see `check_prop` is still nearly equal to `p_neg` in every row, and the issues are fixed:

* it works for negative proportions greater than 0.5, unlike the previous code.
* There are always the same number of samples in the imbalanced and balanced subsets (`extra_imb=0` in every row).
* Every row has a valid `neg <= Tneg = 5170877` (no more `neg` values that are too large to create), even for large values of the target proportion for the negative class, `p_neg >= 0.45`. This is verified above: all rows have non-negative numbers in the `unused` columns.

Below we do the computation for a simple example with 140 positive samples, and 60 negative samples, which should be easy to split into

* a balanced subset with 50 positive and 50 negative samples,
* and an imbalanced subset with 90 positive and 10 negative samples (10% imbalance).

```{r}
compute_target_counts(seq(0.05, 0.15, by=0.01), 140L, 60L)
```

Above we can see that this method is accurate:

* for `p_neg=0.1`, we get `N_pos_neg=50`, `n_pos=90`, and `n_neg=10`.
* for other `p_neg` values (target proportion of negative samples in imbalanced subset), we get reasonable results (small `prop_diff` values).

## Visualization

The goal of this visualization is to understand how the number of samples in each class varies, as a function of `p_neg`, the target proportion of negative samples in the imbalanced subset.

```{r}
br <- c(10^seq(-5, -1), 0.5)
breaks <- unique(c(br, 1-br))
p_lo <- c(
  seq(0.01, 0.5, by=0.01),
  10^seq(-5, -1, by=0.1))
p_grid <- unique(sort(c(p_lo, 1-p_lo)))
(grid_dt <- compute_target_counts(p_grid, Tpos, Tneg))
```

The table above shows the results computed on a grid of input values.
The code above includes `p_lo` on the log and linear scale, because we want to visualize the results on both scales below.
First, the code below does the linear scale visualization.

```{r higgs}
library(ggplot2)
mygg <- function(dt, ref.vec){
  grid_long <- melt(
    dt,
    measure.vars=measure(prefix, class, pattern="^(|n_|N_)(pos|neg)$")
  )[, subset := ifelse(prefix=="", "imbalanced\n+balanced", "imbalanced")]
  n_max <- grid_long[, .SD[value==max(value)], by=.(subset, class)]
  ref.df <- data.frame(
    vjust=c(-0.5, 1.5),
    class=c("pos","neg"),
    value=ref.vec)
  ggplot()+
   theme_bw()+
   scale_linewidth_manual(values=c(
     neg=3, pos=1))+
   scale_linetype_manual(values=c(
     "solid",
     "dotted",
     "dashed"))+
   geom_line(aes(
     p_neg, value, color=class, linewidth=class, linetype=subset),
     data=grid_long)+
   geom_hline(aes(
     yintercept=value),
     data=ref.df)+
   geom_text(aes(
     0, value,
     vjust=vjust,
     label=sprintf(
       " Total %s samples = %s",
       class, format(value, big.mark=",", scientific=FALSE, trim=TRUE))),
     hjust=0,
     data=ref.df)+
    scale_fill_manual(values=c(max="black"))+
    scale_y_continuous("Number of samples (linear scale)")+
    scale_x_continuous(
      "Proportion of negative samples in imbalanced subset (linear scale)")+
   geom_point(aes(
     p_neg, value, color=class, fill=point),
     shape=21,
     data=n_max[, point := "max"])
}
(gg <- mygg(grid_dt, c(Tpos, Tneg)))
```

Above, the linear scale visualization emphasizes the details near the largest function values.
In these data, we see that 

* for small values of the proportion of negative samples in the imbalanced subset, all positive samples are used, and some negative samples are unused.
* for large values of the proportion of negative samples in the imbalanced subset, all negative samples are used, and some positive samples are unused.

Below, we do the log scale visualization,

```{r higgs-log}
mylog <- function(g){
  g+
    scale_x_continuous(
      "Proportion of negative samples in imbalanced subset (logit scale)",
      transform="logit", breaks=breaks)+
    scale_y_log10("Number of samples (log scale)")
}
mylog(gg)
```

Above, the log scale visualization emphasizes the details near the smallest values.
In particular, the linear trends on the log scale indicate that the number of positive/negative samples calculated is linearly related to `p_neg`, the desired proportion of negative samples in the imbalanced subset.

## Laribi2024 data

These data come from the [zenodo](https://zenodo.org/records/12954673) of Laribi, Raymond, Taseen, Poenaru, Vallières. Leveraging patients’ longitudinal data to improve the Hospital One-year Mortality Risk. Health Information Science and Systems (2024).

```{r Laribi}
Laribi_dt <- fread("~/projects/stratified-group-cv/data/Laribi2024.csv")
(ltab <- table(Laribi_dt$target))
min.n <- 10
(Laribi_grid_dt <- compute_target_counts(p_grid, ltab[1], ltab[2])[min.n <= n_neg & min.n <= n_pos])
(gg <- mygg(Laribi_grid_dt, as.numeric(ltab)))
```

In these data we see that for any values of the proportion of negative samples in the imbalanced subset, all negative samples are used, and some positive samples are unused.

```{r Laribi-log}
mylog(gg)
```

Note the x axis limits in the figure above denote the extreme negative class proportions which can be supported using these data, with a max of at least 10 samples from each class in each subset.

## Conclusions

We have studied a robust method for dividing a binary classification data set into one balanced subset, and another imbalanced subset, with a variable amount of class imbalance. We showed that the proposed method guarantees the same number of samples in each subset, so this method can be used with [SOAK](https://onlinelibrary.wiley.com/doi/10.1002/sam.70055), for comparing machine learning algorithms for imbalanced binary classification. Future work includes

* application to benchmark data (Higgs, MNIST, etc).
* generalization to multiple classes.

## session info

```{r}
sessionInfo()
```
