---
layout: post
title: More precise imbalanced data generation
description: No more rounding issues
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-06-11-exact-downsampling"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
options(width=120)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=10,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to continue the exploration of [Creating large imbalanced data benchmarks](https://tdhock.github.io/blog/2025/imbalance-openml/).

## Problem

We are creating machine learning algorithms for imbalanced classification problems.
We would like to know if these algorithms work well for training on imbalanced data, and predicting on balanced data, and vice versa.
We therefore would like to create different subsets for testing.
Assume we start with the `higgs` data, binary classification with this many data in each class:

```{r}
Tpos=5829123
Tneg=5170877
```

Below are the target proportions of the minority negative class:

```{r}
Target_prop <- 10^seq(-1, -5)
```

Below we repeat the calculations in the previous post:

```{r}
library(data.table)
(orig_dt <- data.table(
  Target_prop,
  N_pos_neg = as.integer(Tpos/(3-2*Target_prop))
)[
, n_pos := Tpos-N_pos_neg
][
, n_neg := as.integer(Target_prop*n_pos/(1-Target_prop))
][
, n_imb := n_pos+n_neg
][, let(
  pos=n_pos+N_pos_neg,
  neg=n_neg+N_pos_neg,
  check_prop = n_neg/n_imb
)][])
```
The numbers above seem reasonable, but

* `Target_prop=5e-01` (balanced) has `neg>Tneg` which is not possible (not enough negative samples).
* the `Target_prop` (desired imbalance) is often a bit larger than `check_prop` (actual imbalance).
* the `check_N` columns are off by one or two.

## Proposed fix

To fix the issues above, we start with the idea that for any target proportion, we want `n_neg` to be the largest integer such that the number of positive samples used is less than or equal the total number of samples: `N + n_pos <= Tpos` (and same for negative).
That gives us the code below, which also works for negative proportions greater than 0.5:

```{r}
compute_target_counts <- function(p_neg, Tpos, Tneg){
  p_small <- ifelse(p_neg<0.5, p_neg, 1-p_neg)
  n_small = as.integer(pmin(
    2*Tpos*p_small/(3*(1-p_neg)+p_neg),
    2*Tneg*p_small/(1+2*p_neg)
  ))
  data.table(p_neg)[, let(
    n_pos = ifelse(
      p_neg<0.5,
      n_small*(1-p_neg)/p_neg,
      n_small),
    n_neg = ifelse(
      p_neg<0.5,
      n_small,
      n_small*p_neg/(1-p_neg))
  )][
  , n_imb := n_pos+n_neg
  ][
  , N_pos := n_imb/2
  ][
  , N_neg := n_imb-N_pos
  ][, let(
    pos=n_pos+N_pos,
    neg=n_neg+N_neg
  )][
  , check_prop := n_neg/n_imb
  ][]
}
p_neg <- sort(c(Target_prop, 1-Target_prop), decreasing = TRUE)
#p_neg <- sort(seq(0.45, 0.55, by=0.01), decreasing = TRUE)
compute_target_counts(p_neg, Tpos, Tneg)
```

## Visualization

```{r higgs}
br <- c(10^seq(-5, -1), 0.5)
breaks=unique(c(br, 1-br))
p_lo <- c(
  seq(0.01, 0.5, by=0.01),
  10^seq(-5, -1, by=0.1))
p_grid <- unique(sort(c(p_lo, 1-p_lo)))
(grid_dt <- compute_target_counts(p_grid, Tpos, Tneg))
grid_dt[, diff := p_neg-check_prop][order(abs(diff))]
library(ggplot2)
(grid_long <- melt(grid_dt, measure.vars=measure(prefix, class, pattern="^(|n_|N_)(pos|neg)$")))

n_max <- grid_long[, .SD[value==max(value)], by=.(prefix, class)]

(gg <- ggplot()+
   ggtitle("total counts")+
   theme_bw()+
   scale_size_manual(values=c(
     neg=3, pos=1))+
   scale_linetype_manual(values=c(
     "solid",
     "dotted",
     "dashed"))+
   geom_line(aes(
     p_neg, value, color=class, size=class, linetype=prefix),
     data=grid_long)+
   geom_hline(aes(
     yintercept=value),
     data=data.frame(value=c(Tpos, Tneg)))+
   scale_fill_manual(values=c(max="black"))+
   geom_point(aes(
     p_neg, value, color=class, fill=point),
     shape=21,
     data=n_max[, point := "max"]))
```

```{r higgs-log}
gg+
  scale_x_continuous(transform="logit", breaks=breaks)+
  scale_y_log10()
```

## other data

```{r}
Laribi_dt <- fread("~/projects/stratified-group-cv/data/Laribi2024.csv")
```

```{r Laribi}
ltab <- table(Laribi_dt$target)
(Laribi_grid_dt <- compute_target_counts(p_grid, ltab[1], ltab[2])[is.finite(check_prop)])
Laribi_grid_dt[, diff := p_neg-check_prop][order(abs(diff))]
(grid_long <- melt(Laribi_grid_dt, measure.vars=measure(prefix, class, pattern="^(|n_|N_)(pos|neg)$")))

n_max <- grid_long[, .SD[value==max(value)], by=.(prefix, class)]

(gg <- ggplot()+
   ggtitle("total counts")+
   theme_bw()+
   scale_size_manual(values=c(
     neg=3, pos=1))+
   scale_linetype_manual(values=c(
     "solid",
     "dotted",
     "dashed"))+
   geom_line(aes(
     p_neg, value, color=class, size=class, linetype=prefix),
     data=grid_long)+
   scale_fill_manual(values=c(max="black"))+
   geom_point(aes(
     p_neg, value, color=class, fill=point),
     shape=21,
     data=n_max[, point := "max"]))
```

```{r Laribi-log}
gg+
  scale_x_continuous(transform="logit", breaks=breaks)+
  scale_y_log10()
```

## session info

```{r}
sessionInfo()
```
