---
layout: post
title: mlr3 learners list
description: And class projects
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-01-21-mlr3-learners"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=20,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(repo.dir, ".Rmd"))
}
```

The purpose of this vignette is to make lists of

* learners that currently exist in mlr3,
* CRAN packages that do not yet have a learner.

## Current Learners

mlr3 web site has [a list of learners](https://mlr-org.com/learners.html), created using [learners.qmd](https://github.com/mlr-org/mlr3website/blob/main/mlr-org/learners.qmd).

```{r}
pkgs <- c("mlr3", "mlr3learners", "mlr3extralearners", "mlr3proba", "mlr3cluster", "mlr3torch", "torchvision")
remotes::install_github("mlr-org/mlr3extralearners")
for(pkg in pkgs){
  if(!requireNamespace(pkg))install.packages(pkg)
  requireNamespace(pkg)
}
library(data.table)
content = as.data.table(mlr3::mlr_learners, objects = TRUE)
content[, base_package := purrr::map(object, function(x) strsplit(x$man, "::", TRUE)[[1]][1])]
content[, packages := purrr::pmap(list(packages, base_package), function(x, y) setdiff(x, c(y, "mlr3")))]
learners = rlang::set_names(content$object, content$key)
content[, `:=`(object = NULL, task_type = NULL)]

# fix mlr3probaproba
content[is.na(base_package), base_package := "mlr3proba"]

content
```

The table above seems consistent with the learners web page.

## wish list

I asked in [an issue](https://github.com/mlr-org/mlr3extralearners/issues/539), and there is a [list of issues which are learners to implement](https://github.com/mlr-org/mlr3extralearners/issues?q=is%3Aissue%20state%3Aopen%20label%3A%22Learner%20Status%3A%20Request%22) (CRAN packages with no mlr3 Learner written yet).
These would be good options for student projects in my class.

## session info

```{r}
sessionInfo()
```
