---
layout: post
title: Asypmptotic timing of text parsing
description: Regular expressions exercise
---

```{r Ropts, echo=FALSE}
repo.dir <- normalizePath("..")
post.id <- "2026-03-20-atime-regex"
fig.path <- file.path(repo.dir, "assets", "img", post.id)
knitr::opts_chunk$set(
  dpi=100,
  fig.path=paste0(fig.path, "/"),
  fig.width=20,
  fig.process=function(path)sub(repo.dir, "", path, fixed=TRUE),
  fig.height=6)
if(FALSE){
  knitr::knit(paste0(post.id, ".Rmd"))
}
```

The purpose of this page is to compare efficiency of different methods for parsing text.

## Problem

I recently wrote the test code below,

```{r}
sacct.txt <- system.file(
  package="slurm", "extdata", "sacct-e-rorqual-2026-03-20.txt", mustWork=TRUE)
computed <- slurm::sacct_fields(paste("cat", sacct.txt))
sacct.lines <- readLines(sacct.txt)
expected <- strsplit(gsub(" +", " ", paste(sacct.lines, collapse=" ")), " ")[[1]]
identical(computed, expected)
```

Are the two methods comparable speed?

## Test

The atime package allows us to see differences in computation time as a function of data size.

```{r atime}
ares <- atime::atime(
  setup={
    Nlines <- rep(sacct.lines, N)
  },
  seconds.limit = 0.1,
  do.call=do.call(c, strsplit(Nlines, " +")),
  unlist=unlist(strsplit(Nlines, " +")),
  strsplit=strsplit(paste(Nlines, collapse=" "), " +")[[1]],
  capture_all_str=nc::capture_all_str(Nlines, field="\\w+")$field)
plot(ares)
```

We can see above that there are some constant factor time and memory differences.

## session info

```{r}
sessionInfo()
```
